### 2014-02-25 paper.tex
Initial check in

\documentclass[conference]{sig-alt-gov2}
\pdfpagewidth=8.5in
\pdfpageheight=11in
%\usepackage{subfig}
\usepackage{subfigure}
\usepackage[pdftex]{}
\usepackage{todonotes}
\usepackage{listings}
\lstset {% general command to set parameter(s)
         language=C,
     basicstyle=\footnotesize,               % print whole listing small
%     keywordstyle=\color{black}\bfseries, % underlined bold black keywords
%     identifierstyle =\color{black},  % nothing happens
%     commentstyle=\color{black}\emph, % white comments
     %stringstyle=\ttfamily,          % typewriter type for strings
%     stringstyle=\color{black},       % typewriter type for strings
         tabsize=4,
         showtabs=false,
     showstringspaces=false}
%can't figure this one out for particles bandwidth
%\usepackage[caption,label]{subfig}

%\newcommand{\TITLE}{D\textsuperscript{\huge{2}}T: Doubly Distributed Transactions for High Performance and Distributed Computing}

\newcommand{\DDT}{D\textsuperscript{2}T~}
\newcommand{\DDTns}{D\textsuperscript{2}T}

\hyphenation{sub-trans-ac-tion}

% in order for balance columns to work, it has to be before begin document...
%\balancecolumns

\begin{document}

\conferenceinfo{MSST'14,} {November 17, 2013, Denver, Colorado, USA.}
\CopyrightYear{2014}
%\crdata{978-1-4503-1103-8/11/11}
\clubpenalty=10000
\widowpenalty = 10000

\title{Considerations for the Next Iteration of the DOE Fast Forward Storage and IO Project}

\numberofauthors{3}
\author{
\alignauthor Jay Lofstead\\
       \affaddr{Sandia National Laboratories}\\
       \email{gflofst@sandia.gov}
\alignauthor Ivo Jimenez\\
       \affaddr{University of California, Santa Cruz}\\
       \email{ivo@cs.ucsc.edu}
\alignauthor Carlos Maltzahn\\
       \affaddr{University of California, Santa Cruz}\\
       \email{carlosm@soe.ucsc.edu}
%\and  % use '\and' if you need 'another row' of author names
}
\maketitle

\begin{abstract}
The Fast Forward effort to define a next generation filesystem has made
admirable contributions in architecture and design. Formalizing the general
idea of data staging as burst buffers for the file system will help manage
the performance variability and offer the data processing opportunities
outside the main compute and file system. Adding a tranasctional mechanism to
manage faults and data visilibilty helping to enable effective analytics
without having to work around the IO stack semantics.

While these and other contributions are valuable, improvements can be
incorporated. For example, the Doubly Distributed Transactions (\DDTns)
protocol offers an alternative approach for incorporating transactional
semantics into the data path. The differing semantics between the two
approaches and the functionality provided offer alternatives. Further, the
lessons learned in implementing \DDT has generated insights into what the
FastForward file system could and perhaps should do, given different features.

This paper examines some of the choices made by the FastForward team and
compares them with other options and offers suggestions based on experiences.

\end{abstract}

\category{D.4}{Software}{Operating Systems}
\category{D.4.7}{Operating Systems}{Organization and Design}[hierarchical design]

\terms{Design, Performance}

\section{Introduction}

Current production HPC IO stack design is unlikely to offer sufficient features
and performance to adequately serve the needs of an extreme scale platform. To
address these limitations, the US Department of Energy commissioned an effort
to develop a design and prototype for an IO stack suitable for the extreme
scale environment. This is a joint effort led by the Intel Lustre team, Los
Alamos National Laboratory, EMC, and the HDF Group. This team has developed a
specification for a future IO stack to address the identified challenges.

The basic architecture incorporates five layers. The top layer is a high level
IO library, such as the demonstration HDF-5 library. Below that is an IO
forwarding layer that redirects IO calls from the compute nodes to the IO
dispatching layer. This IO forwarding layer is analogous to the function of the
IO nodes in a BlueGene machine. The IO dispatcher (IOD) has considerable
functionality.

The IOD serves as the sole interface between an application and the persistent
storage array. The core idea for IOD is to provide a way to manage the IO load
that is separate from the compute nodes and the storage array. Communication
intensive acitivies, such as two-phase, data sieving IO's data rearrangement
phase, can be moved to the IOD layer offloading the communication load from the
compute nodes. IOD has three main purposes. First, if the optional burst
buffer is availble, it works as a fast cache absorbing write write operations
for the slower trickle out to the central storage array. It can also be used to
retrieve objects from the central storage array for more efficient read
operations. Second, it offers the transaction mechanism for controlling data
set visibility prior to it being correct and complete. Third, data processing
operations can be placed in the IOD. This operations are inteded to offer data
rearrangement, filtering, and similar operations prior to it reaching the
central storage array.

This offloads the requirement for collective two-phase data sieving
on the compute nodes to reorganize data. This technique has been proven
effective at reducing the total time for writing data due to fewer participants
involved in the communication patterns~\cite{lofstead:2011:nessie-staging}.

The bottom two layers are the Data Access Object Storage (DAOS) and Virtual
Object Storage Device (VOSD). DAOS is the typical interface that represents
the individual objects stored including the semantics related to ``files''.
VSOD is the actual storage mechanism for the objects defined at the DAOS layer.

Incorporated into the design of these layers and the function of some of these
layers themsleves, such as the IOD layer, are features to manage both burst IO
performance and faults. In particular, the IOD layer's use of burst buffers
attempts to absorb a massive data dump from the compute area that is then trickled to the persistent storage area only if desired. Otherwise, it can be stored
locally on optional SSDs for use by analysis routines eliminating the need to
use slower rotational media as the intermediary between a simulation and the
analysis code.

A second feature is the incorporation of a transaction like mechanism into the
IOD layer and a related epochs feature into the DAOS layer. This feature offers
a mechanism to avoid making incomplete data visible to other users and offer a
level of protection if the data can be stored on the optional SSDs.

These features aim to address both the bursty nature of HPC IO and to make the
data storage stack resilient to faults. While these features can address these
concerns, a deeper evaluation of the design suggests exploring alternatives
may improve the performance and/or functionality desired.

The rest of the paper is organized as follows. Section~\ref{sec:burst}
discusses some of the features of incoporating burst buffers as designed and
suggests some considerations and alternatives for the next generation of this
project. Section~\ref{sec:transactions} discusses the transactions approach
offered in the IOD layer and offers a comparison to the \DDT system.
Section~\ref{sec:optimizations} discusses some of the optimizations offered
by the current design and how and when these optimizations will work best. It
also explores some of the dark corners that may yield performance issues with
the current design. Finally, Section~\ref{sec:summary} disucsses the system
overall with recommendations on what design elements should be reconsidered
based on broader issues with current HPC data centers.

\section{Burst Buffers}
\label{sec:burst}

The idea of burst buffers were initially explored in the context of data
staging~\cite{abbasi:2007:datatap,Abbasi:2009:datatap,nisar:2008:staging,zheng:2010:predata}
and later incorporated as part of the IO
stack~\cite{bent:2012:challenges,bent:2012:burst-buffer}. The proposed optional
use of SSDs is problematic. First, with these burst buffers being optional, the
semantics of how IOD works must change to use DAOS to store the data directly
rather than storing them locally until explicitly persisted by the user. How
this work work, the impact on metadata, and the ability to support
functionality such as data reordering, such as chainging the fast dimension of
an array, is undefined. How doing function shipping into an IOD without a burst
buffer would work is also not explored. Further thought about how to have an
IOD layer both with and without a burst buffer is required before they can be
considered optional. As the design stands today, they are a required part of
the IOD layer for proper functioning.

\section{Transactions and Epochs}
\label{sec:transactions}

The transaction mechanism manifests in two forms. At the IOD layer, they are
called transactions and are used to judge whether or not an output is complete
or not. At the DAOS layer, they are called epochs and represent persisted
transactions from the IOD layer. Each of these offers different functionality,
but are connected as is explained below.

\subsection{IOD Transactions}
To understand how transactions are used in the IOD layer, some terminology and
concepts must be explained first. At the coarsest grain level is a container.
Each container serves to host a transaction and contains a collection of
objects. Conceptually, containers corresponds to a file in a traditional file
system. The objects in each container represent different data within a file.
The three initially defined object types are key-value store, array shard,
and blob. The easiest way to understand these types is to evaluate these from
the perspective of an HDF-5 file. The key-value store represents a collection
of attributes. The array shard represents part of a potentially
multi-dimensional array. It is refered to as a shard because it is likely a
small piece of a globally defined array. The blob represents a byte stream.
The fundamental difference between an array shard and a blob is that the array
shard has metadata identifying its portion within the global, logical space
while the blob is simply a 1-dimensional array of bytes that is not shared
across IOD nodes. Should an operation be deployed into IOD to manipulate an
array within a container, it would operate on the array shards rather than on
blobs. Given this context, the transactions come in two forms.

First is a single leader transaction where the IOD manages based on calls from
a single client. The underlying assumption is that the client side will manage
the transactional operations itself and the single client is capable of
reporting to the IOD how to evolve the transaction state.

The second form is called multi-leader and has the IOD layer manage the
transactions. In this case, when the transaction is created, a count of clients
is provided to the IOD layer. As clients write to the container, the reference
count is reduced. Once the count reaches 0, the transaction is automatically
committed.

\subsection{DAOS Epochs}
The Epoch mechanism differs from transactions. Instead of focusing on when a
particular output is complete, an epoch represents incremental persisted copies
of a container. Each persisted copy increments the epoch. NEED TO CORRECT
THIS INFORMATION AS IT IS WRONG.

\subsection{Metadata Management}
While a central design goal of IOD and DAOS is to eliminate the metadata server
as a core component, it was not fully accomplished.  The purpose behind the goal
is to eliminate the serialized bottleneck caused by having a centralized
metadata service. Much of the design of IOD with objects and continers does
work to eliminate these bottlenecks. However, IOD does not go fully to a no
metadata service model. Instead, the IOD layer describes a key-value metadata
service it provides. This service maintains four different items:

\begin{itemize}
\item The list of IOD objects within the container
\item The mapping from each IOD object to the corresponding DAOS object.
\item The sharding and striping of the IOD object
\item The maximum valid offset of the object
\end{itemize}

While a pure no metadata model would be ideal for performance, this current
proposal has a few challenges. First, there are no list of containers. This
is closer to the no metadata ideal, but given the rest of the functionality,
it exports to the user level a requirement to track what containers are in the
file system. Second, given the flattening that occurs between IOD and DAOS,
the mapping from IOD objects to DAOS objects is not likely to be 1-to-1. This
is likely a list instead. Further, given the ability to re-stripe in the IOD
layer, the list is not a list of objects completely included in the IOD object,
but instead objects from which part of the data came from some listed DAOS
object.  Third, sharding IOD objects makes sense based on the compute process
connection a shard has with the complete object. Striping of an IOD object
makes sense in connection with the DAOS object(s) underneath, but striping
within the IOD layer is not discussed. Further, using striping from clients
into the IOD layer introduces a level of coordination that IOD seeks to avoid.
Fourth, given the previous items, there is ambiguity with the maximum valid
offset of the object. This could be for the shard or for the collection of
shard objects that represent a distributed variable.

The \DDT project created a simplistic model for a metadata service that has
different features that is described below and has been published about
previously~\cite{lofstead:2011:txn-metadata}. Some of the major differences
include the following. First, \DDTns's metadata service has a way to get a list
of the ojbjects in the metadata service. While it does not address having a
file equivalence like the container concept, it does introduce a way to figure
out what is available. Stored along with this object description is a list of
the global dimensions for array objects. One observtion made as part of
creating this service is that this is insufficient for many engineering codes
that use non-regular meshes. Second, for each object, a list of the equivalent
to shards are kept with a link back to the master object and the offset and
size for each dimension for that shard. This approach is not scalable because
of the explosion in the process count. Using a striping approach like IOD does
is superior. Third, there is no mapping to another layer with different object
layout and counts.

Based on the lessons from the \DDT metadata service construction, having a
completely separate metadata service is workable. Rather than making it a
bottleneck in the IO path, it is another service that users must interact with
if they need those services. Otherwise, it is 100% out of the way. Users can
manage everything by maintaining the metadata including the list of objects
themselves. However, there are drawbacks to this 100% client-side approach.

For the IOD proposal, the service needs some adjustments to be generally
useful.  First, there needs to be a way to query the list of available
containers. Second, the maximum transaction ID for a container is mentioned as
being maintained by this metadata service, but when and how this is done is
never mentioned. Second, it should be extracted from IOD entirely and kept as
a separate service along side. This will eliminate any contention caused by
interaction with the IOD layer. This also offers a way to link with the DAOS
layer in a machine-independent way. The need for this is described below.
Third, the definition of an object needs to be made more clear. In some cases,
it seems that an object is a globally distributed object is represented by a
collection of shards. In other cases, it is a single shard.

\subsection{Comparison to \DDTns}
The \DDT project~\cite{lofstead:2012:txn} sought to develop an efficient
approach for handling ACID-style transactions in an environment with parallel
clients and multiple servers (doubly distributed). Rather than being aimed
solely at data movement operations, \DDT seeks to address the general problem
of managing any operation with multiple clients and servers.  Consider the
management of the analysis/visualization area, potentially similar to the IOD
concept. The transaction protocol is used to help manage resizing of the
resource allocation to the various analysis and visualization components.  For
the purposes of this discussion, \DDT could also be used to manage changing how
IOD processes and/or nodes are used without exposing these changes to the
client processes prematurely.  This has been described and analyzed
previously~\cite{dayal:2013:io-containers}.

To address the gap between pure \DDT and IOD's transactions, example data
storage and metadata services were created. These examples were not built for
performance but to determine the minimum functionality requirement for a
functional service. While these services were built without knowledge of how
IOD and DAOS operate, we came to similar functionality. The major difference is
that there are no dependencies between transactions that prevent visibility
should an older version be incomplete. This additional, intentional requirement
by IOD offers different functionality than \DDTns's example services. In the
case of \DDTns, the functionality is more minimal, but also avoids some of the
concerns outlined below.

The second iteration of the protocol~\cite{lofstead:2013:pdsw-txn} fixed
scalability issues and demonstrated a scalable client-side coordiation model
with excellent performance. The performance measured for a complex transaction
with \DDT is illustrated in Figure~\ref{fig:performance}. This performance is
explored in detail in a previous paper~\cite{lofstead:2013:pdsw-txn}. The
breakdown of the number of participants in each role is shown in
Table~\ref{tab:scaling}. For comparison, consider the Number of
Sub-Coordinators equivalent to an IOD process. The Processes Per
Sub-Coordinator represents the number of clients that use a particular IOD. For
these tests, we maintained a balanced distribution and always used at least two
sub-coordinators to slow down the processing.

\begin{figure}[ht]
%\vspace{-0.15in}
\centering
\includegraphics[keepaspectratio=true, width=0.9\columnwidth]{images/performance}
\vspace{-0.15in}
\caption{Total Transaction Overhead}
\label{fig:performance}
\vspace{-0.15in}
\end{figure}

\begin{table}[ht]
    \vspace{-0.15in}
    \centering
    \caption[Scaling Configuration]{Performance Tests Scaling Configuration}
    \bigskip
    \vspace{-0.15in}

    \begin{tabular}{|r|r|r|}
\hline
Processes & \vtop{\hbox{\strut Number of}\hbox{\strut Sub-Coordinators}} & \vtop{\hbox{\strut Processes Per} \hbox{\strut Sub-Coordinator}}\\
\hline
%8 & 2 & 4 \\
%16 & 2 & 8 \\
%32 & 2 & 16 \\
%64 & 2 & 32 \\
%128 & 2 & 64 \\
256 & 2 & 128 \\
512 & 2 & 256 \\
1024 & 4 & 256 \\
2048 & 8 & 256 \\
4096 & 16 & 256 \\
8192 & 32 & 256 \\
16384 & 64 & 256 \\
32768 & 128 & 256 \\
65536 & 256 & 256 \\
\hline
    \end{tabular}
    \label{tab:scaling}
\end{table}

At a a high level, both \DDT and the IOD transactions have the same high-level
design. In both cases, a hierarchical model is employed. In the case of \DDTns,
it is a purely client-side tree. For IOD, it is a server-side tree. In both
cases, there is a master in charge of managing the transaction and a collection
of workers that aggregate into the master through second-level leaders. Beyond
that, there are some significant differences. Some of the different choices
made by IOD raise some possible concern.

First, \DDT has a timeout mechanism to detect failures and offer an ability to
recover and clean up the incomplete transactions. IOD's transactions are
asynchronous precluding a relatively simple, short timeout value to detect
failures. Unfortunately, at this point, there is no mechanism in IOD's
transaction handling to detect a fault and clean up any incomplete operations.

Second, should a transaction on a container not complete, all subsequent
transactions on that container, even if they are complete and correct, will not
be accessible.

Third, in the single leader model, if the process that manages the transaction
were to fail, there is no ability to abort that transaction. The data will not
be made available because the transaction is still listed as in progress. The
problem is that without a way to clean up this failed transaction, no newer
versions of the same container can be made readable. In reality, the single
leader model would ideally outsource the transaction processing to something
like \DDT on the client side.

Fourth, in the multi-leader model, using a count of client connections to
determine if transaction is complete is problematic.  First, should this be in
an evnironment where lost or repeated network messages are possible, then the
count could be wrong inappropriately. This is a general problem and unlikely
in the current environment, but is a concern on less integrated platforms.
The idea of a count precludes any tracking of expected messages from sources
introducing a potential hole. Second, the aggregation of completion messages
are based on the local aggregation point reporting to the transaction
leader the state.  There is no ability to detect or recover from a failure of
this IOD process that would report to the transaction leader. Third, because a
simple count of participating processes is sent with the beginning of the
transaction, each process is strictly limited to a single output operation.
This precludes a process writing multiple array shards, for example. To put
this in concrete terms, it means that a file could only contain a single,
globally distributed variable (at the root level of HDF5) OR an attribute per
process. If this count actually represents a begin/end transaction pair
instead, it suffers from a lack of accountability for missing object (shard)
writes. All it would indicate is that a process started and ended a transaction
connection.

\begin{figure}[ht]
\centering
\includegraphics[keepaspectratio=true, width=0.9\columnwidth]{images/optimized-protocol}
\vspace{-0.15in}
\caption{Optimized Protocol}
\label{fig:optimized-protocol}
\vspace{-0.25in}
\end{figure}

\DDT has addressed this count issue in a couple of ways. First, the
sub-coordinators each have a list of processes from which they expect messages.
Should a message be missed, it is noticed and corrective action can be taken.
Second, \DDT has the concept of sub-transactions. The messaging requirements
are illustrated in Figure~\ref{fig:optimized-protocol}.  Sub-transactions
represent finer grained operations than the entire output, \DDT can manage
multiple writes per client by using a sub-transaction to represent the output
for any item to the file (container). Because of how the sub-transactions are
managed, the singleton sub-transactions must be declared before the transaction
begins ensuring there is global knowledge that the sub-transaction is expected.
That way if the coordinator (transaction leader) fails, which ever process
takes over that role knows to expect a completion message for that
sub-transaction or the overall transaction cannot complete. While this
additional layer does introduce messaging, the overhead is quite small.

\subsection{Comparision to Other Protocols}
Alternatives, such as Paxos~\cite{Lamport:1998:paxos} algorithms like
ZooKeeper~\cite{Hunt:2010:zookeeper}, suffer from two limitations making them
unsuitable for this environment. First, the distributed servers in Paxos
systems are all distributed copies of each other that eventually become
consistent. Given the scale we wish to address, a single node's memory is
unlikely to be able to hold all of the data necessary for many operations at
scale. They also do not have a guaranteed for when consensus will be
achieved without using slower synchronous calls. For the tight timing we wish
to support, we need guarantees of when a consistent state has been achieved.
Second, these systems also all assume that updates are initiated from a single
client process rather than a parallel set of processes as is the standard in
HPC environments. The Intel/Lustre FastForward epoch
approach~\cite{barton:2013:fastforward,lombardi:2013:epochs} is discussed in
Section~\ref{lab:evaluation}.

Third, the server must offer a way to mark an operation as ``committed'' during
the ``commit'' phase of the two-phase commit.  Again, this may be implemented
in a transactional wrapper.  As an alternative, a scalable locking service like
Chubby~\cite{burrows:2006:chubby} might be employed.

Fourth, a second layer of coordination on the client side is introduced that
greatly increases the scalability by consolidating messages from clients into
unique sets prior to sending to the overall coordinator. A gossip
protocol~\cite{ganesh:2003:gossip-protocols} may appear sufficient for this
purpose, but the delay of eventual consistency is strictly avoided with this
protocol to ensure guarantees at particular states in the code. For example, if
a failure occurs, the global understanding of the role of all processes is
required in order for effective communication to occur for operations like
creating sub-transactions or voting. In this case, the protocol can offer
stronger statements about consistency than these protocols offer.  These
features offer a way to easily scale the transaction protocol given the
guarantees we wish to offer.

Another effort to offer consistency and data integrity for the ZFS file
system~\cite{zhang:2010:zfs} covers some of the same territory. Instead of a
focus on the processes all having a notion of completion as a transaction, this
work focuses on the integrity of the data movement operations. We view this
work as something that should be considered hand-in-hand with a transaction
approach to ensure the integrity of the movement of the data in addition to the
agreement of processes about the successful completion of a parallel operation.

\section{Optimizations}
\label{sec:optimizations}

The ability to place functionality at the IOD layer was explored earlier. In
particular, PreDatA~\cite{zheng:2010:predata} demonstrated the advantage to
the approach as well as identified that the placement of the operators should
consider the compute nodes, the staging or burst buffer nodes, or offline
depending on the characteristics of the operation. This consideration should
be incorporated into the design. In particular, one of the key observations
is that the reduced process count may require the entire time between output
operations to perform many of the data processing tasks requested. This was
part of the motivation prompting evaluating placement decisions at the compute
node, in a staging area, or offline. In the case of the IOD design, it forces
a staging area deployment and then shares that staging area across all users
simultaneously. This is unlikely to be useful because of the limited compute
and communication capacity to spare to perform these operations at a bottleneck
in the IO path. The use of a separate staging area to perform these computations
intentionally selects a location separate from the IO path to avoid these
bottlenecks and potential scalability issues.

\section{Broader Design}
\label{sec:summary}

Consider a shared file system across an HPC data center. The current design
maintains the metadata in the IOD layer localized to a single machine
effectively making the data inaccessible from another platform.

There is a change in definitions between the IOD layer and the DAOS layer.
For the IOD layer, a container is a collection of objects. For the DAOS layer,
a continer is a collection of shards. For the IOD an object may be a shard of
a global array. For DAOS, a shard can host a set of DAOS objects. Having the
same names with locally correct, globally conflicting definitions serves to
confuse how the system should work.

\section{Conclusions}
\label{lab:conclusion}

We have demonstrated that synchronous two-phase commit transactions can have
low overhead and be used with roughly synchronous operations with little
additional overhead. While other use cases, particularly those that require a
greater degree of asynchrony, require a different approach, \DDT offers a
working solution for a wide variety of scenarios today.  The bigger question of
what sort of synchronization mechanism is appropriate for various use cases is
currently under investigation.

The evaluation of the fault detection and recovery are beyond scope for this
paper and are in preparation for presentation elsewhere.

\section{Acknowledgements}
\includegraphics[scale=0.07]{logos/doe_logo}
\includegraphics[scale=0.30]{logos/snl_logo}
\includegraphics[scale=0.35]{logos/nnsa_logo}
Sandia National Laboratories is a multi-program laboratory managed and operated
by Sandia Corporation, a wholly owned subsidiary of Lockheed Martin
Corporation, for the U.S. Department of Energy's National Nuclear Security
Administration under contract DE-AC04-94AL85000.

\bibliographystyle{abbrv}
\bibliography{paper}

\vfill\eject

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\begin{figure}[ht]
%\vspace{-0.15in}
%\centering
%\includegraphics[keepaspectratio=true, width=0.9\columnwidth]{images/hierarchy}
%\vspace{-0.15in}
%\caption{Example Hierarchy}
%\label{fig:hierarchy}
%\vspace{-0.15in}
%\end{figure}

%\begin{figure}[ht]
%\centering
%\includegraphics[keepaspectratio=true, width=0.9\columnwidth]{images/stagingmodel}
%\vspace{-0.15in}
%\caption{Staging Model}
%\label{fig:staging-model}
%\vspace{-0.15in}
%\end{figure}

%\section{New Protocol Design}
%\label{lab:design}

%To summarize the semantics of \DDT for those unfamiliar with it, consider the
%following. Since we have parallel clients and servers (MxN), it is necessary to
%have a mechanism to manage each single operation by a collection of clients 
%with a collection of servers to ensure that it is completed correctly. We must
%also have a way to manage the set of these operations as a single, atomic
%action. The difficulty with this scenario is the need to maintain global
%knowledge across all of the clients and servers. This global knowledge allows
%decisions across all clients. In a traditional distributed transaction, the
%single client always knows the state reported by all of the ``server''
%resources used. When the client is split into multiple pieces, each piece
%ultimately needs to have the same knowledge, including any knowledge that might
%be local only. If this does not happen, no guarantees can be made. The message
%count explosion implied by this level of knowledge is daunting and what
%prompted this project. Adding complexity for data movement operations is that
%a typical data set will contain multiple globally distributed variables that
%each must be treated as an atomic unit.

%To accommodate these requirements, we have the concept of a transaction
%representing the total set of operations and the sub-transaction representing a
%single operation. Each sub-transaction can represent either a single process
%performing an action (singleton) or all of the parallel clients participating
%in a single action (global). In either case, each sub-transaction is managed as
%part of the overall transaction. Voting and commit operations are performed
%both at the sub-transaction level and at the transaction level. With the
%re-implementation of the protocol, these semantics have not changed.

%\section{Evaluation}
%\label{lab:evaluation}

%To evaluate the scalability and performance of \DDTns, tests are performed on
%Cielo at Sandia National Laboratories. Cielo is a Cray XE6 platform with 8,944
%nodes each containing dual, eight core AMD Opteron 6136 processors and 32 GB of
%RAM. It uses the Cray Gemini network in a 3D torus and uses the Cray Linux
%Environment. Tests are performed at powers of two scale from 256 up to 65536
%processes. We use two pairs of datastore and metadata services along side the
%process counts mentioned here. Each of these other services is given a single
%node on which to operate.

%The represented IAW consists of a writing application creating ten 3-D
%variables. Each of these variables is distributed across all of the processes
%with each process containing a 32x32x32 double precision floating point number
%chunk for each variable. The total space is scaled by doubling each dimension
%keeping the total dimensions as close to square as is possible. All ten of
%these variables and the corresponding chunks are written to the first pair of
%datastore and metadata servers. Then, an update application marks one variable
%as ``in process'' in the first metadata service, reads the chunks in parallel
%from the first datastore service, updates the data, creates a new variable in a
%second metadata service inserting the updated data into the second datastore
%service, and then deletes the original variable from the first datastore and
%metadata services. For the purposes of this evaluation, only the overhead
%incurred by the transaction protocol is reported.  No attempts were made to
%make the datastore and metadata service efficient and including those
%performance results would obfuscate the impact of having transactions surround
%data movement applications. The overheads are measured during the update
%application's activities and represent the green and blue arrows within
%Figure~\ref{fig:optimized-protocol}.

%With the three-tier structure employed by \DDTns, we choose to always employ at
%least two sub-coordinators, even at small scale, to incur higher overhead
%costs.  An example of this configuration is shown in
%Figure~\ref{fig:hierarchy}.  In this case, process 0 acts as the coordinator,
%sub-coordinator, and a subordinate. Process 3 acts as a sub-coordinator and a
%subordinate. The rest of the processes operate solely as subordinates. The
%configuration for each run consists of a minimum of 2 sub-coordinators and a
%maximum of 256 subordinates per sub-coordinator. That yields 256
%sub-coordinators each with 256 subordinates at 65536 processes. This is
%illustrated in Table~\ref{tab:scaling} showing the various scales evaluated and
%the configuration for each.  Other tests of smaller numbers of processes were
%run, but were omitted because they did not add any additional information to
%the results.

%The test values reported include the following actions:

%\begin{enumerate}
%\item txn\_create\_transaction - create a new transaction
%\item txn\_create\_sub\_transaction - create a new singleton sub-transaction
%\item txn\_create\_sub\_transaction\_all - called three times to create three global sub-transactions
%\item txn\_begin\_transaction - start the transaction processing
%\item txn\_commit\_sub\_transaction - called four times total to commit the four sub-transactions
%\item txn\_vote\_transaction - vote to commit or abort the overall transaction
%\item txn\_commit\_transaction - commit the entire transaction
%\item txn\_finalize\_txn - clean up any transaction specific memory or perform any final operations
%\end{enumerate}

%All operations except txn\_create\_sub\_transaction and \\
%txn\_commit\_sub\_transaction are performed as effectively collective
%operations. These two exceptions are only performed by the single process
%involved. In order for the existence of this singleton sub-transaction to be
%guaranteed to be known, we split the create and begin operations to provide a
%timeframe during which all singleton sub-transactions should be initialized.
%The begin operation gathers knowledge of all of these singleton
%sub-transactions and broadcasts it to all processes to make failure recovery
%possible. This step guarantees that every singleton sub-transaction will be
%known or the begin will fail due to a process failure. Since the global
%sub-transactions are global operations, they can be done either before or after
%the begin transaction call. In our case, we have one global sub-transaction
%created before the begin transaction and two afterwards.

%The selected sequence of operations gives a representative idea of what a
%typical transaction may include. Each test is performed seven times and the
%mean value is presented. Given the small magnitude of the values, there is a
%bit of variance that has a large percentage, but small absolute value. The
%results are presented in Figure~\ref{fig:performance}. The results for 4096
%processes were lost.

%In the interest of saving space, only the results for 64K processes is
%discussed. In all cases, the time spent in operations for the 64K case are the
%longest of all cases tested. At a detailed level, the time for the 64K
%processes case can be itemized as follows. First, the slowest operation, by
%far, is the txn\_create\_sub\_transaction\_all with a maximum time of 0.0310
%seconds for one of the calls. The mean time is 0.01 seconds. Second, the other
%operations are all at less than 0.005 seconds on average and a maximum of 0.012
%seconds.  Third, the time for the initialization and cleanup of the total
%protocol, the calls equivalent to MPI\_Init and MPI\_Finalize, take 0.38
%seconds total. Since the txn\_init call is the first call that uses MPI in the
%system and that it is a full order of magnitude slower than any other operation
%at worst, we believe there is MPI initial communication channel setup being
%established accounting for the extra time. The txn\_finalize takes a maximum of
%0.0002 seconds.  Since these calls are done once for the entire application
%lifetime, we did not include them in the results since they are not part of the
%typical transaction cost that would be incurred for each transactional
%operation.

%\section {Discussion}
%\label{lab:discussion}
%The leading alternative to \DDT is the Intel/Lustre FastForward Epoch
%approach~\cite{barton:2013:fastforward,lombardi:2013:epochs}.  While \DDT is
%intended for a general purposes, epochs are designed specifically to address
%the needs of a shared storage array with a large number of parallel clients.
%The general idea is to mark every operation with a transaction id, record a
%process count somewhere to allow reconciliation later, and have every process
%operate independently for the duration of the writing operation. A second, more
%synchronous mode is also available. In this synchronous mode, the user is
%required to manage the transaction ID and to update the transaction state in
%the storage servers. During a read process, it must be determined what is the
%most current, complete, ``Readable'' version of the file.  This entails
%contacting all of the storage devices to determine all of the versions of
%blocks stored on each device. With that information, it can be determined which
%of these transactions is ``Readable''. While this reconciliation operation may
%happen asynchronously, it must be done before a read operation potentially
%prompting it to be triggered by a client call or potentially stale data is
%presented as the most current.

\end{document}

### 2014-02-25 paper.tex
update for invalid cite text

@@
 different features that is described below and has been published about
-previously~\cite{lofstead:2011:txn-metadata}. Some of the major differences
+previously~\cite{lofstead:2012:txn-metadata}. Some of the major differences
 include the following. First, \DDTns's metadata service has a way to get a list

### 2014-02-25 2013-10-10-FastForward/include/iod_api.h
added Fast Forward docs for reference

@@ -1,0 +1,1188 @@
+/*
+ * The draft API for the IO Dispatcher (IOD).
+ * Last modified:  Feb 20, 2013  11:00AM
+ */
+
+#ifndef _IOD_API_H_
+#define _IOD_API_H_
+
+#include "iod_types.h"
+
+/**
+ * Instantiate IOD service.
+ *
+ * It should be called collectively by every application process at initializing
+ * phase.
+ *
+ * The \a total_cnranks and \a cnranks are for IOD optimization. For example
+ * if one TID's num_ranks == total_cnranks it means that all CN ranks
+ * participate it so IOD can use collective instead lots of P2P messages to
+ * synchronize transaction status; the cnranks can help IOD to know how many
+ * CN ranks are connected to it.
+ * User can pass in zero if it does not know it, but \a total_cnranks and
+ * \a cnranks should be provided with meaningful value or set as zero
+ * simultaneously. If they are provided, for application with dynamic processes,
+ * user should re-call this routine to notify IOD the new value after dynamic
+ * processes are created.
+ *
+ * \param comm [IN]		Global MPI communicator among all IODs
+ * \param hints[IN]		pointer to hints and can be NULL when no hint
+ * \param total_cnranks[IN]	total number of CN ranks in application
+ * \param cnranks[IN]		number of CN ranks that connect to this IOD
+ * \param event [IN]		pointer to completion event
+ *
+ * \return			zero on success, negative value if error
+ */
+iod_ret_t
+iod_initialize(iod_comm_t comm, iod_hint_list_t *hints,
+	       unsigned int total_cnranks, unsigned int cnranks,
+	       iod_event_t *event);
+
+/**
+ * Finalize IOD for resource freeing.
+ *
+ * It should be called collectively by every application process when user
+ * program exiting.
+ *
+ * \param hints[IN]		pointer to hints and can be NULL when no hint
+ * \param event [IN]		pointer to completion event
+ *
+ * \return			zero on success, negative value if error
+ */
+iod_ret_t
+iod_finalize(iod_hint_list_t *hints, iod_event_t *event);
+
+
+/* SECTION 1 ***** CONTAINER FUNCTIONS ********************/
+
+/**
+ * Open an IOD container.
+ *
+ * It should be collectively called by every application process.
+ * IOD will call daos_container_open to create/open DAOS container.
+ *
[…]

### 2014-02-25 2013-10-10-FastForward/include/iod_types.h

@@ -1,0 +1,440 @@
+/*
+ * Basic data type definitions for IOD(I/O Dispatcher).
+ */
+
+#ifndef _IOD_TYPES_H_
+#define _IOD_TYPES_H_
+
+#include <stdint.h>
+#include "mpi.h"
+
+/**
+ * IOD generic handle, which can refer to any local data structure (container,
+ * object, event queue ...)
+ */
+typedef struct {
+	uint64_t	cookie;
+} iod_handle_t;
+
+typedef MPI_Comm	iod_comm_t;	/** communicator among IODs */
+typedef uint64_t	iod_trans_id_t; /** IOD transaction ID */
+typedef int		iod_ret_t;	/** return value */
+typedef unsigned int	iod_mode_t;
+
+typedef uint64_t	iod_off_t;
+typedef uint64_t	iod_size_t;
+
+/** Container open modes */
+/** read-only */
+#define	IOD_CONT_RO			(1)
+/** write-only */
+#define	IOD_CONT_WO			(1 << 1)
+/** read-write */
+#define IOD_CONT_RW			(1 << 2)
+/** create container if it's not existed */
+#define IOD_CONT_CREATE			(1 << 7)
+
+/** IOD checksum value, 128bits */
+typedef struct {
+	uint64_t	cs_hi;
+	uint64_t	cs_lo;
+} iod_checksum_t;
+
+/** IOD object ID */
+typedef struct {
+	uint64_t	oid_hi;
+	uint64_t	oid_lo;
+} iod_obj_id_t;
+
+/** IOD object type */
+typedef enum {
+	IOD_OBJ_ANY,
+	IOD_OBJ_ARRAY,
+	IOD_OBJ_BLOB,
+	IOD_OBJ_KV,
+} iod_obj_type_t;
+
+/**
+ * IOD array object's dimensions sequence, it determines the layout mapping
+ * between logical dimensions and physical layout. The dedault dimensions
+ * sequence is same as logical dimension -- the first dimension is the slowest
+ * changing dimension and the higher dimensions are faster changing, the last
+ * dimension is the fastest changing on disk. User can change the mapping by
+ * calling iod_obj_set_layout.
+ *
+ * For contignous layout, it is an array with num_dims size. For example a 4D
+ * array will have uint32_t[4] ([0][1][2][3]) as default dimensions sequence.
+ * User can change it as [3][2][1][0] as reverse as logical dimension sequence
[…]

### 2014-02-26 paper.tex
Major editing pass. a quarter page too long now. One section still needs to be reconciled.

@@
 outside the main compute and file system. Adding a tranasctional mechanism to
-manage faults and data visilibilty helping to enable effective analytics
-without having to work around the IO stack semantics.
+manage faults and data visilibilty helps enable effective analytics without
+having to work around the IO stack semantics.
 
-While these and other contributions are valuable, improvements can be
-incorporated. For example, the Doubly Distributed Transactions (\DDTns)
-protocol offers an alternative approach for incorporating transactional
-semantics into the data path. The differing semantics between the two
-approaches and the functionality provided offer alternatives. Further, the
-lessons learned in implementing \DDT has generated insights into what the
-FastForward file system could and perhaps should do, given different features.
+While these and other their contributions are valuable, similar efforts made
+elsewhere may offer attractive alternatives or differing semantics that would
+yield a more feature rich environment with little to no additional overhead.
+For example, the Doubly Distributed Transactions (\DDTns) protocol offers an
+alternative approach for incorporating transactional semantics into the data
+path. The PreDatA project examined how to get the best throughput for data
+operators can offer some insights for refining the Burst Buffer concept.
 
 This paper examines some of the choices made by the FastForward team and
-compares them with other options and offers suggestions based on experiences.
+compares them with other options and offers observations and suggestions based
+on these other efforts.  This will include some non-core contributions, such
+as some of the demonstration components generated while implementing \DDT,
+to make suggestios that may help the next generation design for how the IO
+stack works as a whole.
 
@@
 Alamos National Laboratory, EMC, and the HDF Group. This team has developed a
-specification for a future IO stack to address the identified challenges.
+specification set~\cite{fastforward:2014:docs} for a future IO stack to address
+the identified challenges.
 
 The basic architecture incorporates five layers. The top layer is a high level
-IO library, such as the demonstration HDF-5 library. Below that is an IO
-forwarding layer that redirects IO calls from the compute nodes to the IO
-dispatching layer. This IO forwarding layer is analogous to the function of the
-IO nodes in a BlueGene machine. The IO dispatcher (IOD) has considerable
-functionality.
+IO library, such as the demonstration HDF-5 library. The intent is to only have
+access to the storage stack through such an API to manage the complexity of
+working with the lower layers. Below the user API is an IO forwarding layer
+that redirects IO calls from the compute nodes to the IO dispatching layer.
+This IO forwarding layer is analogous to the function of the IO nodes in a
+BlueGene machine. The next two layers have considerable functionality. The IO
+dispatcher (IOD) serves as the primary storage interface for the IO stack and
+is the only way to reach the persistent storage array in lower layers.
 
-The IOD serves as the sole interface between an application and the persistent
-storage array. The core idea for IOD is to provide a way to manage the IO load
-that is separate from the compute nodes and the storage array. Communication
-intensive acitivies, such as two-phase, data sieving IO's data rearrangement
-phase, can be moved to the IOD layer offloading the communication load from the
-compute nodes. IOD has three main purposes. First, if the optional burst
-buffer is availble, it works as a fast cache absorbing write write operations
-for the slower trickle out to the central storage array. It can also be used to
+The core idea for IOD is to provide a way to manage the IO load that is
+separate from the compute nodes and the storage array. Communication intensive
+acitivies, such as two-phase, data sieving IO's data rearrangement process, can
+be moved to the IOD layer offloading the communication load from the compute
+nodes. IOD has three main purposes. First, if the optional burst buffer is
+availble, it works as a fast cache absorbing write write operations for the
+slower trickle out to the central storage array. It can also be used to
 retrieve objects from the central storage array for more efficient read
-operations. Second, it offers the transaction mechanism for controlling data
-set visibility prior to it being correct and complete. Third, data processing
-operations can be placed in the IOD. This operations are inteded to offer data
-rearrangement, filtering, and similar operations prior to it reaching the
-central storage array.
+operations and offer data filtering to make client reads more efficient.
+Second, it offers the transaction mechanism for controlling data set visibility
+and to manage faults. Third, data processing operations can be placed in the
+IOD. This operations are inteded to offer data rearrangement, filtering, and
+similar operations prior to it reaching the central storage array.
 
@@
 involved in the communication patterns~\cite{lofstead:2011:nessie-staging}.
+Beyond these broad items, there are many important details some of which are
+examined in more detail below.
 
@@
 
+The focus of this paper is primarily the IOD layer given the critical role it
+has in the performance and functionality of the entire stack. Some of the key
+features explored in this paper all have a focus on the IOD layer motivating
+the focus of this examination.
+
 Incorporated into the design of these layers and the function of some of these
@@
 performance and faults. In particular, the IOD layer's use of burst buffers
-attempts to absorb a massive data dump from the compute area that is then trickled to the persistent storage area only if desired. Otherwise, it can be stored
-locally on optional SSDs for use by analysis routines eliminating the need to
-use slower rotational media as the intermediary between a simulation and the
-analysis code.
+attempts to absorb a massive data dump from the compute area that is then
+trickled to the persistent storage area only if desired. Otherwise, it can be
+stored locally on optional SSDs for use by analysis routines eliminating the
+need to use slower rotational media as the intermediary between a simulation
+and the analysis code.
 
@@
 a mechanism to avoid making incomplete data visible to other users and offer a
-level of protection if the data can be stored on the optional SSDs.
-
-These features aim to address both the bursty nature of HPC IO and to make the
-data storage stack resilient to faults. While these features can address these
-concerns, a deeper evaluation of the design suggests exploring alternatives
-may improve the performance and/or functionality desired.
+level of protection if the data can be stored on the optional SSDs. The
+implications of how these concepts are designed are important to consider for
+the final design. Because they control data visibility and manage faults for
+output operations, they are key features.
 
@@
 project. Section~\ref{sec:transactions} discusses the transactions approach
-offered in the IOD layer and offers a comparison to the \DDT system.
-Section~\ref{sec:optimizations} discusses some of the optimizations offered
-by the current design and how and when these optimizations will work best. It
-also explores some of the dark corners that may yield performance issues with
-the current design. Finally, Section~\ref{sec:summary} disucsses the system
-overall with recommendations on what design elements should be reconsidered
-based on broader issues with current HPC data centers.
+offered in the IOD layer and the corresponding epochs in the DAOS layer. It
+also offers a comparison to the \DDT system given the very similar high-level
+design and motivating use case.  Section~\ref{sec:optimizations} discusses some
+of the optimizations offered by the current design and how and when these
+optimizations will work best. It also explores some of the dark corners that
+may yield performance issues with the current design. Finally,
+Section~\ref{sec:summary} disucsses the system overall with recommendations on
+what design elements should be considered based on broader issues with current
+HPC data centers.
 
@@
 The idea of burst buffers were initially explored in the context of data
-staging~\cite{abbasi:2007:datatap,Abbasi:2009:datatap,nisar:2008:staging,zheng:2010:predata}
-and later incorporated as part of the IO
-stack~\cite{bent:2012:challenges,bent:2012:burst-buffer}. The proposed optional
-use of SSDs is problematic. First, with these burst buffers being optional, the
-semantics of how IOD works must change to use DAOS to store the data directly
-rather than storing them locally until explicitly persisted by the user. How
-this work work, the impact on metadata, and the ability to support
-functionality such as data reordering, such as chainging the fast dimension of
-an array, is undefined. How doing function shipping into an IOD without a burst
-buffer would work is also not explored. Further thought about how to have an
-IOD layer both with and without a burst buffer is required before they can be
+staging~\cite{abbasi:2007:datatap,Abbasi:2009:datatap,nisar:2008:staging,zheng:2010:predata}.
+These initial designs all use extra compute nodes to represent the data storage
+buffer given the lack of any hardware support for this functionality. The
+desired outcome of these initial studies is to motivate how such functionality
+might be incorporated and the potential benefits.  Later, these concepts were
+proposed to be incorporated as part of the IO
+stack~\cite{bent:2012:challenges,bent:2012:burst-buffer}.  The current
+FastForward IOD design recommends incorporating SSDs, but specifically lists
+these devices as optional. Unfortunately, both incorporating burst buffers and
+the use of SSDs in the IOD layer may be problematic.  First, with these burst
+buffers being optional, the semantics of how IOD works must change to use DAOS
+to store the data directly rather than storing them locally until explicitly
+persisted by the user. The design of DAOS does not incorporate a high
+transaction rate mechanism. Conversely, it assumes that it will only be
+involved when persisting a completed transaction and only for a fraction of the
+total transactions created. Further, with function shipping offering the
+ability to change how the data is stored and arranged prior to it being written
+to DAOS, this functionality may also be optional depending on the existence of
+the burst buffer. Consider the important functionality of data rearrangement
+and doing things like changing the fast array dimension.
+
+One of the bigger concerns is the observation that the original data staging
+proposals all used compute nodes while the newer proposals seek not only to
+make them a fixed portion of the IO stack, but also shared across all machine
+users. The PreDatA paper in particular examines the potential costs and
+advantages of where to place operators similar to the IOD proposed function
+shipping. There are two key takeaways from that work. First, placement matters.
+Depending on the communication intensity vs. computation intensity, where
+along the IO path to place the operation can matter significantly. Second, and
+more importantly, the amount of time spent processing for the operators was
+stretched to the point where it consumed nearly all of the time between IO
+operations. The given ratios of compute processes to staging process examined
+is representative for future extreme scale platforms. If anything, the ratios
+offer more staging processes than IOD processes would be available.
+
+By concentrating this functionality into the storage stack, two problems arise.
+First, the amount of network bandwidth, IO bandwidth, and compute power
+consumed for example operations from a single application is likely to
+completely monopolize the IOD processes. Second, if space and time partitioning
+is used instead, the functionality risks being too small to be useful. Third,
+the hardware performance or SSDs is questionable. Current NAND-based flash
+devices top out at around 400 MB/sec. The key spec that is missing from this
+number is that 400 MB/sec is a measure of the fixed number of available IOPS
+multipled by the block size. This represents the ideal streaming performance
+possible. The problem is that it costs an IOP to read 1 byte or 1 block. It
+costs 1 IOP to write a full block--usually. In some cases, it will cost 2 IOPs.
+This accounts for the required pre-erase write prior to writing the block. In
+the worst case, it can be 3 IOPS per write. This would be for a partial block
+write (read the old block, erase, write the new block). One-third of 400
+MB/sec, about 133 MB/sec, is well below the streaming performance of HDDs.
+Granted, there is still rotational and seek latency to deal with for HDDs, but
+the advantage for SSDs has evaporated and potentially turned into a penalty.
+
+Given these features, the optionality and even incorporation of burst buffers
+in the current design should be reconsidered. Much of the advanced, key
+functionality proposed as they are currently designed ultimatey relies on the
+existence of burst buffers to work. Further thought about how to have an IOD
+layer both with and without a burst buffer is required before they can be
 considered optional. As the design stands today, they are a required part of
-the IOD layer for proper functioning.
+the IOD layer for proper functioning. Unfortunately, it is not clear that they
+can address the performance concerns they are intended to cover.
 
@@
 called transactions and are used to judge whether or not an output is complete
-or not. At the DAOS layer, they are called epochs and represent persisted
-transactions from the IOD layer. Each of these offers different functionality,
-but are connected as is explained below.
+or not and control access through treating the transaction ID as a version
+identifier. At the DAOS layer, they are called epochs and represent persisted
+(durable) transactions from the IOD layer. Each of these offers different
+functionality, but are connected as is explained below. How these differ from
+the \DDT approach is also explored. While they are seemingly very different,
+they use a very similar approach to solve the same problem.
 
@@
 and blob. The easiest way to understand these types is to evaluate these from
-the perspective of an HDF-5 file. The key-value store represents a collection
-of attributes. The array shard represents part of a potentially
-multi-dimensional array. It is refered to as a shard because it is likely a
-small piece of a globally defined array. The blob represents a byte stream.
-The fundamental difference between an array shard and a blob is that the array
-shard has metadata identifying its portion within the global, logical space
-while the blob is simply a 1-dimensional array of bytes that is not shared
-across IOD nodes. Should an operation be deployed into IOD to manipulate an
-array within a container, it would operate on the array shards rather than on
-blobs. Given this context, the transactions come in two forms.
+the perspective of an HDF-5 file, the initial user interface layer. The
+key-value store represents a collection of attributes. The array shard
+represents part of a potentially multi-dimensional array. It is refered to as a
+shard because it is likely a small piece of a globally defined array. The blob
+represents a byte stream of arbitrary contents.  The fundamental difference
+between an array shard and a blob is that the array shard has metadata
+identifying its portion within the global, logical space while the blob is
+simply a 1-dimensional array of bytes that is not shared across IOD nodes.
+Should an operation be deployed into IOD to manipulate an array within a
+container, it would operate on the array shards rather than on blobs. Given
+this context, the transactions come in two forms.
 
@@
 particular output is complete, an epoch represents incremental persisted copies
-of a container. Each persisted copy increments the epoch. NEED TO CORRECT
-THIS INFORMATION AS IT IS WRONG.
+of a container. To simplify the mapping between an IOD transaction and the DAOS
+epochs, the IOD transaction ID is the same as the epoch ID. The key difference
+is that at the DAOS layer, some transaction (epoch) IDs will not be
+represented.
 
 \subsection{Metadata Management}
-While a central design goal of IOD and DAOS is to eliminate the metadata server
-as a core component, it was not fully accomplished.  The purpose behind the goal
-is to eliminate the serialized bottleneck caused by having a centralized
-metadata service. Much of the design of IOD with objects and continers does
-work to eliminate these bottlenecks. However, IOD does not go fully to a no
-metadata service model. Instead, the IOD layer describes a key-value metadata
-service it provides. This service maintains four different items:
+While a central design goal of the Fast Forward project to eliminate the
+requireent for a metadata server as a core component, it was not fully
+accomplished.  This idea is not new and has been explored as part of the Light
+Weight File Systems project~\cite{oldfield:lwfs}. In LWFS, the metadata service
+is explicitly limited to a user task with the storage layer limited to data
+storage/retrieval, authorization, and authentication. This approach proved
+workable. For the Fast Forward project, while a similar goal is desired, it is
+not completely accomplished.
+
+The purpose behind the goal is to eliminate the serialized bottleneck caused by
+having a centralized metadata service. Much of the design of IOD with objects
+and continers does work to eliminate these bottlenecks.  However, IOD does not
+go fully to a no metadata service model. Instead, the IOD layer describes a
+key-value metadata service it provides. This service maintains four different
+items:
 
@@
 While a pure no metadata model would be ideal for performance, this current
-proposal has a few challenges. First, there are no list of containers. This
-is closer to the no metadata ideal, but given the rest of the functionality,
-it exports to the user level a requirement to track what containers are in the
-file system. Second, given the flattening that occurs between IOD and DAOS,
-the mapping from IOD objects to DAOS objects is not likely to be 1-to-1. This
-is likely a list instead. Further, given the ability to re-stripe in the IOD
-layer, the list is not a list of objects completely included in the IOD object,
-but instead objects from which part of the data came from some listed DAOS
-object.  Third, sharding IOD objects makes sense based on the compute process
-connection a shard has with the complete object. Striping of an IOD object
-makes sense in connection with the DAOS object(s) underneath, but striping
-within the IOD layer is not discussed. Further, using striping from clients
-into the IOD layer introduces a level of coordination that IOD seeks to avoid.
-Fourth, given the previous items, there is ambiguity with the maximum valid
-offset of the object. This could be for the shard or for the collection of
-shard objects that represent a distributed variable.
+proposal has a few challenges. By providing this basic functionality, metadata
+management is split between the user layer and the IOD layer. Consdier the
+following items. First, there are no list of containers. This is closer to the
+no metadata ideal, but given the rest of the functionality, it exports to the
+user level a requirement to track what containers are in the file system.
+Second, given the flattening that occurs between IOD and DAOS, the mapping from
+IOD objects to DAOS objects is unlikely to be 1-to-1. This is likely a list
+instead. Further, given the ability to re-stripe in the IOD layer, the list is
+not a list of objects completely included in the IOD object, but instead
+objects from which part of the data came from some listed DAOS object.  Third,
+sharding IOD objects makes sense based on the compute process connection a
+shard has with the complete object. Striping of an IOD object makes sense in
+connection with the DAOS object(s) underneath, but striping within the IOD
+layer is not discussed. Further, using striping from clients into the IOD layer
+introduces a level of coordination that IOD seeks to avoid.  Fourth, given the
+previous items, there is ambiguity with the maximum valid offset of the object.
+This could be for the shard or for the collection of shard objects that
+represent a distributed variable. Ideally, either a robust metadata management
+system should be incorporated into the system, preferably separate from the IOD
+layer. Alternatively, all of this functionality should be left to the user
+layer. The only potential challenge is determining the container for a file
+open command and what the transaction ID should be. This limited operation
+should be a straightforward search using a hash or similar fast mechanism for
+examining a list of container names and/or object IDs.
 
 The \DDT project created a simplistic model for a metadata service that has
-different features that is described below and has been published about
+different features. It is described below and has been published
 previously~\cite{lofstead:2012:txn-metadata}. Some of the major differences
 include the following. First, \DDTns's metadata service has a way to get a list
-of the ojbjects in the metadata service. While it does not address having a
-file equivalence like the container concept, it does introduce a way to figure
-out what is available. Stored along with this object description is a list of
-the global dimensions for array objects. One observtion made as part of
-creating this service is that this is insufficient for many engineering codes
-that use non-regular meshes. Second, for each object, a list of the equivalent
-to shards are kept with a link back to the master object and the offset and
-size for each dimension for that shard. This approach is not scalable because
-of the explosion in the process count. Using a striping approach like IOD does
-is superior. Third, there is no mapping to another layer with different object
-layout and counts.
+of the objects in the metadata service. Since this was intended as a test
+system to examine \DDT performance, we did not put the effort into a making it
+a robust system. While it does not address having a file equivalence like the
+container concept, it does introduce a way to figure out what is available.
+Stored along with this object description is a list of the global dimensions
+for array objects. One observtion made as part of creating this service is that
+this is insufficient for many engineering codes that use non-regular meshes.
+Second, for each object, a list of the equivalent to shards are kept with a
+link back to the master object and the offset and size for each dimension for
+that shard. This approach is not scalable because of the explosion in the
+process count. Using a striping approach like IOD does is superior. Third,
+there is no mapping to another layer with different object layout and counts.
 
-Based on the lessons from the \DDT metadata service construction, having a
-completely separate metadata service is workable. Rather than making it a
-bottleneck in the IO path, it is another service that users must interact with
-if they need those services. Otherwise, it is 100% out of the way. Users can
-manage everything by maintaining the metadata including the list of objects
-themselves. However, there are drawbacks to this 100% client-side approach.
+Based on the lessons from the \DDT metadata service construction and the prior
+experiments with LWFS, having a completely separate metadata service is
+workable. Rather than making it a bottleneck in the IO path, it is another
+service that users must interact with if they need those services. Otherwise,
+it is 100\% out of the way. Users can manage everything by maintaining the
+metadata including the list of objects themselves. However, there are drawbacks
+to this 100\% client-side approach.
 
@@
 useful.  First, there needs to be a way to query the list of available
-containers. Second, the maximum transaction ID for a container is mentioned as
-being maintained by this metadata service, but when and how this is done is
-never mentioned. Second, it should be extracted from IOD entirely and kept as
-a separate service along side. This will eliminate any contention caused by
-interaction with the IOD layer. This also offers a way to link with the DAOS
-layer in a machine-independent way. The need for this is described below.
-Third, the definition of an object needs to be made more clear. In some cases,
-it seems that an object is a globally distributed object is represented by a
-collection of shards. In other cases, it is a single shard.
+containers. This will support an analysis application using data generated by
+the simulation code. This use case is a stated motivating example for the Fast
+Forward project. Second, the maximum transaction ID for a container is
+mentioned as being maintained by the IOD metadata service, but when and how
+this is done is never mentioned.  Third, the definition of an object needs to
+be made more clear. In some cases, it seems that an object is a globally
+distributed object is represented by a collection of shards. In other cases, it
+is a single shard.
 
@@
 
-To address the gap between pure \DDT and IOD's transactions, example data
-storage and metadata services were created. These examples were not built for
-performance but to determine the minimum functionality requirement for a
-functional service. While these services were built without knowledge of how
-IOD and DAOS operate, we came to similar functionality. The major difference is
-that there are no dependencies between transactions that prevent visibility
-should an older version be incomplete. This additional, intentional requirement
-by IOD offers different functionality than \DDTns's example services. In the
-case of \DDTns, the functionality is more minimal, but also avoids some of the
-concerns outlined below.
+The example metadata and data storage services created as part of the \DDT
+project lack one of the key transaction management features used by IOD.
+The major difference is that there are no dependencies between transactions
+that prevent visibility should an older version be incomplete. This additional,
+intentional requirement by IOD offers different functionality than \DDTns's
+example services. In the case of \DDTns, the functionality is more minimal, but
+also avoids some of the concerns outlined below.
 
@@
 design. In both cases, a hierarchical model is employed. In the case of \DDTns,
-it is a purely client-side tree. For IOD, it is a server-side tree. In both
-cases, there is a master in charge of managing the transaction and a collection
-of workers that aggregate into the master through second-level leaders. Beyond
-that, there are some significant differences. Some of the different choices
-made by IOD raise some possible concern.
+it is a purely client-side tree using semi-synchronous messaging. The messaging
+itself, in the current implementation, uses asynchronous MPI messages. The
+synchronous component comes from the timeout mechanism used to detect faults.
+It forces a level of coordiation and synchronization for the protocol. For IOD,
+it is a server-side tree and fully asynchronous. In both cases, there is a
+master in charge of managing the transaction and a collection of workers that
+aggregate into the master through second-level leaders. Beyond that, there are
+some significant differences. Some of the different choices made by IOD raise
+some possible concern.
 
@@
 writes. All it would indicate is that a process started and ended a transaction
-connection.
+connection. There is no validation that a process has done everything required
+of it.
 
@@
 unlikely to be able to hold all of the data necessary for many operations at
-scale. They also do not have a guaranteed for when consensus will be
-achieved without using slower synchronous calls. For the tight timing we wish
-to support, we need guarantees of when a consistent state has been achieved.
+scale. They also do not have a guarantee for when consensus will be achieved
+without using slower synchronous calls. For the tight timing we wish to
+support, we need guarantees of when a consistent state has been achieved.
 Second, these systems also all assume that updates are initiated from a single
 client process rather than a parallel set of processes as is the standard in
-HPC environments. The Intel/Lustre FastForward epoch
-approach~\cite{barton:2013:fastforward,lombardi:2013:epochs} is discussed in
-Section~\ref{lab:evaluation}.
+HPC environments.
 
-Third, the server must offer a way to mark an operation as ``committed'' during
-the ``commit'' phase of the two-phase commit.  Again, this may be implemented
-in a transactional wrapper.  As an alternative, a scalable locking service like
-Chubby~\cite{burrows:2006:chubby} might be employed.
+As an alternative, a scalable locking service like Chubby~\cite{burrows:2006:chubby} might be employed.
 
-Fourth, a second layer of coordination on the client side is introduced that
-greatly increases the scalability by consolidating messages from clients into
-unique sets prior to sending to the overall coordinator. A gossip
+\DDT uses a second layer of coordination on the client side that greatly
+increases the scalability by consolidating messages from clients into unique
+sets prior to sending to the overall coordinator. A gossip
 protocol~\cite{ganesh:2003:gossip-protocols} may appear sufficient for this
@@
 features offer a way to easily scale the transaction protocol given the
-guarantees we wish to offer.
+guarantees we wish to offer. The IOD approach does nothing to address these
+concerns.
 
@@
 \label{sec:optimizations}
+
+THIS MUST BE RECONCILED WITH THE EARLIER TEXT THAT SAYS LARGELY THE SAME THING.
 
@@
 
-We have demonstrated that synchronous two-phase commit transactions can have
-low overhead and be used with roughly synchronous operations with little
-additional overhead. While other use cases, particularly those that require a
-greater degree of asynchrony, require a different approach, \DDT offers a
-working solution for a wide variety of scenarios today.  The bigger question of
-what sort of synchronization mechanism is appropriate for various use cases is
-currently under investigation.
+The Fast Forward Storage and IO Stack project has designed a good first pass at
+addressing the requirements for an extreme scale data storage mechanism. The
+split between the IOD layer and the DAOS layer offers a fast place for
+intermediate data without requiring the overhead of writing to persistent
+storage. The envisioned transaction mechanism, while not perfect in the current
+form, is another good attempt to address both failures and prevent access to
+incomplete or incorrect data by downstream data consumers. Integrated with the
+IOD functionality, this concept represents the concensus approach for what will
+be required.
 
-The evaluation of the fault detection and recovery are beyond scope for this
-paper and are in preparation for presentation elsewhere.
+The partial metadata management incorporated into the IOD layer and the lack of
+consideration for how to handle and recover from failures are glaring
+oversights in the current documents. It is our undertanding that these will be
+addressed in the next phase and we hope to help inform that effort with our
+experiences.
+
+We hope that the efforts made in the \DDT project and other efforts to explore
+the requirements for this space, along with the analysis presented in this
+paper will prove useful for the next phase of the Fast Forward project.
 

### 2014-02-28 paper.tex
relatively minor edits. Some more notes of sections that need help. The others are in pretty good shape.

@@
 \begin{abstract}
-The Fast Forward effort to define a next generation filesystem has made
-admirable contributions in architecture and design. Formalizing the general
-idea of data staging as burst buffers for the file system will help manage
-the performance variability and offer the data processing opportunities
+With phase 1 of the Fast Forward Storage and IO Stack project complete, it is
+an excellent opportunity to evaluate many of the decisions made to feed into
+the phase 2 effort. The initial effort to define a next generation filesystem
+has made admirable contributions in architecture and design. Formalizing the
+general idea of data staging as burst buffers for the file system will help
+manage the performance variability and offer the data processing opportunities
 outside the main compute and file system. Adding a tranasctional mechanism to
@@
 compares them with other options and offers observations and suggestions based
-on these other efforts.  This will include some non-core contributions, such
-as some of the demonstration components generated while implementing \DDT,
-to make suggestios that may help the next generation design for how the IO
-stack works as a whole.
+on these other efforts.  This will include some non-core contributions of other
+projects, such as some of the demonstration metadata and data storage
+components generated while implementing \DDTns, to make suggestions that may
+help the next generation design for how the IO stack works as a whole.
 
@@
 specification set~\cite{fastforward:2014:docs} for a future IO stack to address
-the identified challenges.
+the identified challenges. The first phase recently completed with a second
+phase getting underway. The core focus of the first phase was basic
+functionality and design. The second phase will refine this design
+incorporating fault recovery and other features missing from the first phase.
 
@@
 Second, it offers the transaction mechanism for controlling data set visibility
-and to manage faults. Third, data processing operations can be placed in the
-IOD. This operations are inteded to offer data rearrangement, filtering, and
-similar operations prior to it reaching the central storage array.
+and to manage faults that should prevent a data set from being used. Third,
+data processing operations can be placed in the IOD. This operations are
+inteded to offer data rearrangement, filtering, and similar operations prior to
+it reaching the central storage array.
 
-This offloads the requirement for collective two-phase data sieving
-on the compute nodes to reorganize data. This technique has been proven
-effective at reducing the total time for writing data due to fewer participants
-involved in the communication patterns~\cite{lofstead:2011:nessie-staging}.
-Beyond these broad items, there are many important details some of which are
-examined in more detail below.
+Offloading the collective two-phase data sieving from the compute nodes to
+reorganize data has proven effective at reducing the total time for writing
+data due to fewer participants involved in the communication
+patterns~\cite{lofstead:2011:nessie-staging}.  Beyond these broad items, there
+are many important details some of which are examined in more detail below.
 
-The bottom two layers are the Data Access Object Storage (DAOS) and Virtual
+The bottom two layers are the Data Access Object Storage (DAOS) and Versioning
 Object Storage Device (VOSD). DAOS is the typical interface that represents
@@
 The focus of this paper is primarily the IOD layer given the critical role it
-has in the performance and functionality of the entire stack. Some of the key
-features explored in this paper all have a focus on the IOD layer motivating
-the focus of this examination.
+has in the performance and functionality of the entire stack. Most of the key
+features explored in this paper all have a strong presence in the IOD layer
+motivating the focus of this examination.
 
-Incorporated into the design of these layers and the function of some of these
-layers themsleves, such as the IOD layer, are features to manage both burst IO
-performance and faults. In particular, the IOD layer's use of burst buffers
-attempts to absorb a massive data dump from the compute area that is then
-trickled to the persistent storage area only if desired. Otherwise, it can be
-stored locally on optional SSDs for use by analysis routines eliminating the
-need to use slower rotational media as the intermediary between a simulation
-and the analysis code.
+%Incorporated into the design of these layers and the function of some of these
+%layers themsleves, such as the IOD layer, are features to manage both burst IO
+%performance and faults. In particular, the IOD layer's use of burst buffers
+%attempts to absorb a massive data dump from the compute area that is then
+%trickled to the persistent storage area only if desired. Otherwise, it can be
+%stored locally on optional SSDs for use by analysis routines eliminating the
+%need to use slower rotational media as the intermediary between a simulation
+%and the analysis code.
 
-A second feature is the incorporation of a transaction like mechanism into the
-IOD layer and a related epochs feature into the DAOS layer. This feature offers
-a mechanism to avoid making incomplete data visible to other users and offer a
-level of protection if the data can be stored on the optional SSDs. The
-implications of how these concepts are designed are important to consider for
-the final design. Because they control data visibility and manage faults for
-output operations, they are key features.
+%A second feature is the incorporation of a transaction like mechanism into the
+%IOD layer and a related epochs feature into the DAOS layer. This feature offers
+%a mechanism to avoid making incomplete data visible to other users and offer a
+%level of protection if the data can be stored on the optional SSDs. The
+%implications of how these concepts are designed are important to consider for
+%the final design. Because they control data visibility and manage faults for
+%output operations, they are key features.
 
@@
 optimizations will work best. It also explores some of the dark corners that
-may yield performance issues with the current design. Finally,
+may yield performance issues with the current design. 
 Section~\ref{sec:summary} disucsses the system overall with recommendations on
 what design elements should be considered based on broader issues with current
-HPC data centers.
+HPC data centers. The paper is conclued in Section~\ref{sec:conclusion} with
+a summary of the broad issues covered in the paper.
 
@@
 
-By concentrating this functionality into the storage stack, two problems arise.
-First, the amount of network bandwidth, IO bandwidth, and compute power
+By concentrating this functionality into the storage stack, three problems
+arise.  First, the amount of network bandwidth, IO bandwidth, and compute power
 consumed for example operations from a single application is likely to
@@
 is used instead, the functionality risks being too small to be useful. Third,
-the hardware performance or SSDs is questionable. Current NAND-based flash
-devices top out at around 400 MB/sec. The key spec that is missing from this
-number is that 400 MB/sec is a measure of the fixed number of available IOPS
-multipled by the block size. This represents the ideal streaming performance
-possible. The problem is that it costs an IOP to read 1 byte or 1 block. It
-costs 1 IOP to write a full block--usually. In some cases, it will cost 2 IOPs.
-This accounts for the required pre-erase write prior to writing the block. In
-the worst case, it can be 3 IOPS per write. This would be for a partial block
-write (read the old block, erase, write the new block). One-third of 400
-MB/sec, about 133 MB/sec, is well below the streaming performance of HDDs.
-Granted, there is still rotational and seek latency to deal with for HDDs, but
-the advantage for SSDs has evaporated and potentially turned into a penalty.
+the hardware performance advantage for SSDs is questionable. Current NAND-based
+flash devices top out at around 400 MB/sec. The key spec that is missing from
+this number is that 400 MB/sec is a measure of the fixed number of available
+IOPS multipled by the block size. This represents the ideal streaming
+performance possible. The problem is that it costs an IOP to read 1 byte or 1
+block (4KB or 8KB, depending on the device). It costs 1 IOP to write a full
+block--usually. In some cases, it will cost 2 IOPs.  This accounts for the
+required pre-erase write prior to writing to a reused block. In the worst case,
+it can be 3 IOPS per write. This would be for a partial block write (read the
+old block, erase, write the modified block). One-third of 400 MB/sec, about 133
+MB/sec, is well below the streaming performance of HDDs.  Granted, there is
+still rotational and seek latency to deal with for HDDs, but the advantage for
+SSDs has evaporated and potentially turned into a penalty at a considerable
+cost premium.  There are faster SSD solutions on the market that incorporate
+DRAM for caching and using the PCIe bus, for example, but their price precludes
+them from use in an extreme scale platform.
 
 Given these features, the optionality and even incorporation of burst buffers
-in the current design should be reconsidered. Much of the advanced, key
+in the current design should be carefully considered. Much of the advanced, key
 functionality proposed as they are currently designed ultimatey relies on the
@@
 functionality, but are connected as is explained below. How these differ from
-the \DDT approach is also explored. While they are seemingly very different,
-they use a very similar approach to solve the same problem.
+the \DDT approach is also explored. While IOD's and \DDTns's transactions are
+seemingly very different, they use a similar high-level design, but very
+different implementation, to solve the same problem.
 
@@
 of a container. To simplify the mapping between an IOD transaction and the DAOS
-epochs, the IOD transaction ID is the same as the epoch ID. The key difference
-is that at the DAOS layer, some transaction (epoch) IDs will not be
-represented.
+epochs, when an IOD transaction is persisted to DAOS, the IOD transaction ID is
+the used as the epoch ID. The key difference is that at the DAOS layer, some
+transaction (epoch) IDs will not be represented.
 
@@
 system should be incorporated into the system, preferably separate from the IOD
-layer. Alternatively, all of this functionality should be left to the user
-layer. The only potential challenge is determining the container for a file
-open command and what the transaction ID should be. This limited operation
-should be a straightforward search using a hash or similar fast mechanism for
-examining a list of container names and/or object IDs.
+layer or all of this functionality should be left to the user layer. The only
+potential challenge is determining the container for a file open command and
+what the transaction ID should be. This limited operation should be a
+straightforward search using a hash or similar fast mechanism for examining a
+list of container names and/or object IDs.
 
@@
 to this 100\% client-side approach.
+
+NEED TO ENUMERATE SOME OF THE CLIENT-SIDE ONLY PROBLEMS
 
@@
 transaction handling to detect a fault and clean up any incomplete operations.
+While this is intended for phase 2, some initial suggestion on how this might
+be handled would give some confidence that the asynchronous approach is viable.
 
@@
 leader model would ideally outsource the transaction processing to something
-like \DDT on the client side.
+like \DDT on the client side. This does not address the failure case though.
 
@@
 for any item to the file (container). Because of how the sub-transactions are
-managed, the singleton sub-transactions must be declared before the transaction
-begins ensuring there is global knowledge that the sub-transaction is expected.
-That way if the coordinator (transaction leader) fails, which ever process
-takes over that role knows to expect a completion message for that
-sub-transaction or the overall transaction cannot complete. While this
-additional layer does introduce messaging, the overhead is quite small.
+managed, the singleton sub-transactions, ones in which only a single process
+participates, must be declared before the transaction begins so that its
+existence can be broadcast as part of the begin transaction message. This
+ensures there is global knowledge that the sub-transaction is expected.  That
+way if the coordinator (transaction leader) fails, which ever process takes
+over that role knows to expect a completion message for that sub-transaction or
+the overall transaction cannot complete. Global sub-transactions can be defined
+at any time since they are a global, synchronized operation broadcasting their
+existence. While this additional layer does introduce messaging, the overhead
+is quite small.
+
+The advantages of eliminating these messages is not performance as demonstrated
+by the performance of \DDTns. Instead, it offers a much less synchronous model
+that matches with different programming models, such as Charm++ or other
+task-based approaches. Since it can work for the bulk-synchronous model also,
+it is a more broadly applicable approach. This assumes that the observed
+potential issues can be addressed successfully.
 
@@
 HPC environments.
-
-As an alternative, a scalable locking service like Chubby~\cite{burrows:2006:chubby} might be employed.
 
@@
 
+THiS SECTION NEEDS SOME HELP
+
 Consider a shared file system across an HPC data center. The current design
@@
 \section{Conclusions}
-\label{lab:conclusion}
+\label{sec:conclusion}
 
@@
 The partial metadata management incorporated into the IOD layer and the lack of
-consideration for how to handle and recover from failures are glaring
-oversights in the current documents. It is our undertanding that these will be
-addressed in the next phase and we hope to help inform that effort with our
-experiences.
+consideration for how to handle and recover from failures are oversights in the
+current documents. It is our undertanding that these will be addressed in the
+next phase and we hope to help inform that effort with our experiences.
 
@@
 
-\newpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\begin{figure}[ht]
-%\vspace{-0.15in}
-%\centering
-%\includegraphics[keepaspectratio=true, width=0.9\columnwidth]{images/hierarchy}
-%\vspace{-0.15in}
-%\caption{Example Hierarchy}
-%\label{fig:hierarchy}
-%\vspace{-0.15in}
-%\end{figure}
-
-%\begin{figure}[ht]
-%\centering
-%\includegraphics[keepaspectratio=true, width=0.9\columnwidth]{images/stagingmodel}
-%\vspace{-0.15in}
-%\caption{Staging Model}
-%\label{fig:staging-model}
-%\vspace{-0.15in}
-%\end{figure}
-
-%\section{New Protocol Design}
-%\label{lab:design}
-
-%To summarize the semantics of \DDT for those unfamiliar with it, consider the
-%following. Since we have parallel clients and servers (MxN), it is necessary to
-%have a mechanism to manage each single operation by a collection of clients 
-%with a collection of servers to ensure that it is completed correctly. We must
-%also have a way to manage the set of these operations as a single, atomic
-%action. The difficulty with this scenario is the need to maintain global
-%knowledge across all of the clients and servers. This global knowledge allows
-%decisions across all clients. In a traditional distributed transaction, the
-%single client always knows the state reported by all of the ``server''
-%resources used. When the client is split into multiple pieces, each piece
-%ultimately needs to have the same knowledge, including any knowledge that might
-%be local only. If this does not happen, no guarantees can be made. The message
-%count explosion implied by this level of knowledge is daunting and what
-%prompted this project. Adding complexity for data movement operations is that
-%a typical data set will contain multiple globally distributed variables that
-%each must be treated as an atomic unit.
-
-%To accommodate these requirements, we have the concept of a transaction
-%representing the total set of operations and the sub-transaction representing a
-%single operation. Each sub-transaction can represent either a single process
-%performing an action (singleton) or all of the parallel clients participating
-%in a single action (global). In either case, each sub-transaction is managed as
-%part of the overall transaction. Voting and commit operations are performed
-%both at the sub-transaction level and at the transaction level. With the
-%re-implementation of the protocol, these semantics have not changed.
-
-%\section{Evaluation}
-%\label{lab:evaluation}
-
-%To evaluate the scalability and performance of \DDTns, tests are performed on
-%Cielo at Sandia National Laboratories. Cielo is a Cray XE6 platform with 8,944
-%nodes each containing dual, eight core AMD Opteron 6136 processors and 32 GB of
-%RAM. It uses the Cray Gemini network in a 3D torus and uses the Cray Linux
-%Environment. Tests are performed at powers of two scale from 256 up to 65536
-%processes. We use two pairs of datastore and metadata services along side the
-%process counts mentioned here. Each of these other services is given a single
-%node on which to operate.
-
-%The represented IAW consists of a writing application creating ten 3-D
-%variables. Each of these variables is distributed across all of the processes
-%with each process containing a 32x32x32 double precision floating point number
-%chunk for each variable. The total space is scaled by doubling each dimension
-%keeping the total dimensions as close to square as is possible. All ten of
-%these variables and the corresponding chunks are written to the first pair of
-%datastore and metadata servers. Then, an update application marks one variable
-%as ``in process'' in the first metadata service, reads the chunks in parallel
-%from the first datastore service, updates the data, creates a new variable in a
-%second metadata service inserting the updated data into the second datastore
-%service, and then deletes the original variable from the first datastore and
-%metadata services. For the purposes of this evaluation, only the overhead
-%incurred by the transaction protocol is reported.  No attempts were made to
-%make the datastore and metadata service efficient and including those
-%performance results would obfuscate the impact of having transactions surround
-%data movement applications. The overheads are measured during the update
-%application's activities and represent the green and blue arrows within
-%Figure~\ref{fig:optimized-protocol}.
-
-%With the three-tier structure employed by \DDTns, we choose to always employ at
-%least two sub-coordinators, even at small scale, to incur higher overhead
-%costs.  An example of this configuration is shown in
-%Figure~\ref{fig:hierarchy}.  In this case, process 0 acts as the coordinator,
-%sub-coordinator, and a subordinate. Process 3 acts as a sub-coordinator and a
-%subordinate. The rest of the processes operate solely as subordinates. The
-%configuration for each run consists of a minimum of 2 sub-coordinators and a
-%maximum of 256 subordinates per sub-coordinator. That yields 256
-%sub-coordinators each with 256 subordinates at 65536 processes. This is
-%illustrated in Table~\ref{tab:scaling} showing the various scales evaluated and
-%the configuration for each.  Other tests of smaller numbers of processes were
-%run, but were omitted because they did not add any additional information to
-%the results.
-
-%The test values reported include the following actions:
-
-%\begin{enumerate}
-%\item txn\_create\_transaction - create a new transaction
-%\item txn\_create\_sub\_transaction - create a new singleton sub-transaction
-%\item txn\_create\_sub\_transaction\_all - called three times to create three global sub-transactions
-%\item txn\_begin\_transaction - start the transaction processing
-%\item txn\_commit\_sub\_transaction - called four times total to commit the four sub-transactions
-%\item txn\_vote\_transaction - vote to commit or abort the overall transaction
-%\item txn\_commit\_transaction - commit the entire transaction
-%\item txn\_finalize\_txn - clean up any transaction specific memory or perform any final operations
-%\end{enumerate}
-
-%All operations except txn\_create\_sub\_transaction and \\
-%txn\_commit\_sub\_transaction are performed as effectively collective
-%operations. These two exceptions are only performed by the single process
-%involved. In order for the existence of this singleton sub-transaction to be
-%guaranteed to be known, we split the create and begin operations to provide a
-%timeframe during which all singleton sub-transactions should be initialized.
-%The begin operation gathers knowledge of all of these singleton
-%sub-transactions and broadcasts it to all processes to make failure recovery
-%possible. This step guarantees that every singleton sub-transaction will be
-%known or the begin will fail due to a process failure. Since the global
-%sub-transactions are global operations, they can be done either before or after
-%the begin transaction call. In our case, we have one global sub-transaction
-%created before the begin transaction and two afterwards.
-
-%The selected sequence of operations gives a representative idea of what a
-%typical transaction may include. Each test is performed seven times and the
-%mean value is presented. Given the small magnitude of the values, there is a
-%bit of variance that has a large percentage, but small absolute value. The
-%results are presented in Figure~\ref{fig:performance}. The results for 4096
-%processes were lost.
-
-%In the interest of saving space, only the results for 64K processes is
-%discussed. In all cases, the time spent in operations for the 64K case are the
-%longest of all cases tested. At a detailed level, the time for the 64K
-%processes case can be itemized as follows. First, the slowest operation, by
-%far, is the txn\_create\_sub\_transaction\_all with a maximum time of 0.0310
-%seconds for one of the calls. The mean time is 0.01 seconds. Second, the other
-%operations are all at less than 0.005 seconds on average and a maximum of 0.012
-%seconds.  Third, the time for the initialization and cleanup of the total
-%protocol, the calls equivalent to MPI\_Init and MPI\_Finalize, take 0.38
-%seconds total. Since the txn\_init call is the first call that uses MPI in the
-%system and that it is a full order of magnitude slower than any other operation
-%at worst, we believe there is MPI initial communication channel setup being
-%established accounting for the extra time. The txn\_finalize takes a maximum of
-%0.0002 seconds.  Since these calls are done once for the entire application
-%lifetime, we did not include them in the results since they are not part of the
-%typical transaction cost that would be incurred for each transactional
-%operation.
-
-%\section {Discussion}
-%\label{lab:discussion}
-%The leading alternative to \DDT is the Intel/Lustre FastForward Epoch
-%approach~\cite{barton:2013:fastforward,lombardi:2013:epochs}.  While \DDT is
-%intended for a general purposes, epochs are designed specifically to address
-%the needs of a shared storage array with a large number of parallel clients.
-%The general idea is to mark every operation with a transaction id, record a
-%process count somewhere to allow reconciliation later, and have every process
-%operate independently for the duration of the writing operation. A second, more
-%synchronous mode is also available. In this synchronous mode, the user is
-%required to manage the transaction ID and to update the transaction state in
-%the storage servers. During a read process, it must be determined what is the
-%most current, complete, ``Readable'' version of the file.  This entails
-%contacting all of the storage devices to determine all of the versions of
-%blocks stored on each device. With that information, it can be determined which
-%of these transactions is ``Readable''. While this reconciliation operation may
-%happen asynchronously, it must be done before a read operation potentially
-%prompting it to be triggered by a client call or potentially stale data is
-%presented as the most current.
-
 \end{document}

### 2014-03-11 paper.tex
completed draft. Needs to be edited down for submission to 6 pages

@@
 
-\title{Considerations for the Next Iteration of the DOE Fast Forward Storage and IO Project}
+\title{Consistency and Fault Tolerance Considerations for the Next Iteration of the DOE Fast Forward Storage and IO Project}
 
@@
 an excellent opportunity to evaluate many of the decisions made to feed into
-the phase 2 effort. The initial effort to define a next generation filesystem
+the phase 2 effort. The initial effort to define a next generation file system
 has made admirable contributions in architecture and design. Formalizing the
@@
 manage the performance variability and offer the data processing opportunities
-outside the main compute and file system. Adding a tranasctional mechanism to
-manage faults and data visilibilty helps enable effective analytics without
+outside the main compute and file system. Adding a transactional mechanism to
+manage faults and data visibility helps enable effective analytics without
 having to work around the IO stack semantics.
@@
 separate from the compute nodes and the storage array. Communication intensive
-acitivies, such as two-phase, data sieving IO's data rearrangement process, can
-be moved to the IOD layer offloading the communication load from the compute
-nodes. IOD has three main purposes. First, if the optional burst buffer is
-availble, it works as a fast cache absorbing write write operations for the
+activities, such as two-phase, data sieving IO's data rearrangement process,
+can be moved to the IOD layer offloading the communication load from the
+compute nodes. IOD has three main purposes. First, if the optional burst buffer
+is available, it works as a fast cache absorbing write operations for the
 slower trickle out to the central storage array. It can also be used to
@@
 Second, it offers the transaction mechanism for controlling data set visibility
-and to manage faults that should prevent a data set from being used. Third,
-data processing operations can be placed in the IOD. This operations are
-inteded to offer data rearrangement, filtering, and similar operations prior to
-it reaching the central storage array.
+and to manage faults that would prevent a data set from being used. Third, data
+processing operations can be placed in the IOD. These operations are intended
+to offer data rearrangement, filtering, and similar operations prior to it
+reaching the central storage array.
 
@@
 %Incorporated into the design of these layers and the function of some of these
-%layers themsleves, such as the IOD layer, are features to manage both burst IO
+%layers themselves, such as the IOD layer, are features to manage both burst IO
 %performance and faults. In particular, the IOD layer's use of burst buffers
@@
 The rest of the paper is organized as follows. Section~\ref{sec:burst}
-discusses some of the features of incoporating burst buffers as designed and
+discusses some of the features of incorporating burst buffers as designed and
 suggests some considerations and alternatives for the next generation of this
@@
 also offers a comparison to the \DDT system given the very similar high-level
-design and motivating use case.  Section~\ref{sec:optimizations} discusses some
-of the optimizations offered by the current design and how and when these
-optimizations will work best. It also explores some of the dark corners that
-may yield performance issues with the current design. 
-Section~\ref{sec:summary} disucsses the system overall with recommendations on
-what design elements should be considered based on broader issues with current
-HPC data centers. The paper is conclued in Section~\ref{sec:conclusion} with
-a summary of the broad issues covered in the paper.
+design and motivating use case.  Section~\ref{sec:summary} discusses the system
+overall with recommendations on what design elements should be considered based
+on broader issues with current HPC data centers. The paper is concluded in
+Section~\ref{sec:conclusion} with a summary of the broad issues covered in the
+paper.
 
@@
 make them a fixed portion of the IO stack, but also shared across all machine
-users. The PreDatA paper in particular examines the potential costs and
-advantages of where to place operators similar to the IOD proposed function
-shipping. There are two key takeaways from that work. First, placement matters.
-Depending on the communication intensity vs. computation intensity, where
-along the IO path to place the operation can matter significantly. Second, and
-more importantly, the amount of time spent processing for the operators was
-stretched to the point where it consumed nearly all of the time between IO
-operations. The given ratios of compute processes to staging process examined
-is representative for future extreme scale platforms. If anything, the ratios
-offer more staging processes than IOD processes would be available.
+users. The PreDatA~\cite{zheng:2010:predata} paper in particular examines the
+potential costs and advantages of where to place operators similar to the IOD
+proposed function shipping. There are two key takeaways from that work. First,
+placement matters.  Depending on the communication intensity vs. computation
+intensity, where along the IO path to place the operation can matter
+significantly. Second, and more importantly, the amount of time spent
+processing for the operators was stretched to the point where it consumed
+nearly all of the time between IO operations. The given ratios of compute
+processes to staging process examined is representative for future extreme
+scale platforms. If anything, the ratios offer more staging processes than IOD
+processes would be available.
+
+In the case of the IOD design, it forces a staging area deployment and then
+shares the staging area across all users simultaneously. This is unlikely to be
+useful because of the limited compute and communication capacity to spare to
+perform these operations at a bottleneck in the IO path. The use of a separate
+data staging area intentionally separate from the IO path allows using
+operators on limited resources leaving the IO path clear for strictly data
+movement.
 
@@
 this number is that 400 MB/sec is a measure of the fixed number of available
-IOPS multipled by the block size. This represents the ideal streaming
+IOPS multiplied by the block size. This represents the ideal streaming
 performance possible. The problem is that it costs an IOP to read 1 byte or 1
@@
 in the current design should be carefully considered. Much of the advanced, key
-functionality proposed as they are currently designed ultimatey relies on the
+functionality proposed as they are currently designed ultimately relies on the
 existence of burst buffers to work. Further thought about how to have an IOD
@@
 key-value store represents a collection of attributes. The array shard
-represents part of a potentially multi-dimensional array. It is refered to as a
-shard because it is likely a small piece of a globally defined array. The blob
-represents a byte stream of arbitrary contents.  The fundamental difference
-between an array shard and a blob is that the array shard has metadata
-identifying its portion within the global, logical space while the blob is
-simply a 1-dimensional array of bytes that is not shared across IOD nodes.
-Should an operation be deployed into IOD to manipulate an array within a
-container, it would operate on the array shards rather than on blobs. Given
+represents part of a potentially multi-dimensional array. It is referred to as
+a shard because it is likely a small piece of a globally defined array. The
+blob represents a byte stream of arbitrary contents.  The fundamental
+difference between an array shard and a blob is that the array shard has
+metadata identifying its portion within the global, logical space while the
+blob is simply a 1-dimensional array of bytes that is not shared across IOD
+nodes.  Should an operation be deployed into IOD to manipulate an array within
+a container, it would operate on the array shards rather than on blobs. Given
 this context, the transactions come in two forms.
@@
 While a central design goal of the Fast Forward project to eliminate the
-requireent for a metadata server as a core component, it was not fully
+requirement for a metadata server as a core component, it was not fully
 accomplished.  This idea is not new and has been explored as part of the Light
@@
 having a centralized metadata service. Much of the design of IOD with objects
-and continers does work to eliminate these bottlenecks.  However, IOD does not
+and containers does work to eliminate these bottlenecks.  However, IOD does not
 go fully to a no metadata service model. Instead, the IOD layer describes a
@@
 proposal has a few challenges. By providing this basic functionality, metadata
-management is split between the user layer and the IOD layer. Consdier the
-following items. First, there are no list of containers. This is closer to the
+management is split between the user layer and the IOD layer. Consider the
+following items. First, there is no list of containers. This is closer to the
 no metadata ideal, but given the rest of the functionality, it exports to the
@@
 Stored along with this object description is a list of the global dimensions
-for array objects. One observtion made as part of creating this service is that
+for array objects. One observation made as part of creating this service is that
 this is insufficient for many engineering codes that use non-regular meshes.
@@
 
-NEED TO ENUMERATE SOME OF THE CLIENT-SIDE ONLY PROBLEMS
+With a client-side only approach, there is a serious risk of the metadata
+service and the object store getting out of sync.  While having a metadata-less
+object storage service is desirable, the different semantics from traditional
+file systems requires some considerations. In this case, should these services
+get out of sync, three particular risks introduced.  First, a client could
+create an entry in the metadata service that does not correspond to any objects
+in the object store. Second, a client could create objects that have no
+associated metadata entries. Third, updates to the metadata or object store
+service should be an atomic operation, but due to a lack of coordination, a
+window where the system is inconsistent appears.  Ultimately, the consistency
+semantics required must be determined. If a metadata service is required and it
+must be in sync with the object storage service, then additional work must be
+performed. In traditional file systems, the metadata and object storage updates
+are atomic. With decoupling metadata from object storage, should this atomicity
+still be desired, it requires both the ability for the services to participate
+in a task that is part of a larger atomic operation and a higher-level
+mechanism to manage the atomic operation. Overall, while additional work is
+required to maintain a client-side only metadata service, eliminating any
+potential bottlenecks related to updating metadata related to the object
+storage are eliminated. The burden of tracking striping and other metadata that
+has traditionally been part of the metadata associated with the file system
+will have to be maintained by the object storage service. The lack of a
+centralized, serialized bottleneck to store that information improves
+concurrency.
 
@@
 The second iteration of the protocol~\cite{lofstead:2013:pdsw-txn} fixed
-scalability issues and demonstrated a scalable client-side coordiation model
+scalability issues and demonstrated a scalable client-side coordination model
 with excellent performance. The performance measured for a complex transaction
@@
 
-At a a high level, both \DDT and the IOD transactions have the same high-level
+At a high level, both \DDT and the IOD transactions have the same high-level
 design. In both cases, a hierarchical model is employed. In the case of \DDTns,
@@
 synchronous component comes from the timeout mechanism used to detect faults.
-It forces a level of coordiation and synchronization for the protocol. For IOD,
-it is a server-side tree and fully asynchronous. In both cases, there is a
+It forces a level of coordination and synchronization for the protocol. For
+IOD, it is a server-side tree and fully asynchronous. In both cases, there is a
 master in charge of managing the transaction and a collection of workers that
@@
 determine if transaction is complete is problematic.  First, should this be in
-an evnironment where lost or repeated network messages are possible, then the
+an environment where lost or repeated network messages are possible, then the
 count could be wrong inappropriately. This is a general problem and unlikely
@@
 
-\subsection{Comparision to Other Protocols}
+\subsection{Comparison to Other Protocols}
 Alternatives, such as Paxos~\cite{Lamport:1998:paxos} algorithms like
@@
 
-\section{Optimizations}
-\label{sec:optimizations}
-
-THIS MUST BE RECONCILED WITH THE EARLIER TEXT THAT SAYS LARGELY THE SAME THING.
-
-The ability to place functionality at the IOD layer was explored earlier. In
-particular, PreDatA~\cite{zheng:2010:predata} demonstrated the advantage to
-the approach as well as identified that the placement of the operators should
-consider the compute nodes, the staging or burst buffer nodes, or offline
-depending on the characteristics of the operation. This consideration should
-be incorporated into the design. In particular, one of the key observations
-is that the reduced process count may require the entire time between output
-operations to perform many of the data processing tasks requested. This was
-part of the motivation prompting evaluating placement decisions at the compute
-node, in a staging area, or offline. In the case of the IOD design, it forces
-a staging area deployment and then shares that staging area across all users
-simultaneously. This is unlikely to be useful because of the limited compute
-and communication capacity to spare to perform these operations at a bottleneck
-in the IO path. The use of a separate staging area to perform these computations
-intentionally selects a location separate from the IO path to avoid these
-bottlenecks and potential scalability issues.
-
 \section{Broader Design}
@@
 
-THiS SECTION NEEDS SOME HELP
-
 Consider a shared file system across an HPC data center. The current design
 maintains the metadata in the IOD layer localized to a single machine
-effectively making the data inaccessible from another platform.
+effectively making the data inaccessible from another platform. This requires
+either the IOD layer be shared across machines, something clearly beyond the
+design scope, or that an additional metadata service that extends the minimal
+metadata service provided in the IOD layer. This higher level metadata service
+would be required to be reachable from any machine that shares the file system.
+As explained above, having this client-side only metadata service introduces
+additional functionality requirements.
 
-There is a change in definitions between the IOD layer and the DAOS layer.
-For the IOD layer, a container is a collection of objects. For the DAOS layer,
-a continer is a collection of shards. For the IOD an object may be a shard of
-a global array. For DAOS, a shard can host a set of DAOS objects. Having the
-same names with locally correct, globally conflicting definitions serves to
-confuse how the system should work.
+A point of confusion rather than a potential design challenge is the change in
+definitions between the IOD layer and the DAOS layer.  For the IOD layer, a
+container is a collection of objects. For the DAOS layer, a container is a
+collection of shards. For the IOD an object may be a shard of a global array.
+For DAOS, a shard can host a set of DAOS objects. Having the same names with
+locally correct, globally conflicting definitions serves to confuse how the
+system should work.
 
@@
 incomplete or incorrect data by downstream data consumers. Integrated with the
-IOD functionality, this concept represents the concensus approach for what will
+IOD functionality, this concept represents the consensus approach for what will
 be required.
@@
 consideration for how to handle and recover from failures are oversights in the
-current documents. It is our undertanding that these will be addressed in the
+current documents. It is our understanding that these will be addressed in the
 next phase and we hope to help inform that effort with our experiences.
 
-We hope that the efforts made in the \DDT project and other efforts to explore
-the requirements for this space, along with the analysis presented in this
-paper will prove useful for the next phase of the Fast Forward project.
+We hope that the efforts made in the \DDTns, Lightweight File Systems, and
+other efforts to explore the requirements for this space, along with the
+analysis presented in this paper will prove useful for the next phase of the
+Fast Forward project.
 

### 2014-03-15 paper.tex
revised according to conversation with Eric Barton

@@
 scale environment. This is a joint effort led by the Intel Lustre team, Los
-Alamos National Laboratory, EMC, and the HDF Group. This team has developed a
-specification set~\cite{fastforward:2014:docs} for a future IO stack to address
-the identified challenges. The first phase recently completed with a second
-phase getting underway. The core focus of the first phase was basic
-functionality and design. The second phase will refine this design
+Alamos National Laboratory, EMC, DDN, and the HDF Group. This team has
+developed a specification set~\cite{fastforward:2014:docs} for a future IO
+stack to address the identified challenges. The first phase recently completed
+with a second phase getting underway. The core focus of the first phase was
+basic functionality and design. The second phase will refine this design
 incorporating fault recovery and other features missing from the first phase.
+
+The overall design seeks to offer a byte-granular, multi-version concurrency
+control. Through the use of a copy-on-write style mechanism, multiple versions
+of a file can be stored in potentially greatly reduced space. By assuming the
+client interface will be through an IO library affords a more complicated
+interface that offers richer functionality without requiring user code changes.
+By managing data in the IOD layer rather than requring writing to centralized
+storage will better support the performnce requirements of integrated
+application workflows. At a more detailed view, the various layers of the IO
+stack each contribute different functionality and performance implications.
 
@@
 
-Offloading the collective two-phase data sieving from the compute nodes to
-reorganize data has proven effective at reducing the total time for writing
-data due to fewer participants involved in the communication
-patterns~\cite{lofstead:2011:nessie-staging}.  Beyond these broad items, there
-are many important details some of which are examined in more detail below.
+While these ideas are not necessarily new, they are new twists on best of class
+efforts for these techologies. For example, offloading the collective two-phase
+data sieving from the compute nodes to reorganize data has proven effective at
+reducing the total time for writing data due to fewer participants involved in
+the communication patterns~\cite{lofstead:2011:nessie-staging}.  Beyond these
+broad items, there are many important details some of which are examined in
+more detail below.
 
@@
 
-%Incorporated into the design of these layers and the function of some of these
-%layers themselves, such as the IOD layer, are features to manage both burst IO
-%performance and faults. In particular, the IOD layer's use of burst buffers
-%attempts to absorb a massive data dump from the compute area that is then
-%trickled to the persistent storage area only if desired. Otherwise, it can be
-%stored locally on optional SSDs for use by analysis routines eliminating the
-%need to use slower rotational media as the intermediary between a simulation
-%and the analysis code.
-
-%A second feature is the incorporation of a transaction like mechanism into the
-%IOD layer and a related epochs feature into the DAOS layer. This feature offers
-%a mechanism to avoid making incomplete data visible to other users and offer a
-%level of protection if the data can be stored on the optional SSDs. The
-%implications of how these concepts are designed are important to consider for
-%the final design. Because they control data visibility and manage faults for
-%output operations, they are key features.
+Along with the analysis of the published design documents, a discussion of the
+design philosophy representing the overall intent is presented. This informtion
+reprents information that may or moy not have been written down, but is the 
+intent of ultimate product. These insights were gained based on personal
+converations with the team members discussing some of the potential challenges
+with the design as written. These ideas are presented to give a fuller picture
+of where the project is going rather than dwelling on the limitations of the
+published documents.
 
@@
 These initial designs all use extra compute nodes to represent the data storage
-buffer given the lack of any hardware support for this functionality. The
-desired outcome of these initial studies is to motivate how such functionality
-might be incorporated and the potential benefits.  Later, these concepts were
-proposed to be incorporated as part of the IO
+buffer given the lack of any dedicated hardware support for this functionality.
+The desired outcome of these initial studies is to motivate how such
+functionality might be incorporated and the potential benefits.  Later, these
+concepts were proposed to be incorporated as part of the IO
 stack~\cite{bent:2012:challenges,bent:2012:burst-buffer}.  The current
@@
 the use of SSDs in the IOD layer may be problematic.  First, with these burst
-buffers being optional, the semantics of how IOD works must change to use DAOS
-to store the data directly rather than storing them locally until explicitly
-persisted by the user. The design of DAOS does not incorporate a high
-transaction rate mechanism. Conversely, it assumes that it will only be
+buffers being optional, the semantics must be clarified of how IOD works must
+change to use DAOS to store the data directly rather than storing them locally
+until explicitly persisted by the user. The design of DAOS does not incorporate
+a high transaction rate mechanism. Conversely, it assumes that it will only be
 involved when persisting a completed transaction and only for a fraction of the
@@
 to DAOS, this functionality may also be optional depending on the existence of
-the burst buffer. Consider the important functionality of data rearrangement
-and doing things like changing the fast array dimension.
+the burst buffer or slow the IO path to accommodate the data processing.
+Consider the important functionality of data rearrangement and doing things
+like changing the fast array dimension.
 
@@
 
-In the case of the IOD design, it forces a staging area deployment and then
-shares the staging area across all users simultaneously. This is unlikely to be
-useful because of the limited compute and communication capacity to spare to
-perform these operations at a bottleneck in the IO path. The use of a separate
-data staging area intentionally separate from the IO path allows using
+In the case of the written IOD design, it forces a staging area deployment and
+then shares the staging area across all users simultaneously. This is unlikely
+to be useful because of the limited compute and communication capacity to spare
+to perform these operations at a bottleneck in the IO path. The use of a
+separate data staging area intentionally separate from the IO path allows using
 operators on limited resources leaving the IO path clear for strictly data
@@
 can address the performance concerns they are intended to cover.
+
+\subsection{Design Philosophy}
+The burst buffers design, as presented in the IOD documents, limits the
+placement of the function operators and SSDs for buffers to the IO nodes. The
+team does acknowledge the limitations of this design and intend to ultimately
+focus on spreading the IOD from the IO nodes into the compute area as well.
+This is intended both to help address the limitations of the IO bandwidth
+and compute capaibility of these few nodes for data processing, but also to
+take advantage of new layers in the storage hierarchy. By incorporating NVRAM
+into compute nodes, new options for buffering data prior to being moved to
+centralized storage become available. This lessens the impact of some operators
+while offering additional options for places to store data.
+
+Burst buffers being optional is a high level goal, but not one considered at a
+detailed level with the design. While this forces some serious thought on how
+to adapt the functionality proposed for IOD to work directly on DAOS will have
+to be considered. With the additional desire to support using compute node
+resources for these operations, serious work will be required to make a fully
+functional end-to-end IOD layer implementation for a production system.
+
+Another concern that is acknowledged, but no thought as been applied to is the
+requirement that a single IOD process be the master for any operation. Should
+the number of concurrent applications exceed the available nodes, sharing an
+IOD process will be required. Since partitioning of the IOD processes for
+exclusive use by particular applications is the proposed operating mode, should
+insufficient IOD resources be available, either a job could be delayed or IOD
+resources could be reallocated from a different process could be redeployed for
+use by the new job. These sorts of consideraitons still need to be made for a
+full production system.
 
@@
 
+\subsubsection{Design Philosophy}
+Undocumented, but inherent in the design of these transactions is how faults
+are detected. The initial design assumes that the current Lustre fault
+detection mechanism that can determine if a process or node is no longer
+reachable is used. This defines how a fault will be detected and what will trigger a passive fault recovery (i.e., transaction abort).
+
+A new term, ``Primary'', is used to define the process(es) that contact the IOD
+layer directly to open a container. Unfortunately, this term is never used nor
+defined within any of the published documents. There are some very positive and
+negative performance implications of this approach. On the positive side, like
+PVFS, it is possible for a single process to open a container and share that
+handle with other processes reducing the load on the IOD layer. The key idea is
+that the single leader vs. multi-leader transaction mode is automatically defined based on the number of Primary processes that open a container. This greatly
+simplifies any additional client-side load in using the different styles of transactions. Also on the positive side, it offers the ability to have a subset of
+processes that are considered transaction participants affording reduced load
+on the IOD processes and more localized transaction management on the client
+side. On the negative side, for a fully IOD managed transaction, ALL processes
+must open the container forcing a potentially immense number of simultaneous
+calls to the IOD layer to open a container. This sort of onslaught is what is
+dealt with on metadata servers today with mass file opens. The reality is that
+this ends up having the same sort of performance implications as distributed
+denial of service attacks on the IOD processes.
+
+Ultimately, with the passive detection of faults for Primaries, the transaction
+mechanism can work very well. A mostly unstated restriction that will be
+relaxed in the next phase is that every sequential transaction on a container
+is considered dependent on the earlier transaction. Should one output be
+delayed and the subsequent five succeed, when the delayed process finally
+fails, all six transactions are rolled back. The thought of using this
+mechanism to store subsequent checkpoint outputs in the same container to both
+save space, but not care if one fails, cannot work in the current form. This
+has been acknowledged and is planned on being relaxed in phase two.
+
 \subsection{DAOS Epochs}
@@
 \subsection{Metadata Management}
-While a central design goal of the Fast Forward project to eliminate the
-requirement for a metadata server as a core component, it was not fully
-accomplished.  This idea is not new and has been explored as part of the Light
-Weight File Systems project~\cite{oldfield:lwfs}. In LWFS, the metadata service
-is explicitly limited to a user task with the storage layer limited to data
+Eliminating metaata management as a special case and instead treating it just
+as data is a central design goal of the Fast Forward project. Unfortunately,
+it is not completely achieved.
+
+Eliminating metadata as a core component of a file system is not new. It has
+been explored as part of the Light Weight File Systems
+project~\cite{oldfield:lwfs}. In LWFS, the metadata service is explicitly
+limited to a user task with the storage layer limited to data
 storage/retrieval, authorization, and authentication. This approach proved
-workable. For the Fast Forward project, while a similar goal is desired, it is
-not completely accomplished.
+workable. Using this hybrid approach is less common and introduces other
+issues.
 
-The purpose behind the goal is to eliminate the serialized bottleneck caused by
-having a centralized metadata service. Much of the design of IOD with objects
-and containers does work to eliminate these bottlenecks.  However, IOD does not
-go fully to a no metadata service model. Instead, the IOD layer describes a
-key-value metadata service it provides. This service maintains four different
-items:
+IOD and DAOS both share a philosophy that they will have to maintain the
+metadata about how objects are striped and where they are placed. It was
+decided that the additional data to track the namespace as well was a small
+enough additional increment that it was worth maintaining within the system.
 
-\begin{itemize}
-\item The list of IOD objects within the container
-\item The mapping from each IOD object to the corresponding DAOS object.
-\item The sharding and striping of the IOD object
-\item The maximum valid offset of the object
-\end{itemize}
-
-While a pure no metadata model would be ideal for performance, this current
-proposal has a few challenges. By providing this basic functionality, metadata
-management is split between the user layer and the IOD layer. Consider the
-following items. First, there is no list of containers. This is closer to the
-no metadata ideal, but given the rest of the functionality, it exports to the
-user level a requirement to track what containers are in the file system.
-Second, given the flattening that occurs between IOD and DAOS, the mapping from
-IOD objects to DAOS objects is unlikely to be 1-to-1. This is likely a list
-instead. Further, given the ability to re-stripe in the IOD layer, the list is
-not a list of objects completely included in the IOD object, but instead
-objects from which part of the data came from some listed DAOS object.  Third,
-sharding IOD objects makes sense based on the compute process connection a
-shard has with the complete object.  Striping of an IOD object makes sense in
-connection with the DAOS object(s) underneath, but striping within the IOD
-layer is not discussed. Further, using striping from clients into the IOD layer
-introduces a level of coordination that IOD seeks to avoid.  Fourth, given the
-previous items, there is ambiguity with the maximum valid offset of the object.
-This could be for the shard or for the collection of shard objects that
-represent a distributed variable. Ideally, either a robust metadata management
-system should be incorporated into the system, preferably separate from the IOD
-layer or all of this functionality should be left to the user layer. The only
-potential challenge is determining the container for a file open command and
-what the transaction ID should be. This limited operation should be a
-straightforward search using a hash or similar fast mechanism for examining a
-list of container names and/or object IDs.
+\subsubsection{Design Philosophy}
+While the metadata design is not fully defined, there are a few things that
+are intended as part of this. For example, there is a standard, well known
+container that is the system metadata. This includes the list of all other
+containers. This container is treated like any other data in the system and
+striped as appropriate at both the IOD and the DAOS layers. Unfortunately, this
+still couples the metadata to a single object that must serialize access. If
+the metadata, including information about striping and other data layout
+operations were separated completely from the data path, more scalable
+throughput could be achieved. The real challenge of this is introduced by the
+the IOD, DAOS, and VOSD layers. Each of these requires some different metadata
+storage and the migration is transparent to the user. Supporting fully
+independent metadata with this model is difficult, if not impossible. Serious
+thought on how to do this outside the data path effectively should be
+considered for phase two.
 
@@
 concurrency.
-
-For the IOD proposal, the service needs some adjustments to be generally
-useful.  First, there needs to be a way to query the list of available
-containers. This will support an analysis application using data generated by
-the simulation code. This use case is a stated motivating example for the Fast
-Forward project. Second, the maximum transaction ID for a container is
-mentioned as being maintained by the IOD metadata service, but when and how
-this is done is never mentioned.  Third, the definition of an object needs to
-be made more clear. In some cases, it seems that an object is a globally
-distributed object is represented by a collection of shards. In other cases, it
-is a single shard.
 
@@
 
-First, \DDT has a timeout mechanism to detect failures and offer an ability to
-recover and clean up the incomplete transactions. IOD's transactions are
-asynchronous precluding a relatively simple, short timeout value to detect
-failures. Unfortunately, at this point, there is no mechanism in IOD's
-transaction handling to detect a fault and clean up any incomplete operations.
-While this is intended for phase 2, some initial suggestion on how this might
-be handled would give some confidence that the asynchronous approach is viable.
+First, should a transaction on a container not complete, all subsequent
+transactions on that container, even if they are complete and correct, will not
+be accessible. Once the transaction fails, all of these subsequent transactions
+will be aborted. While this is being reconsidered for phase two, it does
+complicate the interface to offer the additional transaction mode.
 
-Second, should a transaction on a container not complete, all subsequent
-transactions on that container, even if they are complete and correct, will not
-be accessible.
-
-Third, in the single leader model, if the process that manages the transaction
-were to fail, there is no ability to abort that transaction. The data will not
-be made available because the transaction is still listed as in progress. The
-problem is that without a way to clean up this failed transaction, no newer
-versions of the same container can be made readable. In reality, the single
-leader model would ideally outsource the transaction processing to something
-like \DDT on the client side. This does not address the failure case though.
-
-Fourth, in the multi-leader model, using a count of client connections to
+Second, in the multi-leader model, using a count of client connections to
 determine if transaction is complete is problematic.  First, should this be in
@@
 
-Consider a shared file system across an HPC data center. The current design
-maintains the metadata in the IOD layer localized to a single machine
-effectively making the data inaccessible from another platform. This requires
-either the IOD layer be shared across machines, something clearly beyond the
-design scope, or that an additional metadata service that extends the minimal
-metadata service provided in the IOD layer. This higher level metadata service
-would be required to be reachable from any machine that shares the file system.
-As explained above, having this client-side only metadata service introduces
-additional functionality requirements.
+At a broader level, there are some concerns that were partially clarified
+through conversations with the team.  Consider a shared file system across an
+HPC data center. The current design maintains the metadata in its own
+container. Since copying data from the IOD layer to the DAOS layer requires an
+explicit persist call, how and when synchronizing the metadata across the 
+layers and potentially across machines occurs is undefined. By delaying the
+synchronization until an explicit persist is called will reduce the update
+frequency, it does delay the visibility of data. Ideally, the metadata object
+would need to be automatically persisted every time a container transaction is
+persisted to the DAOS layer.
+
+The implication of this is that every transaction persist is double operation
+to account for the metadata persist. More importantly, the IOD-layer version of
+the metadata container may contain readable transactions that have not been
+persisted to the DAOS layer. How to handle this inconsistency between the two
+layers still needs to be explored.
 

### 2014-03-16 paper.tex
version sent to potential team mates.

@@
-\documentclass[conference]{sig-alt-gov2}
+\documentclass[conference]{IEEEtran}
 \pdfpagewidth=8.5in
@@
 
+\addtolength{\parskip}{-0.04in}
+
 % in order for balance columns to work, it has to be before begin document...
@@
 
-\conferenceinfo{MSST'14,} {November 17, 2013, Denver, Colorado, USA.}
-\CopyrightYear{2014}
+%\conferenceinfo{MSST'14,} {November 17, 2013, Denver, Colorado, USA.}
+%\CopyrightYear{2014}
 %\crdata{978-1-4503-1103-8/11/11}
-\clubpenalty=10000
-\widowpenalty = 10000
+%\clubpenalty=10000
+%\widowpenalty = 10000
 
@@
 
-\numberofauthors{3}
 \author{
-\alignauthor Jay Lofstead\\
-       \affaddr{Sandia National Laboratories}\\
-       \email{gflofst@sandia.gov}
-\alignauthor Ivo Jimenez\\
-       \affaddr{University of California, Santa Cruz}\\
-       \email{ivo@cs.ucsc.edu}
-\alignauthor Carlos Maltzahn\\
-       \affaddr{University of California, Santa Cruz}\\
-       \email{carlosm@soe.ucsc.edu}
-%\and  % use '\and' if you need 'another row' of author names
+\IEEEauthorblockN{Jay Lofstead}
+\IEEEauthorblockA{Sandia National Laboratories\\gflofst@sandia.gov}
+\and
+\IEEEauthorblockN{Ivo Jimenez}
+\IEEEauthorblockA{University of California, Santa Cruz\\ivo@cs.ucsc.edu}
+\and
+\IEEEauthorblockN{Carlos Maltzahn}
+\IEEEauthorblockA{University of California, Santa Cruz\\carlosm@soe.ucsc.edu}
 }
@@
 While these and other their contributions are valuable, similar efforts made
-elsewhere may offer attractive alternatives or differing semantics that would
+elsewhere may offer attractive alternatives or differing semantics that could
 yield a more feature rich environment with little to no additional overhead.
@@
 
-\category{D.4}{Software}{Operating Systems}
-\category{D.4.7}{Operating Systems}{Organization and Design}[hierarchical design]
+%\category{D.4}{Software}{Operating Systems}
+%\category{D.4.7}{Operating Systems}{Organization and Design}[hierarchical design]
 
-\terms{Design, Performance}
+%\terms{Design, Performance}
 
@@
 control. Through the use of a copy-on-write style mechanism, multiple versions
-of a file can be stored in potentially greatly reduced space. By assuming the
-client interface will be through an IO library affords a more complicated
+of a file can be stored in potentially greatly reduced space. It assumes the
+client interface will be through an IO library affording a more complicated
 interface that offers richer functionality without requiring user code changes.
-By managing data in the IOD layer rather than requring writing to centralized
-storage will better support the performnce requirements of integrated
-application workflows. At a more detailed view, the various layers of the IO
-stack each contribute different functionality and performance implications.
+Managing most data access in a machine-local layer rather than requiring
+writing to centralized storage will better support the performance and energy
+requirements of extreme scale integrated application workflows. At a more
+detailed view, the various layers of the IO stack each contribute different
+functionality and performance implications.
 
@@
 dispatcher (IOD) serves as the primary storage interface for the IO stack and
-is the only way to reach the persistent storage array in lower layers.
+is the only way to reach the persistent storage array in lower layers. The
+Data Access Object Storage (DAOS) layer serves as the persistent storage
+interface. At the bottom is the Versioning Object Storage Device (VOSD). It
+serves as the interface for storing objects of all types efficiently.
+
+The focus of this paper is primarily the IOD layer given the critical role it
+has in the performance and functionality of the entire stack. Most of the key
+features explored in this paper all have a strong presence in the IOD layer
+motivating the focus of this examination.
 
@@
 processing operations can be placed in the IOD. These operations are intended
-to offer data rearrangement, filtering, and similar operations prior to it
+to offer data rearrangement, filtering, and similar operations prior to data
 reaching the central storage array.
@@
 While these ideas are not necessarily new, they are new twists on best of class
-efforts for these techologies. For example, offloading the collective two-phase
-data sieving from the compute nodes to reorganize data has proven effective at
-reducing the total time for writing data due to fewer participants involved in
-the communication patterns~\cite{lofstead:2011:nessie-staging}.  Beyond these
-broad items, there are many important details some of which are examined in
-more detail below.
-
-The bottom two layers are the Data Access Object Storage (DAOS) and Versioning
-Object Storage Device (VOSD). DAOS is the typical interface that represents
-the individual objects stored including the semantics related to ``files''.
-VSOD is the actual storage mechanism for the objects defined at the DAOS layer.
-
-The focus of this paper is primarily the IOD layer given the critical role it
-has in the performance and functionality of the entire stack. Most of the key
-features explored in this paper all have a strong presence in the IOD layer
-motivating the focus of this examination.
+efforts for these technologies. For example, offloading the collective
+two-phase data sieving from the compute nodes to reorganize data has proven
+effective at reducing the total time for writing data due to fewer participants
+involved in the communication patterns~\cite{lofstead:2011:nessie-staging}.
+Beyond these broad items, there are many important details some of which are
+examined in more detail below.
 
 Along with the analysis of the published design documents, a discussion of the
-design philosophy representing the overall intent is presented. This informtion
-reprents information that may or moy not have been written down, but is the 
-intent of ultimate product. These insights were gained based on personal
-converations with the team members discussing some of the potential challenges
-with the design as written. These ideas are presented to give a fuller picture
-of where the project is going rather than dwelling on the limitations of the
-published documents.
+design philosophy representing the overall intent is presented. This
+information represents information that may or may not have been written down,
+but is the intent of ultimate product. These insights were gained based on
+personal conversations with the team members discussing some of the potential
+challenges with the design as written. These ideas are presented to give a
+fuller picture of where the project is going rather than dwelling on the
+limitations of the published documents.
 
@@
 these devices as optional. Unfortunately, both incorporating burst buffers and
-the use of SSDs in the IOD layer may be problematic.  First, with these burst
-buffers being optional, the semantics must be clarified of how IOD works must
-change to use DAOS to store the data directly rather than storing them locally
-until explicitly persisted by the user. The design of DAOS does not incorporate
-a high transaction rate mechanism. Conversely, it assumes that it will only be
-involved when persisting a completed transaction and only for a fraction of the
-total transactions created. Further, with function shipping offering the
-ability to change how the data is stored and arranged prior to it being written
-to DAOS, this functionality may also be optional depending on the existence of
-the burst buffer or slow the IO path to accommodate the data processing.
-Consider the important functionality of data rearrangement and doing things
-like changing the fast array dimension.
+the use of SSDs in the IOD layer may be problematic.  First, the IOD and DAOS
+designs currently are written assuming burst buffers. This means that the
+bulk of IO operations will only hit the IOD layer and proposed functionality,
+such as the function operators, do not discuss the impact on the DAOS layer or
+the operators themselves should a burst buffer not be available.  Consider the
+important functionality of data rearrangement and doing things like changing
+the fast array dimension on shared, spinning media.  The design of DAOS assumes
+that it will only be involved when persisting a completed transaction and only
+for a fraction of the total transactions created.
 
@@
 potential costs and advantages of where to place operators similar to the IOD
-proposed function shipping. There are two key takeaways from that work. First,
+proposed function shipping. There are two key takeaways from PreDatA. First,
 placement matters.  Depending on the communication intensity vs. computation
@@
 
-In the case of the written IOD design, it forces a staging area deployment and
-then shares the staging area across all users simultaneously. This is unlikely
-to be useful because of the limited compute and communication capacity to spare
-to perform these operations at a bottleneck in the IO path. The use of a
-separate data staging area intentionally separate from the IO path allows using
-operators on limited resources leaving the IO path clear for strictly data
-movement.
+In the case of the written IOD design, it describes a fixed-sized staging area
+shared across all users simultaneously.  This is unlikely to be useful because
+of the limited compute and communication capacity to spare to perform these
+operations at a bottleneck in the IO path.  The use of a separate data staging
+area intentionally separate from the IO path allows using operators on limited
+resources leaving the IO path clear for strictly data movement. A nuance of
+this design is discussed in the Design Philosophy below.
 
@@
 team does acknowledge the limitations of this design and intend to ultimately
-focus on spreading the IOD from the IO nodes into the compute area as well.
-This is intended both to help address the limitations of the IO bandwidth
-and compute capaibility of these few nodes for data processing, but also to
-take advantage of new layers in the storage hierarchy. By incorporating NVRAM
-into compute nodes, new options for buffering data prior to being moved to
-centralized storage become available. This lessens the impact of some operators
-while offering additional options for places to store data.
+focus on spreading the IOD layer from the IO nodes into the compute area as
+well.  This is intended both to help address the limitations of the IO
+bandwidth and compute capability of these few nodes for data processing, but
+also to take advantage of new layers in the storage hierarchy. By incorporating
+NVRAM into compute nodes, new options for buffering data prior to being moved
+to centralized storage become available and addresses some of the concerns
+about SSD performance. This lessens the impact of some operators while offering
+additional options for places to store data.
 
@@
 IOD process will be required. Since partitioning of the IOD processes for
-exclusive use by particular applications is the proposed operating mode, should
+exclusive use by particular applications is the assumed operating mode, should
 insufficient IOD resources be available, either a job could be delayed or IOD
 resources could be reallocated from a different process could be redeployed for
-use by the new job. These sorts of consideraitons still need to be made for a
+use by the new job. These sorts of considerations still need to be made for a
 full production system.
@@
 Undocumented, but inherent in the design of these transactions is how faults
-are detected. The initial design assumes that the current Lustre fault
-detection mechanism that can determine if a process or node is no longer
-reachable is used. This defines how a fault will be detected and what will trigger a passive fault recovery (i.e., transaction abort).
+are detected. The initial design assumes the current Lustre fault detection
+mechanism that can determine if a process or node is no longer reachable. This
+defines how a fault will be detected and what will trigger a passive fault
+recovery (i.e., transaction abort).
 
@@
 handle with other processes reducing the load on the IOD layer. The key idea is
-that the single leader vs. multi-leader transaction mode is automatically defined based on the number of Primary processes that open a container. This greatly
-simplifies any additional client-side load in using the different styles of transactions. Also on the positive side, it offers the ability to have a subset of
-processes that are considered transaction participants affording reduced load
-on the IOD processes and more localized transaction management on the client
-side. On the negative side, for a fully IOD managed transaction, ALL processes
-must open the container forcing a potentially immense number of simultaneous
-calls to the IOD layer to open a container. This sort of onslaught is what is
-dealt with on metadata servers today with mass file opens. The reality is that
-this ends up having the same sort of performance implications as distributed
-denial of service attacks on the IOD processes.
+that the single leader vs. multi-leader transaction mode is automatically
+defined based on the number of Primary processes that open a container. This
+greatly simplifies any additional client-side load in using the different
+styles of transactions. Also on the positive side, it offers the ability to
+have a subset of processes that are considered transaction participants
+affording reduced load on the IOD processes and more localized transaction
+management on the client side. On the negative side, for a fully IOD managed
+transaction, ALL processes must open the container forcing a potentially
+immense number of simultaneous calls to the IOD layer to open a container. This
+sort of onslaught is what is dealt with on metadata servers today with mass
+file opens. The reality is that this ends up having the same sort of
+performance implications as distributed denial of service attacks on the IOD
+processes.
 
@@
 \subsection{Metadata Management}
-Eliminating metaata management as a special case and instead treating it just
+Eliminating metadata management as a special case and instead treating it just
 as data is a central design goal of the Fast Forward project. Unfortunately,
@@
 While the metadata design is not fully defined, there are a few things that
-are intended as part of this. For example, there is a standard, well known
-container that is the system metadata. This includes the list of all other
-containers. This container is treated like any other data in the system and
-striped as appropriate at both the IOD and the DAOS layers. Unfortunately, this
-still couples the metadata to a single object that must serialize access. If
-the metadata, including information about striping and other data layout
-operations were separated completely from the data path, more scalable
-throughput could be achieved. The real challenge of this is introduced by the
-the IOD, DAOS, and VOSD layers. Each of these requires some different metadata
-storage and the migration is transparent to the user. Supporting fully
-independent metadata with this model is difficult, if not impossible. Serious
-thought on how to do this outside the data path effectively should be
-considered for phase two.
+are intended. For example, there is a standard, well-known container that is
+the system metadata. This includes the list of all other containers. This
+container is treated like any other data in the system and striped as
+appropriate at both the IOD and the DAOS layers. Unfortunately, this still
+couples the metadata to a single object that must serialize access. If the
+metadata, including information about striping and other data layout operations
+were separated completely from the data path, more scalable throughput could be
+achieved. The real challenge of this is introduced by the IOD, DAOS, and VOSD
+layers. Each of these requires some different metadata storage and the
+migration is transparent to the user. Supporting fully independent metadata
+with this model is difficult. Serious thought on how to do this effectively
+outside the data path should be considered for phase two.
 
-The \DDT project created a simplistic model for a metadata service that has
-different features. It is described below and has been published
-previously~\cite{lofstead:2012:txn-metadata}. Some of the major differences
-include the following. First, \DDTns's metadata service has a way to get a list
-of the objects in the metadata service. Since this was intended as a test
-system to examine \DDT performance, we did not put the effort into a making it
-a robust system. While it does not address having a file equivalence like the
-container concept, it does introduce a way to figure out what is available.
-Stored along with this object description is a list of the global dimensions
-for array objects. One observation made as part of creating this service is that
-this is insufficient for many engineering codes that use non-regular meshes.
-Second, for each object, a list of the equivalent to shards are kept with a
-link back to the master object and the offset and size for each dimension for
-that shard. This approach is not scalable because of the explosion in the
-process count. Using a striping approach like IOD does is superior. Third,
-there is no mapping to another layer with different object layout and counts.
-
-Based on the lessons from the \DDT metadata service construction and the prior
+Based on the lessons from the \DDT metadata
+service~\cite{lofstead:2012:txn-metadata} construction and the prior
 experiments with LWFS, having a completely separate metadata service is
 workable. Rather than making it a bottleneck in the IO path, it is another
-service that users must interact with if they need those services. Otherwise,
-it is 100\% out of the way. Users can manage everything by maintaining the
-metadata including the list of objects themselves. However, there are drawbacks
-to this 100\% client-side approach.
+service that users must interact with if they need those services.  Users can
+manage everything by maintaining the metadata including the list of objects
+themselves. However, there are drawbacks to this 100\% client-side approach.
 
@@
 mechanism to manage the atomic operation. Overall, while additional work is
-required to maintain a client-side only metadata service, eliminating any
+required to maintain a client-side only metadata service, it eliminates any
 potential bottlenecks related to updating metadata related to the object
-storage are eliminated. The burden of tracking striping and other metadata that
-has traditionally been part of the metadata associated with the file system
-will have to be maintained by the object storage service. The lack of a
-centralized, serialized bottleneck to store that information improves
-concurrency.
+storage. The burden of tracking striping and other metadata that has
+traditionally been part of the metadata associated with the file system will
+have to be maintained by the object storage service. The lack of a centralized,
+serialized bottleneck to store that information improves concurrency.
 
@@
 The example metadata and data storage services created as part of the \DDT
-project lack one of the key transaction management features used by IOD.
-The major difference is that there are no dependencies between transactions
-that prevent visibility should an older version be incomplete. This additional,
-intentional requirement by IOD offers different functionality than \DDTns's
-example services. In the case of \DDTns, the functionality is more minimal, but
-also avoids some of the concerns outlined below.
+project have no dependencies between transactions that prevent visibility
+should an older version be incomplete. This additional, intentional requirement
+by IOD offers different functionality than \DDTns's example services. In the
+case of \DDTns, the functionality is more minimal, but also avoids some of the
+concerns outlined below.
 
@@
 Table~\ref{tab:scaling}. For comparison, consider the Number of
-Sub-Coordinators equivalent to an IOD process. The Processes Per
-Sub-Coordinator represents the number of clients that use a particular IOD. For
-these tests, we maintained a balanced distribution and always used at least two
+Sub-Coordinators equivalent to IOD processes. The Processes Per Sub-Coordinator
+represents the number of clients that use a particular IOD process. For these
+tests, we maintained a balanced distribution and always used at least two
 sub-coordinators to slow down the processing.
@@
 \label{fig:performance}
-\vspace{-0.15in}
+%\vspace{-0.15in}
 \end{figure}
@@
 
-At a high level, both \DDT and the IOD transactions have the same high-level
-design. In both cases, a hierarchical model is employed. In the case of \DDTns,
-it is a purely client-side tree using semi-synchronous messaging. The messaging
-itself, in the current implementation, uses asynchronous MPI messages. The
-synchronous component comes from the timeout mechanism used to detect faults.
-It forces a level of coordination and synchronization for the protocol. For
-IOD, it is a server-side tree and fully asynchronous. In both cases, there is a
-master in charge of managing the transaction and a collection of workers that
-aggregate into the master through second-level leaders. Beyond that, there are
-some significant differences. Some of the different choices made by IOD raise
-some possible concern.
+At a high level, both \DDT and the IOD transactions have the same design. In
+both cases, a hierarchical model is employed. In the case of \DDTns, it is a
+purely client-side tree using semi-synchronous messaging. The messaging itself,
+in the current implementation, uses asynchronous MPI messages. The synchronous
+component comes from the timeout mechanism used to detect faults.  It forces a
+level of coordination and synchronization for the protocol. For IOD, it is a
+server-side tree and fully asynchronous relying on the existing Lustre fault
+detection mechanism for failure detection. In both cases, there is a master in
+charge of managing the transaction and a collection of workers that aggregate
+into the master through second-level leaders. Beyond that, there are some
+significant differences. Some of the different choices made by IOD raise some
+possible concern.
 
-First, should a transaction on a container not complete, all subsequent
-transactions on that container, even if they are complete and correct, will not
-be accessible. Once the transaction fails, all of these subsequent transactions
-will be aborted. While this is being reconsidered for phase two, it does
-complicate the interface to offer the additional transaction mode.
-
-Second, in the multi-leader model, using a count of client connections to
-determine if transaction is complete is problematic.  First, should this be in
-an environment where lost or repeated network messages are possible, then the
-count could be wrong inappropriately. This is a general problem and unlikely
-in the current environment, but is a concern on less integrated platforms.
-The idea of a count precludes any tracking of expected messages from sources
-introducing a potential hole. Second, the aggregation of completion messages
-are based on the local aggregation point reporting to the transaction
-leader the state.  There is no ability to detect or recover from a failure of
-this IOD process that would report to the transaction leader. Third, because a
-simple count of participating processes is sent with the beginning of the
-transaction, each process is strictly limited to a single output operation.
-This precludes a process writing multiple array shards, for example. To put
-this in concrete terms, it means that a file could only contain a single,
-globally distributed variable (at the root level of HDF5) OR an attribute per
-process. If this count actually represents a begin/end transaction pair
-instead, it suffers from a lack of accountability for missing object (shard)
-writes. All it would indicate is that a process started and ended a transaction
-connection. There is no validation that a process has done everything required
-of it.
+In the multi-leader model, using a count of client connections to determine if
+transaction is complete is problematic.  First, the aggregation of completion
+messages are based on the local aggregation point reporting to the transaction
+leader.  There is no ability to detect or recover from a failure of this IOD
+process that would report to the transaction leader. The current assumption
+that the IOD layer will be an MPI program is not workable until MPI can fully
+support fault detection and recovery including removing and adding processes.
+Second, because a simple count of participating processes is sent with the
+beginning of the transaction, each process is strictly limited to a single
+output operation.  This precludes a process writing multiple array shards, for
+example. To put this in concrete terms, it means that a file could only contain
+a single, globally distributed variable (at the root level of HDF5) OR an
+attribute per process. If this count actually represents a begin/end
+transaction pair instead, it suffers from a lack of accountability for missing
+object (shard) writes. All it would indicate is that a process started and
+ended a transaction connection. There is no validation that a process has done
+everything required of it.
 
@@
 \label{fig:optimized-protocol}
-\vspace{-0.25in}
+%\vspace{-0.25in}
 \end{figure}
 
-\DDT has addressed this count issue in a couple of ways. First, the
+\DDT has addressed theses issues in a couple of ways. First, the
 sub-coordinators each have a list of processes from which they expect messages.
@@
 
-\subsection{Comparison to Other Protocols}
-Alternatives, such as Paxos~\cite{Lamport:1998:paxos} algorithms like
-ZooKeeper~\cite{Hunt:2010:zookeeper}, suffer from two limitations making them
-unsuitable for this environment. First, the distributed servers in Paxos
-systems are all distributed copies of each other that eventually become
-consistent. Given the scale we wish to address, a single node's memory is
-unlikely to be able to hold all of the data necessary for many operations at
-scale. They also do not have a guarantee for when consensus will be achieved
-without using slower synchronous calls. For the tight timing we wish to
-support, we need guarantees of when a consistent state has been achieved.
-Second, these systems also all assume that updates are initiated from a single
-client process rather than a parallel set of processes as is the standard in
-HPC environments.
-
-\DDT uses a second layer of coordination on the client side that greatly
-increases the scalability by consolidating messages from clients into unique
-sets prior to sending to the overall coordinator. A gossip
-protocol~\cite{ganesh:2003:gossip-protocols} may appear sufficient for this
-purpose, but the delay of eventual consistency is strictly avoided with this
-protocol to ensure guarantees at particular states in the code. For example, if
-a failure occurs, the global understanding of the role of all processes is
-required in order for effective communication to occur for operations like
-creating sub-transactions or voting. In this case, the protocol can offer
-stronger statements about consistency than these protocols offer.  These
-features offer a way to easily scale the transaction protocol given the
-guarantees we wish to offer. The IOD approach does nothing to address these
-concerns.
-
-Another effort to offer consistency and data integrity for the ZFS file
-system~\cite{zhang:2010:zfs} covers some of the same territory. Instead of a
-focus on the processes all having a notion of completion as a transaction, this
-work focuses on the integrity of the data movement operations. We view this
-work as something that should be considered hand-in-hand with a transaction
-approach to ensure the integrity of the movement of the data in addition to the
-agreement of processes about the successful completion of a parallel operation.
+%\subsection{Comparison to Other Protocols}
+%Alternatives, such as Paxos~\cite{Lamport:1998:paxos} algorithms like
+%ZooKeeper~\cite{Hunt:2010:zookeeper}, suffer from two limitations making them
+%unsuitable for this environment. First, the distributed servers in Paxos
+%systems are all distributed copies of each other that eventually become
+%consistent. Given the scale we wish to address, a single node's memory is
+%unlikely to be able to hold all of the data necessary for many operations at
+%scale. They also do not have a guarantee for when consensus will be achieved
+%without using slower synchronous calls. For the tight timing we wish to
+%support, we need guarantees of when a consistent state has been achieved.
+%Second, these systems also all assume that updates are initiated from a single
+%client process rather than a parallel set of processes as is the standard in
+%HPC environments.
+%
+%\DDT uses a second layer of coordination on the client side that greatly
+%increases the scalability by consolidating messages from clients into unique
+%sets prior to sending to the overall coordinator. A gossip
+%protocol~\cite{ganesh:2003:gossip-protocols} may appear sufficient for this
+%purpose, but the delay of eventual consistency is strictly avoided with this
+%protocol to ensure guarantees at particular states in the code. For example, if
+%a failure occurs, the global understanding of the role of all processes is
+%required in order for effective communication to occur for operations like
+%creating sub-transactions or voting. In this case, the protocol can offer
+%stronger statements about consistency than these protocols offer.  These
+%features offer a way to easily scale the transaction protocol given the
+%guarantees we wish to offer. The IOD approach does nothing to address these
+%concerns.
+%
+%Another effort to offer consistency and data integrity for the ZFS file
+%system~\cite{zhang:2010:zfs} covers some of the same territory. Instead of a
+%focus on the processes all having a notion of completion as a transaction, this
+%work focuses on the integrity of the data movement operations. We view this
+%work as something that should be considered hand-in-hand with a transaction
+%approach to ensure the integrity of the movement of the data in addition to the
+%agreement of processes about the successful completion of a parallel operation.
 

### 2014-03-18 paper.tex
added Carlos's ref for Ceph and Gary Grider's comments

@@
 and performance to adequately serve the needs of an extreme scale platform. To
-address these limitations, the US Department of Energy commissioned an effort
-to develop a design and prototype for an IO stack suitable for the extreme
-scale environment. This is a joint effort led by the Intel Lustre team, Los
-Alamos National Laboratory, EMC, DDN, and the HDF Group. This team has
-developed a specification set~\cite{fastforward:2014:docs} for a future IO
-stack to address the identified challenges. The first phase recently completed
-with a second phase getting underway. The core focus of the first phase was
-basic functionality and design. The second phase will refine this design
-incorporating fault recovery and other features missing from the first phase.
+address these limitations, a joint effort between the US Department of Energy's
+Office of Advanced Simulation and Computing and Advanced Scientific Computing
+Research commissioned an effort to develop a design and prototype for an IO
+stack suitable for the extreme scale environment. This is a joint effort led by
+Lawrence Livermore National Laboratory, with the DOE Data Management Nexus
+leads Rob Ross and Gary Grider as coordinators and contract lead Mark Gary. The
+participating labs are LLNL, SNL, LANL, ORNL, PNL, LBNL, and ANL. Additional
+industrial partners contracted include the Intel Lustre team, EMC, DDN, and the
+HDF Group. This team has developed a specification
+set~\cite{fastforward:2014:docs} for a future IO stack to address the
+identified challenges. The first phase recently completed with a second phase
+getting underway. The core focus of the first phase was basic functionality and
+design. Overriding many of the decisions during this and any subsequent phases
+is the reality of budgets. Placing GBs of NVRAM on every node, while a
+potentially advantageous approach, is not financially feasible. With this in
+mind, the second phase will refine this design incorporating fault recovery and
+other features missing from the first phase.
 
@@
 storage/retrieval, authorization, and authentication. This approach proved
-workable. Using this hybrid approach is less common and introduces other
-issues.
+workable. Using this hybrid approach is less common~\cite{weil:2006:ceph} and
+introduces other issues.
 

### 2014-03-21 paper.tex
added comments from Quincey and expanded acknowledgements.

@@
 
-While these and other their contributions are valuable, similar efforts made
+While these and their other contributions are valuable, similar efforts made
 elsewhere may offer attractive alternatives or differing semantics that could
@@
 separate from the compute nodes and the storage array. Communication intensive
-activities, such as two-phase, data sieving IO's data rearrangement process,
-can be moved to the IOD layer offloading the communication load from the
-compute nodes. IOD has three main purposes. First, if the optional burst buffer
-is available, it works as a fast cache absorbing write operations for the
-slower trickle out to the central storage array. It can also be used to
-retrieve objects from the central storage array for more efficient read
-operations and offer data filtering to make client reads more efficient.
-Second, it offers the transaction mechanism for controlling data set visibility
-and to manage faults that would prevent a data set from being used. Third, data
-processing operations can be placed in the IOD. These operations are intended
-to offer data rearrangement, filtering, and similar operations prior to data
-reaching the central storage array.
+activities, such as data rearrangement, can be moved to the IOD layer
+offloading the communication load from the compute nodes. IOD has three main
+purposes. First, if the optional burst buffer is available, it works as a fast
+cache absorbing write operations for the slower trickle out to the central
+storage array. It can also be used to retrieve objects from the central storage
+array for more efficient read operations and offers data filtering to make
+client reads more efficient.  Second, it offers the transaction mechanism for
+controlling data set visibility and to manage faults that would prevent a data
+set from being used. Third, data processing operations can be placed in the
+IOD. These operations are intended to offer data rearrangement, filtering, and
+similar operations prior to data reaching the central storage array.
 
@@
 Burst buffers being optional is a high level goal, but not one considered at a
-detailed level with the design. While this forces some serious thought on how
-to adapt the functionality proposed for IOD to work directly on DAOS will have
-to be considered. With the additional desire to support using compute node
-resources for these operations, serious work will be required to make a fully
-functional end-to-end IOD layer implementation for a production system.
+detailed level with the design. For example, if there is no burst buffer, all
+of the advanced functionality proposed for the IOD layer would have to work
+against the DAOS layer instead. For example, function shipping assumes it will
+operate on fast, local data within the IOD layer rather than against the
+globally shared DAOS layer. With the additional desire to support using compute
+node resources for these operations, serious work will be required to make a
+fully functional end-to-end IOD layer implementation for a production system.
 
-Another concern that is acknowledged, but no thought as been applied to is the
+Another concern that is acknowledged, but no thought has been applied to is the
 requirement that a single IOD process be the master for any operation. Should
@@
 system. The objects in each container represent different data within a file.
-The three initially defined object types are key-value store, array shard,
-and blob. The easiest way to understand these types is to evaluate these from
-the perspective of an HDF-5 file, the initial user interface layer. The
-key-value store represents a collection of attributes. The array shard
-represents part of a potentially multi-dimensional array. It is referred to as
+The three initially defined object types are key-value store, array, and blob.
+The easiest way to understand these types is to evaluate these from the
+perspective of an HDF-5 file, the initial user interface layer. The key-value
+store represents a collection of attributes or groups. The array represents
+part of a potentially multi-dimensional array. It is referred to as
 a shard because it is likely a small piece of a globally defined array. The
 blob represents a byte stream of arbitrary contents.  The fundamental
-difference between an array shard and a blob is that the array shard has
-metadata identifying its portion within the global, logical space while the
-blob is simply a 1-dimensional array of bytes that is not shared across IOD
-nodes.  Should an operation be deployed into IOD to manipulate an array within
-a container, it would operate on the array shards rather than on blobs. Given
-this context, the transactions come in two forms.
+difference between an array and a blob is that the array has metadata
+identifying its portion within the global, logical space while the blob is
+simply a 1-dimensional array of bytes that is not shared across IOD nodes.
+Should an operation be deployed into IOD to manipulate an array within a
+container, it would operate on the array rather than on blobs. Given this
+context, the transactions come in two forms.
 
@@
 transactions. In this case, when the transaction is created, a count of clients
-is provided to the IOD layer. As clients write to the container, the reference
-count is reduced. Once the count reaches 0, the transaction is automatically
-committed.
+is provided to the IOD layer. As clients commit their changes to the container,
+the reference count is reduced. Once the count reaches 0, the transaction is
+automatically committed.
 
@@
 
-A new term, ``Primary'', is used to define the process(es) that contact the IOD
-layer directly to open a container. Unfortunately, this term is never used nor
-defined within any of the published documents. There are some very positive and
-negative performance implications of this approach. On the positive side, like
-PVFS, it is possible for a single process to open a container and share that
-handle with other processes reducing the load on the IOD layer. The key idea is
-that the single leader vs. multi-leader transaction mode is automatically
-defined based on the number of Primary processes that open a container. This
-greatly simplifies any additional client-side load in using the different
-styles of transactions. Also on the positive side, it offers the ability to
-have a subset of processes that are considered transaction participants
-affording reduced load on the IOD processes and more localized transaction
-management on the client side. On the negative side, for a fully IOD managed
-transaction, ALL processes must open the container forcing a potentially
-immense number of simultaneous calls to the IOD layer to open a container. This
-sort of onslaught is what is dealt with on metadata servers today with mass
-file opens. The reality is that this ends up having the same sort of
-performance implications as distributed denial of service attacks on the IOD
-processes.
+The team uses the term ``Primary'' to define the process(es) that contact the
+IOD layer directly to open a container. Unfortunately, this term is never used
+nor defined within any of the published documents. There are some very positive
+and negative performance implications of this approach. On the positive side,
+like PVFS, it is possible for a single process to open a container and share
+that handle with other processes reducing the load on the IOD layer. The key
+idea is that the single leader vs. multi-leader transaction mode is
+automatically defined based on the number of Primary processes that start a
+transaction on a container. This greatly simplifies any additional client-side
+load in using the different styles of transactions. Also on the positive side,
+it offers the ability to have a subset of processes that are considered
+transaction participants affording reduced load on the IOD processes and more
+localized transaction management on the client side. On the negative side, for
+a fully IOD managed transaction, ALL processes must start the transaction on
+the container forcing a potentially immense number of simultaneous calls to the
+IOD layer to open a container. This sort of onslaught is what is dealt with on
+metadata servers today with mass file opens. The reality is that this ends up
+having the same sort of performance implications as distributed denial of
+service attacks on the IOD processes.
 
 Ultimately, with the passive detection of faults for Primaries, the transaction
-mechanism can work very well. A mostly unstated restriction that will be
-relaxed in the next phase is that every sequential transaction on a container
-is considered dependent on the earlier transaction. Should one output be
-delayed and the subsequent five succeed, when the delayed process finally
-fails, all six transactions are rolled back. The thought of using this
-mechanism to store subsequent checkpoint outputs in the same container to both
-save space, but not care if one fails, cannot work in the current form. This
-has been acknowledged and is planned on being relaxed in phase two.
+mechanism can work very well. A mostly unstated restriction that is being
+relaxed is that every sequential transaction on a container is considered
+dependent on the earlier transaction. Should one output be delayed and the
+subsequent five succeed, when the delayed process finally fails, all six
+transactions are rolled back. The thought of using this mechanism to store
+subsequent checkpoint outputs in the same container to both save space, but not
+care if one fails, cannot work in the current form. This has been acknowledged
+and is planned on being relaxed in phase two.
 
@@
 Eliminating metadata management as a special case and instead treating it just
-as data is a central design goal of the Fast Forward project. Unfortunately,
-it is not completely achieved.
+as data is a central design goal of the Fast Forward project. This is a hybrid
+approach to metadata management that is half-way between providing no inherent
+metadata support and having a fully integrated, but separate metadata
+manaegement system.
 
@@
 file systems requires some considerations. In this case, should these services
-get out of sync, three particular risks introduced.  First, a client could
+get out of sync, three particular risks are introduced.  First, a client could
 create an entry in the metadata service that does not correspond to any objects
@@
 service should be an atomic operation, but due to a lack of coordination, a
-window where the system is inconsistent appears.  Ultimately, the consistency
-semantics required must be determined. If a metadata service is required and it
-must be in sync with the object storage service, then additional work must be
-performed. In traditional file systems, the metadata and object storage updates
-are atomic. With decoupling metadata from object storage, should this atomicity
-still be desired, it requires both the ability for the services to participate
-in a task that is part of a larger atomic operation and a higher-level
-mechanism to manage the atomic operation. Overall, while additional work is
-required to maintain a client-side only metadata service, it eliminates any
-potential bottlenecks related to updating metadata related to the object
-storage. The burden of tracking striping and other metadata that has
-traditionally been part of the metadata associated with the file system will
-have to be maintained by the object storage service. The lack of a centralized,
-serialized bottleneck to store that information improves concurrency.
+window where the system is inconsistent appears.
+
+Ultimately, the consistency semantics required must be determined. If a
+metadata service is required and it must be in sync with the object storage
+service, then additional work must be performed. In traditional file systems,
+the metadata and object storage updates are atomic. With decoupling metadata
+from object storage, should this atomicity still be desired, it requires both
+the ability for the services to participate in a task that is part of a larger
+atomic operation and a higher-level mechanism to manage the atomic operation.
+
+Overall, while additional work is required to maintain a client-side only
+metadata service, it eliminates any potential bottlenecks related to updating
+metadata related to the object storage. The burden of tracking striping and
+other metadata that has traditionally been part of the metadata associated with
+the file system will have to be maintained by the object storage service. The
+lack of a centralized, serialized bottleneck to store that information improves
+concurrency.
 
@@
 support fault detection and recovery including removing and adding processes.
-Second, because a simple count of participating processes is sent with the
-beginning of the transaction, each process is strictly limited to a single
-output operation.  This precludes a process writing multiple array shards, for
-example. To put this in concrete terms, it means that a file could only contain
-a single, globally distributed variable (at the root level of HDF5) OR an
-attribute per process. If this count actually represents a begin/end
-transaction pair instead, it suffers from a lack of accountability for missing
-object (shard) writes. All it would indicate is that a process started and
-ended a transaction connection. There is no validation that a process has done
-everything required of it.
+Second, since the counts represent a begin/end transaction pair, it suffers
+from a lack of accountability for missing object (shard) writes. All it would
+indicate is that a process started and ended a transaction connection. There is
+no validation that a process has done everything required of it.
 
@@
 ensures there is global knowledge that the sub-transaction is expected.  That
-way if the coordinator (transaction leader) fails, which ever process takes
-over that role knows to expect a completion message for that sub-transaction or
-the overall transaction cannot complete. Global sub-transactions can be defined
-at any time since they are a global, synchronized operation broadcasting their
+way if the coordinator (transaction leader) fails, whichever process takes over
+that role knows to expect a completion message for that sub-transaction or the
+overall transaction cannot complete. Global sub-transactions can be defined at
+any time since they are a global, synchronized operation broadcasting their
 existence. While this additional layer does introduce messaging, the overhead
@@
 explicit persist call, how and when synchronizing the metadata across the
-layers and potentially across machines occurs is undefined. By delaying the
+layers and potentially across machines occurs is undefined. Delaying
 synchronization until an explicit persist is called will reduce the update
-frequency, it does delay the visibility of data. Ideally, the metadata object
-would need to be automatically persisted every time a container transaction is
+frequency, but delays the data visibility. Ideally, the metadata object would
+need to be automatically persisted every time a container transaction is
 persisted to the DAOS layer.
@@
 \section{Acknowledgements}
+The authors would like to thank Gary Grider, Eric Barton, Quincey Koziol, and
+John Bent for their early review comments and discussions that clarified the
+details of the design, the intent of the design, and the future plans.
+
 \includegraphics[scale=0.07]{logos/doe_logo}

### 2014-03-21 paper.tex
incorporated John Bent's comments

@@
 an excellent opportunity to evaluate many of the decisions made to feed into
-the phase 2 effort. The initial effort to define a next generation file system
-has made admirable contributions in architecture and design. Formalizing the
-general idea of data staging as burst buffers for the file system will help
-manage the performance variability and offer the data processing opportunities
-outside the main compute and file system. Adding a transactional mechanism to
-manage faults and data visibility helps enable effective analytics without
-having to work around the IO stack semantics.
+the phase 2 effort. The initial effort to define a next generation storage
+system has made admirable contributions in architecture and design. Formalizing
+the general idea of data staging into burst buffers for the storage system will
+help manage the performance variability and offer additional data processing
+opportunities outside the main compute and storage system. Adding a
+transactional mechanism to manage faults and data visibility helps enable
+effective analytics without having to work around the IO stack semantics.
 
-While these and their other contributions are valuable, similar efforts made
+While these and other contributions are valuable, similar efforts made
 elsewhere may offer attractive alternatives or differing semantics that could
@@
 alternative approach for incorporating transactional semantics into the data
-path. The PreDatA project examined how to get the best throughput for data
-operators can offer some insights for refining the Burst Buffer concept.
+path. Another project, PreDatA, examined how to get the best throughput for
+data operators and may offer additional insights into further refinements of
+the Burst Buffer concept.
 
@@
 identified challenges. The first phase recently completed with a second phase
-getting underway. The core focus of the first phase was basic functionality and
-design. Overriding many of the decisions during this and any subsequent phases
-is the reality of budgets. Placing GBs of NVRAM on every node, while a
-potentially advantageous approach, is not financially feasible. With this in
-mind, the second phase will refine this design incorporating fault recovery and
-other features missing from the first phase.
+currently getting underway as of early 2014. The core focus of the first phase
+was basic functionality and design. Overriding many of the decisions during
+this and any subsequent phases is the reality of budgets. Placing GBs of NVRAM
+on every node, while a potentially advantageous approach, is not financially
+feasible. With this in mind, the second phase will refine this design
+incorporating fault recovery and other features missing from the first phase.
 
-The overall design seeks to offer a byte-granular, multi-version concurrency
+The overall design seeks to offer byte-granular, multi-version concurrency
 control. Through the use of a copy-on-write style mechanism, multiple versions
-of a file can be stored in potentially greatly reduced space. It assumes the
+of an object can be stored in potentially greatly reduced space. It assumes the
 client interface will be through an IO library affording a more complicated
-interface that offers richer functionality without requiring user code changes.
-Managing most data access in a machine-local layer rather than requiring
+interface that offers richer functionality requiring only minimal end-user code
+changes.  Managing most data access in a platform-local layer rather than
+requiring
 writing to centralized storage will better support the performance and energy
@@
 The basic architecture incorporates five layers. The top layer is a high level
-IO library, such as the demonstration HDF-5 library. The intent is to only have
+IO library, such as the demonstration HDF-5 library~\cite{hdf5}. The intent is to only have
 access to the storage stack through such an API to manage the complexity of
@@
 concepts were proposed to be incorporated as part of the IO
-stack~\cite{bent:2012:challenges,bent:2012:burst-buffer}.  The current
-FastForward IOD design recommends incorporating SSDs, but specifically lists
-these devices as optional. Unfortunately, both incorporating burst buffers and
-the use of SSDs in the IOD layer may be problematic.  First, the IOD and DAOS
-designs currently are written assuming burst buffers. This means that the
-bulk of IO operations will only hit the IOD layer and proposed functionality,
-such as the function operators, do not discuss the impact on the DAOS layer or
-the operators themselves should a burst buffer not be available.  Consider the
-important functionality of data rearrangement and doing things like changing
-the fast array dimension on shared, spinning media.  The design of DAOS assumes
-that it will only be involved when persisting a completed transaction and only
-for a fraction of the total transactions created.
+stack~\cite{nowoczynski:2008:zest,bent:2012:challenges,bent:2012:burst-buffer}.
+The current FastForward IOD design recommends incorporating SSDs, but
+specifically lists these devices as optional. Unfortunately, not incorporating
+burst buffers and the use of SSDs in the IOD layer may be problematic.  First,
+the IOD and DAOS designs currently are written assuming burst buffers. This
+means that the bulk of IO operations will only hit the IOD layer and proposed
+functionality, such as the function shipping, do not discuss the impact on the
+DAOS layer or the operators themselves should a burst buffer not be available.
+Consider the important functionality of data rearrangement and doing things
+like changing the fast array dimension on shared, spinning media.  The design
+of DAOS assumes that it will only be involved when persisting a completed
+transaction and only for a fraction of the total transactions created.
+Transaction frequency at the IOD layer can be higher since IOD does not
+``flatten'' transactions. Flattening is the process of making each transaction
+fully-independent by duplicating any data unchanged from previous transactions.
+DAOS epochs are persisted, flattened transactions from the IOD layer.
 
@@
 In the case of the written IOD design, it describes a fixed-sized staging area
-shared across all users simultaneously.  This is unlikely to be useful because
-of the limited compute and communication capacity to spare to perform these
-operations at a bottleneck in the IO path.  The use of a separate data staging
-area intentionally separate from the IO path allows using operators on limited
-resources leaving the IO path clear for strictly data movement. A nuance of
-this design is discussed in the Design Philosophy below.
+that is partitioned on a per-application basis.  This is unlikely to be useful
+because of the limited compute and communication capacity to spare to perform
+these operations at a bottleneck in the IO path.  The use of a separate data
+staging area intentionally separate from the IO path allows using operators on
+limited resources leaving the IO path clear for strictly data movement. A
+nuance of this design is discussed in the Design Philosophy below.
 
-By concentrating this functionality into the storage stack, three problems
-arise.  First, the amount of network bandwidth, IO bandwidth, and compute power
-consumed for example operations from a single application is likely to
-completely monopolize the IOD processes. Second, if space and time partitioning
-is used instead, the functionality risks being too small to be useful. Third,
-the hardware performance advantage for SSDs is questionable. Current NAND-based
-flash devices top out at around 400 MB/sec. The key spec that is missing from
-this number is that 400 MB/sec is a measure of the fixed number of available
-IOPS multiplied by the block size. This represents the ideal streaming
-performance possible. The problem is that it costs an IOP to read 1 byte or 1
-block (4KB or 8KB, depending on the device). It costs 1 IOP to write a full
-block--usually. In some cases, it will cost 2 IOPs.  This accounts for the
-required pre-erase write prior to writing to a reused block. In the worst case,
-it can be 3 IOPS per write. This would be for a partial block write (read the
-old block, erase, write the modified block). One-third of 400 MB/sec, about 133
-MB/sec, is well below the streaming performance of HDDs.  Granted, there is
-still rotational and seek latency to deal with for HDDs, but the advantage for
-SSDs has evaporated and potentially turned into a penalty at a considerable
-cost premium.  There are faster SSD solutions on the market that incorporate
-DRAM for caching and using the PCIe bus, for example, but their price precludes
-them from use in an extreme scale platform.
+By concentrating the Burst Buffers and function shipping into the storage
+stack, three problems arise.  First, the amount of network bandwidth, IO
+bandwidth, and compute power consumed for example operations from a single
+application is likely to completely monopolize the IOD processes. Second, if
+space and time partitioning is used instead, the functionality risks being too
+small to be useful. Third, the hardware performance advantage for SSDs is
+questionable. Current NAND-based flash devices top out at around 400 MB/sec.
+The key spec that is missing from this number is that 400 MB/sec is a measure
+of the fixed number of available IOPS multiplied by the block size. This
+represents the ideal streaming performance possible. The problem is that it
+costs an IOP to read 1 byte or 1 block (4KB or 8KB, depending on the device).
+It costs 1 IOP to write a full block--usually. In some cases, it will cost 2
+IOPs.  This accounts for the required pre-erase write prior to writing to a
+reused block. In the worst case, it can be 3 IOPS per write. This would be for
+a partial block write (read the old block, erase, write the modified block).
+One-third of 400 MB/sec, about 133 MB/sec, is well below the streaming
+performance of HDDs.  Granted, there is still rotational and seek latency to
+deal with for HDDs, but the advantage for SSDs has evaporated and potentially
+turned into a penalty at a considerable cost premium.  There are faster SSD
+solutions on the market that incorporate DRAM for caching and using the PCIe
+bus, for example, but their price precludes them from use in an extreme scale
+platform.
 
@@
 The burst buffers design, as presented in the IOD documents, limits the
-placement of the function operators and SSDs for buffers to the IO nodes. The
-team does acknowledge the limitations of this design and intend to ultimately
-focus on spreading the IOD layer from the IO nodes into the compute area as
-well.  This is intended both to help address the limitations of the IO
-bandwidth and compute capability of these few nodes for data processing, but
-also to take advantage of new layers in the storage hierarchy. By incorporating
-NVRAM into compute nodes, new options for buffering data prior to being moved
-to centralized storage become available and addresses some of the concerns
-about SSD performance. This lessens the impact of some operators while offering
+placement of the function operators and SSD buffers to the IO nodes. The team
+does acknowledge the limitations of this design and intend to ultimately focus
+on spreading the IOD layer from the IO nodes into the compute area as well.
+This is intended both to help address the limitations of the IO bandwidth and
+compute capability of these few nodes for data processing, but also to take
+advantage of new layers in the storage hierarchy. By incorporating NVRAM into
+compute nodes, new options for buffering data prior to being moved to
+centralized storage become available and addresses some of the concerns about
+SSD performance. This lessens the impact of some operators while offering
 additional options for places to store data.
@@
 Burst buffers being optional is a high level goal, but not one considered at a
-detailed level with the design. For example, if there is no burst buffer, all
+detailed level within the design. For example, if there is no burst buffer, all
 of the advanced functionality proposed for the IOD layer would have to work
@@
 
-Another concern that is acknowledged, but no thought has been applied to is the
-requirement that a single IOD process be the master for any operation. Should
-the number of concurrent applications exceed the available nodes, sharing an
-IOD process will be required. Since partitioning of the IOD processes for
-exclusive use by particular applications is the assumed operating mode, should
-insufficient IOD resources be available, either a job could be delayed or IOD
-resources could be reallocated from a different process could be redeployed for
-use by the new job. These sorts of considerations still need to be made for a
-full production system.
+Another concern that is acknowledged, but no thought has been applied to, is
+the requirement that a single IOD process be the master for any operation.
+Should the number of concurrent applications exceed the available nodes,
+sharing an IOD process will be required. Since partitioning of the IOD
+processes for exclusive use by particular applications is the assumed operating
+mode, should insufficient IOD resources be available, either a job could be
+delayed or IOD resources could be reallocated from a different process could be
+redeployed for use by the new job. Handling resilience concerns for the IOD
+processes must also be address. These sorts of considerations still need to be
+made for a full production system.
 
@@
 The transaction mechanism manifests in two forms. At the IOD layer, they are
-called transactions and are used to judge whether or not an output is complete
-or not and control access through treating the transaction ID as a version
-identifier. At the DAOS layer, they are called epochs and represent persisted
-(durable) transactions from the IOD layer. Each of these offers different
-functionality, but are connected as is explained below. How these differ from
-the \DDT approach is also explored. While IOD's and \DDTns's transactions are
-seemingly very different, they use a similar high-level design, but very
-different implementation, to solve the same problem.
+called transactions and are used to judge whether or not a set of distributed,
+asynchronous modifications across a set of related objects is complete or not
+and control access through treating the transaction ID of committed transaction
+as a version identifier.  At the DAOS layer, they are called epochs and
+represent persisted (durable) transactions from the IOD layer. Each of these
+offers different functionality, but are connected as is explained below. How
+these differ from the \DDT approach is also explored. While IOD's and \DDTns's
+transactions are seemingly very different, they use a similar high-level
+design, but very different implementation, to solve the same problem.
 
@@
 system. The objects in each container represent different data within a file.
-The three initially defined object types are key-value store, array, and blob.
-The easiest way to understand these types is to evaluate these from the
-perspective of an HDF-5 file, the initial user interface layer. The key-value
-store represents a collection of attributes or groups. The array represents
-part of a potentially multi-dimensional array. It is referred to as
-a shard because it is likely a small piece of a globally defined array. The
-blob represents a byte stream of arbitrary contents.  The fundamental
-difference between an array and a blob is that the array has metadata
-identifying its portion within the global, logical space while the blob is
-simply a 1-dimensional array of bytes that is not shared across IOD nodes.
-Should an operation be deployed into IOD to manipulate an array within a
-container, it would operate on the array rather than on blobs. Given this
+The three initially defined object types are key-value stores,
+multi-dimensional arrays, and blobs.  The easiest way to understand these types
+is to evaluate these from the perspective of an HDF-5 file, the initial user
+interface layer. The key-value store represents a collection of attributes or
+groups. The array represents part of a potentially multi-dimensional array. It
+is referred to as a shard because it is likely a small piece of a globally
+defined array. The blob represents a byte stream of arbitrary contents.  The
+fundamental difference between an array and a blob is that the array has
+metadata identifying its portion within the global, logical space while the
+blob is simply a 1-dimensional array of bytes that is not shared across IOD
+nodes.  Should an operation be deployed into IOD to manipulate an array within
+a container, it would operate on the array rather than on blobs. Given this
 context, the transactions come in two forms.
@@
 the used as the epoch ID. The key difference is that at the DAOS layer, some
-transaction (epoch) IDs will not be represented.
+transaction (epoch) IDs will not be represented since not all IOD transactions
+are necessarily persisted.
 
@@
 get out of sync, three particular risks are introduced.  First, a client could
-create an entry in the metadata service that does not correspond to any objects
-in the object store. Second, a client could create objects that have no
-associated metadata entries. Third, updates to the metadata or object store
-service should be an atomic operation, but due to a lack of coordination, a
-window where the system is inconsistent appears.
+create a dangling entry in the metadata service that does not correspond to any
+objects in the object store. Second, a client could create orphaned objects
+that have no associated metadata entries. Third, updates to the metadata or
+object store service should be an atomic operation, but due to a lack of
+coordination, a window where the system is inconsistent appears.
 
@@
 synchronization until an explicit persist is called will reduce the update
-frequency, but delays the data visibility. Ideally, the metadata object would
-need to be automatically persisted every time a container transaction is
-persisted to the DAOS layer.
+frequency, but delays the data visibility on other platforms. Ideally, the
+metadata object would need to be automatically persisted every time a container
+transaction is persisted to the DAOS layer.
 
@@
 container is a collection of objects. For the DAOS layer, a container is a
-collection of shards. For the IOD an object may be a shard of a global array.
-For DAOS, a shard can host a set of DAOS objects. Having the same names with
-locally correct, globally conflicting definitions serves to confuse how the
-system should work.
+collection of objects across a set of shards. For the IOD an object may be a
+shard of a global array.  For DAOS, a shard can host a set of DAOS objects.
+Having the same names with locally correct, globally conflicting definitions
+serves to confuse how the system should work.
 

### 2014-03-21 paper.tex
incorporated comments from call today

@@
 changes.  Managing most data access in a platform-local layer rather than
-requiring
-writing to centralized storage will better support the performance and energy
-requirements of extreme scale integrated application workflows. At a more
-detailed view, the various layers of the IO stack each contribute different
-functionality and performance implications.
+requiring writing to centralized storage will better support the performance
+and energy requirements of extreme scale integrated application workflows. At a
+more detailed view, the various layers of the IO stack each contribute
+different functionality and performance implications.
 
 The basic architecture incorporates five layers. The top layer is a high level
-IO library, such as the demonstration HDF-5 library~\cite{hdf5}. The intent is to only have
-access to the storage stack through such an API to manage the complexity of
-working with the lower layers. Below the user API is an IO forwarding layer
-that redirects IO calls from the compute nodes to the IO dispatching layer.
-This IO forwarding layer is analogous to the function of the IO nodes in a
-BlueGene machine. The next two layers have considerable functionality. The IO
-dispatcher (IOD) serves as the primary storage interface for the IO stack and
-is the only way to reach the persistent storage array in lower layers. The
-Data Access Object Storage (DAOS) layer serves as the persistent storage
-interface. At the bottom is the Versioning Object Storage Device (VOSD). It
-serves as the interface for storing objects of all types efficiently.
+IO library, such as the demonstration HDF-5 library~\cite{hdf5}. The intent is
+to only have access to the storage stack through such an API to manage the
+complexity of working with the lower layers. Below the user API is an IO
+forwarding layer that redirects IO calls from the compute nodes to the IO
+dispatching layer.  This IO forwarding layer is analogous to the function of
+the IO nodes in a BlueGene machine. The next two layers have considerable
+functionality. The IO dispatcher (IOD) serves as the primary storage interface
+for the IO stack and is the only way to reach the persistent storage array in
+lower layers. The Data Access Object Storage (DAOS) layer serves as the
+persistent storage interface and is intended to be the foundation on which
+everything else is built with no dependence on any technologies specified above
+it. For example, the IOD layer and burst buffers are not required for DAOS to
+operate properly. At the bottom is the Versioning Object Storage Device (VOSD).
+It serves as the interface for storing objects of all types efficiently.
 
@@
 but is the intent of ultimate product. These insights were gained based on
-personal conversations with the team members discussing some of the potential
-challenges with the design as written. These ideas are presented to give a
-fuller picture of where the project is going rather than dwelling on the
+personal conversations with the team
+members~\cite{JohnBent,QuinceyKoziol,EricBarton} discussing some of the
+potential challenges with the design as written. These ideas are presented to
+give a fuller picture of where the project is going rather than dwelling on the
 limitations of the published documents.
@@
 Transaction frequency at the IOD layer can be higher since IOD does not
-``flatten'' transactions. Flattening is the process of making each transaction
-fully-independent by duplicating any data unchanged from previous transactions.
-DAOS epochs are persisted, flattened transactions from the IOD layer.
+``flatten'' transactions. Flattening is the process of transfering the set of
+copy-on-write changes from transaction/epoch $n$ to transaction $n+m$ into the
+DAOS layer. The DAOS layer mirrors the copy-on-write functionality, but stores
+the differences between epochs rather than individual transactions. The primary
+difference is that at the DAOS layer, data is stored in large, logically
+contiguous chunks to manage the metadata load and hopefully improve read
+performance.
 
@@
 space and time partitioning is used instead, the functionality risks being too
-small to be useful. Third, the hardware performance advantage for SSDs is
-questionable. Current NAND-based flash devices top out at around 400 MB/sec.
-The key spec that is missing from this number is that 400 MB/sec is a measure
-of the fixed number of available IOPS multiplied by the block size. This
-represents the ideal streaming performance possible. The problem is that it
-costs an IOP to read 1 byte or 1 block (4KB or 8KB, depending on the device).
-It costs 1 IOP to write a full block--usually. In some cases, it will cost 2
-IOPs.  This accounts for the required pre-erase write prior to writing to a
-reused block. In the worst case, it can be 3 IOPS per write. This would be for
-a partial block write (read the old block, erase, write the modified block).
-One-third of 400 MB/sec, about 133 MB/sec, is well below the streaming
-performance of HDDs.  Granted, there is still rotational and seek latency to
-deal with for HDDs, but the advantage for SSDs has evaporated and potentially
-turned into a penalty at a considerable cost premium.  There are faster SSD
-solutions on the market that incorporate DRAM for caching and using the PCIe
-bus, for example, but their price precludes them from use in an extreme scale
-platform.
+small to be useful. Third, the long-term hardware performance advantage for
+SSDs is questionable. Recent studies have shown that the erase-before-write
+and interference between reading and writing with flash-based SSDs can cause
+severe performance problems~\cite{skourtis:2013:ssd-performance}. The inclusion
+of an optional use SSD layer in the new Trinity machine at Los Alamos will
+offer a testbed to determine how likely these observed problems would affect a
+production extreme scale platform.
+
+%Current NAND-based flash devices top out at around 400 MB/sec.
+%The key spec that is missing from this number is that 400 MB/sec is a measure
+%of the fixed number of available IOPS multiplied by the block size. This
+%represents the ideal streaming performance possible. The problem is that it
+%costs an IOP to read 1 byte or 1 block (4KB or 8KB, depending on the device).
+%It costs 1 IOP to write a full block--usually. In some cases, it will cost 2
+%IOPs.  This accounts for the required pre-erase write prior to writing to a
+%reused block. In the worst case, it can be 3 IOPS per write. This would be for
+%a partial block write (read the old block, erase, write the modified block).
+%One-third of 400 MB/sec, about 133 MB/sec, is well below the streaming
+%performance of HDDs.  Granted, there is still rotational and seek latency to
+%deal with for HDDs, but the advantage for SSDs has evaporated and potentially
+%turned into a penalty at a considerable cost premium.  There are faster SSD
+%solutions on the market that incorporate DRAM for caching and using the PCIe
+%bus, for example, but their price precludes them from use in an extreme scale
+%platform.
 
@@
 centralized storage become available and addresses some of the concerns about
-SSD performance. This lessens the impact of some operators while offering
-additional options for places to store data.
+SSD performance. For example, including a small amount of Phase Change memory
+into many or most compute nodes offers a way to move data outside of both the
+compute and IO path for data and communication intensive operations. Other 
+projects~\cite{zheng:2010:predata} have suggested this will have value, but the
+cost will have to be considered as part of the overall platform budget. This
+lessens the impact of some operators while offering additional options for
+places to store data.
 
@@
 Should the number of concurrent applications exceed the available nodes,
-sharing an IOD process will be required. Since partitioning of the IOD
-processes for exclusive use by particular applications is the assumed operating
-mode, should insufficient IOD resources be available, either a job could be
-delayed or IOD resources could be reallocated from a different process could be
-redeployed for use by the new job. Handling resilience concerns for the IOD
-processes must also be address. These sorts of considerations still need to be
-made for a full production system.
+sharing an IOD process will be required. The requirements both in terms of
+scheduling and resource management were cut from the project due to funding
+limitations. Since partitioning of the IOD processes for exclusive use by
+particular applications is the assumed operating mode, should insufficient IOD
+resources be available, either a job could be delayed or IOD resources could be
+reallocated from a different process could be redeployed for use by the new
+job. Handling resilience concerns for the IOD processes must also be address.
+These sorts of considerations still need to be made for a full production
+system.
 
@@
 called transactions and are used to judge whether or not a set of distributed,
-asynchronous modifications across a set of related objects is complete or not
-and control access through treating the transaction ID of committed transaction
-as a version identifier.  At the DAOS layer, they are called epochs and
-represent persisted (durable) transactions from the IOD layer. Each of these
-offers different functionality, but are connected as is explained below. How
-these differ from the \DDT approach is also explored. While IOD's and \DDTns's
-transactions are seemingly very different, they use a similar high-level
-design, but very different implementation, to solve the same problem.
+asynchronous modifications across a set of related objects is complete or not.
+It is also used to control access by treating the transaction ID of committed
+transaction as a version identifier.  At the DAOS layer, they are called epochs
+and represent persisted (durable) transactions from the IOD layer. Each of
+these offers different functionality, but are connected as is explained below.
+How these differ from the \DDT approach is also explored.  While IOD's and
+\DDTns's transactions are seemingly very different, they use a similar
+high-level design, but very different implementation, to solve the same
+problem.
 
@@
 concepts must be explained first. At the coarsest grain level is a container.
-Each container serves to host a transaction and contains a collection of
-objects. Conceptually, containers corresponds to a file in a traditional file
-system. The objects in each container represent different data within a file.
-The three initially defined object types are key-value stores,
+Each container provides the single access context through which to access a
+collection of objects. Transactions are the way that a series of modifications
+to the objects within a container are treated atomically. Conceptually,
+containers corresponds to a something akin to an HDF5 file in a traditional
+file system. The objects in each container represent different data within a
+file.  The three initially defined object types are key-value stores,
 multi-dimensional arrays, and blobs.  The easiest way to understand these types
@@
 interface layer. The key-value store represents a collection of attributes or
-groups. The array represents part of a potentially multi-dimensional array. It
-is referred to as a shard because it is likely a small piece of a globally
-defined array. The blob represents a byte stream of arbitrary contents.  The
-fundamental difference between an array and a blob is that the array has
-metadata identifying its portion within the global, logical space while the
-blob is simply a 1-dimensional array of bytes that is not shared across IOD
-nodes.  Should an operation be deployed into IOD to manipulate an array within
-a container, it would operate on the array rather than on blobs. Given this
-context, the transactions come in two forms.
+groups. The array represents a potentially multi-dimensional array.  The blob
+represents a byte stream of arbitrary contents.  The fundamental difference
+between an array and a blob is that the array has metadata specifying the
+dimension(s). At the physical layer within the IO nodes, all of these objects
+may be striped across multiple IO nodes.  Given this context, the transactions
+come in two forms.
 
@@
 
-The team uses the term ``Primary'' to define the process(es) that contact the
-IOD layer directly to open a container. Unfortunately, this term is never used
-nor defined within any of the published documents. There are some very positive
-and negative performance implications of this approach. On the positive side,
-like PVFS, it is possible for a single process to open a container and share
-that handle with other processes reducing the load on the IOD layer. The key
-idea is that the single leader vs. multi-leader transaction mode is
-automatically defined based on the number of Primary processes that start a
-transaction on a container. This greatly simplifies any additional client-side
-load in using the different styles of transactions. Also on the positive side,
-it offers the ability to have a subset of processes that are considered
-transaction participants affording reduced load on the IOD processes and more
-localized transaction management on the client side. On the negative side, for
-a fully IOD managed transaction, ALL processes must start the transaction on
-the container forcing a potentially immense number of simultaneous calls to the
-IOD layer to open a container. This sort of onslaught is what is dealt with on
-metadata servers today with mass file opens. The reality is that this ends up
-having the same sort of performance implications as distributed denial of
-service attacks on the IOD processes.
+There are two steps for beginning a transaction on a container. The first step
+is for one or more process to open the container. This handle can be shared
+eliminating the need for every participating process to hit the IOD layer to
+open the file. The second step is a call to determine how many leaders will
+participate in the transaction. In the single leader case, there is no
+aggregation of success/fail statuses to determine the final transaction state.
+Instead, it is assumed that the client will fully manage the transaction. In
+the multi-leader model, some subset from 2 to $n$ where $n$ is the count of all
+processes, declare themselves a leader for this container operation to the IOD
+layer. Any number of processes can participate in modifying container without
+regard to whether or not they are a leader. Once each leader has finished, with
+the assumption that any clients they may be responsible for are finished as
+well, the IOD layer aggregates those responses to either commit or abort the
+transaction.
 
-Ultimately, with the passive detection of faults for Primaries, the transaction
-mechanism can work very well. A mostly unstated restriction that is being
-relaxed is that every sequential transaction on a container is considered
+Ultimately, with the passive detection of faults for transaction leaders, the
+transaction mechanism can work very well. A mostly unstated restriction that is
+being relaxed is that every sequential transaction on a container is considered
 dependent on the earlier transaction. Should one output be delayed and the
@@
 care if one fails, cannot work in the current form. This has been acknowledged
-and is planned on being relaxed in phase two.
+and is being relaxed requiring a new parameter to the creation of a transaction
+determining if it will be dependent or not.
 
@@
 \subsection{Metadata Management}
-Eliminating metadata management as a special case and instead treating it just
-as data is a central design goal of the Fast Forward project. This is a hybrid
-approach to metadata management that is half-way between providing no inherent
-metadata support and having a fully integrated, but separate metadata
-manaegement system.
+Metadata management has been a perennial challenge for parallel storage
+systems.  Eliminating metadata management as a special case and instead
+treating it just as data is a central design goal of the Fast Forward project.
+This is a hybrid approach to metadata management that is half-way between
+providing no inherent metadata support and having a fully integrated, but
+separate metadata manaegement system.
 
@@
 IOD and DAOS both share a philosophy that they will have to maintain the
-metadata about how objects are striped and where they are placed. It was
-decided that the additional data to track the namespace as well was a small
-enough additional increment that it was worth maintaining within the system.
+metadata about how the physical pieces of the logical objects are striped and
+where they are placed. The primary metadata management is done at the DAOS
+layer with the IOD layer relying on the DAOS layer for all authoritative
+information about containers and objects. The only place where the IOD layer
+manages metadata for itself is to manage how the different objects are striped
+across the IO nodes.
 
@@
 container is treated like any other data in the system and striped as
-appropriate at both the IOD and the DAOS layers. Unfortunately, this still
-couples the metadata to a single object that must serialize access. If the
-metadata, including information about striping and other data layout operations
-were separated completely from the data path, more scalable throughput could be
-achieved. The real challenge of this is introduced by the IOD, DAOS, and VOSD
-layers. Each of these requires some different metadata storage and the
-migration is transparent to the user. Supporting fully independent metadata
-with this model is difficult. Serious thought on how to do this effectively
-outside the data path should be considered for phase two.
+appropriate. Unfortunately, this still couples the metadata to a single object
+that must serialize access. If the metadata, including information about
+striping and other data layout operations were separated completely from the
+data path, more scalable throughput could be achieved. The real challenge of
+this is introduced by the IOD, DAOS, and VOSD layers collectively. Each of
+these requires some different metadata storage and the migration is transparent
+to the user.  Supporting fully independent metadata with this model is
+difficult. Serious thought on how to do this effectively outside the data path
+should be considered for phase two.
 
@@
 with \DDT is illustrated in Figure~\ref{fig:performance}. This performance is
-explored in detail in a previous paper~\cite{lofstead:2013:pdsw-txn}. The
-breakdown of the number of participants in each role is shown in
-Table~\ref{tab:scaling}. For comparison, consider the Number of
-Sub-Coordinators equivalent to IOD processes. The Processes Per Sub-Coordinator
-represents the number of clients that use a particular IOD process. For these
-tests, we maintained a balanced distribution and always used at least two
-sub-coordinators to slow down the processing.
+explored in detail in a previous paper~\cite{lofstead:2013:pdsw-txn}.
+%The
+%breakdown of the number of participants in each role is shown in
+%Table~\ref{tab:scaling}. For comparison, consider the Number of
+%Sub-Coordinators equivalent to IOD processes. The Processes Per Sub-Coordinator
+%represents the number of clients that use a particular IOD process. For these
+%tests, we maintained a balanced distribution and always used at least two
+%sub-coordinators to slow down the processing.
 
@@
 
-\begin{table}[ht]
-    \vspace{-0.15in}
-    \centering
-    \caption[Scaling Configuration]{Performance Tests Scaling Configuration}
-    \bigskip
-    \vspace{-0.15in}
-
-    \begin{tabular}{|r|r|r|}
-\hline
-Processes & \vtop{\hbox{\strut Number of}\hbox{\strut Sub-Coordinators}} & \vtop{\hbox{\strut Processes Per} \hbox{\strut Sub-Coordinator}}\\
-\hline
-%8 & 2 & 4 \\
-%16 & 2 & 8 \\
-%32 & 2 & 16 \\
-%64 & 2 & 32 \\
-%128 & 2 & 64 \\
-256 & 2 & 128 \\
-512 & 2 & 256 \\
-1024 & 4 & 256 \\
-2048 & 8 & 256 \\
-4096 & 16 & 256 \\
-8192 & 32 & 256 \\
-16384 & 64 & 256 \\
-32768 & 128 & 256 \\
-65536 & 256 & 256 \\
-\hline
-    \end{tabular}
-    \label{tab:scaling}
-\end{table}
+%\begin{table}[ht]
+%    \vspace{-0.15in}
+%    \centering
+%    \caption[Scaling Configuration]{Performance Tests Scaling Configuration}
+%    \bigskip
+%    \vspace{-0.15in}
+%
+%    \begin{tabular}{|r|r|r|}
+%\hline
+%Processes & \vtop{\hbox{\strut Number of}\hbox{\strut Sub-Coordinators}} & \vtop{\hbox{\strut Processes Per} \hbox{\strut Sub-Coordinator}}\\
+%\hline
+%%8 & 2 & 4 \\
+%%16 & 2 & 8 \\
+%%32 & 2 & 16 \\
+%%64 & 2 & 32 \\
+%%128 & 2 & 64 \\
+%256 & 2 & 128 \\
+%512 & 2 & 256 \\
+%1024 & 4 & 256 \\
+%2048 & 8 & 256 \\
+%4096 & 16 & 256 \\
+%8192 & 32 & 256 \\
+%16384 & 64 & 256 \\
+%32768 & 128 & 256 \\
+%65536 & 256 & 256 \\
+%\hline
+%    \end{tabular}
+%    \label{tab:scaling}
+%\end{table}
 
@@
 
-In the multi-leader model, using a count of client connections to determine if
-transaction is complete is problematic. First, the aggregation of completion
-messages are based on the local aggregation point reporting to the transaction
-leader.  There is no ability to detect or recover from a failure of this IOD
-process that would report to the transaction leader. The current assumption
-that the IOD layer will be an MPI program is not workable until MPI can fully
-support fault detection and recovery including removing and adding processes.
-Second, since the counts represent a begin/end transaction pair, it suffers
-from a lack of accountability for missing object (shard) writes. All it would
-indicate is that a process started and ended a transaction connection. There is
-no validation that a process has done everything required of it.
+The multi-leader model introduces the possibility of forcing a rollback of an
+entire transaction when a partial retry might be sufficient for success. Since
+the transactions are managed at a high level rather than the individual
+tasks, a failure in a limited distributed task can cause the entire transaction
+to fail. For example, consider 10 processes each have 5 tasks, but 3 of those
+10 have an additional shared task to complete. If the task shared by the 3
+processes fails on any of the three, the entire transaction would roll back
+because it is a coarse-grained success/failure. If a concept like
+sub-transactions at a task granularity were used, then it would be possible for
+the one process that failed to report just that failure. Then the transaction
+manager could reassign the resources for these 3 processes and try just that
+operation again. If it now succeeds, then the overall transaction can be marked
+successful only redoing the minimum amount of work required.
 
@@
 \centering
-\includegraphics[keepaspectratio=true, width=0.9\columnwidth]{images/optimized-protocol}
+\includegraphics[keepaspectratio=true, width=0.5\columnwidth]{images/optimized-protocol}
 \vspace{-0.15in}
@@
 \section{Acknowledgements}
-The authors would like to thank Gary Grider, Eric Barton, Quincey Koziol, and
-John Bent for their early review comments and discussions that clarified the
-details of the design, the intent of the design, and the future plans.
+The authors would like to thank Eric Barton, John Bent, Gary Grider, and
+Quincey Koziol for their early review comments and discussions that clarified
+the details of the design, the intent of the design, and the future plans.
 

### 2014-03-22 paper.tex
submission version

@@
 
-\addtolength{\parskip}{-0.04in}
+\addtolength{\parskip}{-0.06in}
 
@@
 \begin{abstract}
-With phase 1 of the Fast Forward Storage and IO Stack project complete, it is
-an excellent opportunity to evaluate many of the decisions made to feed into
-the phase 2 effort. The initial effort to define a next generation storage
-system has made admirable contributions in architecture and design. Formalizing
-the general idea of data staging into burst buffers for the storage system will
-help manage the performance variability and offer additional data processing
-opportunities outside the main compute and storage system. Adding a
-transactional mechanism to manage faults and data visibility helps enable
-effective analytics without having to work around the IO stack semantics.
+The DOE Extreme-Scale Technology Acceleration Fast Forward Storage and IO Stack
+project is going to have significant impact on storage systems design within
+and beyond the HPC community. With phase 1 of the project complete, it is an
+excellent opportunity to evaluate many of the decisions made to feed into the
+phase 2 effort. With this paper we not only provide a timely summary of
+important aspects of the design specifications but also capture the underlying
+reasoning that is not available elsewhere.
 
-While these and other contributions are valuable, similar efforts made
-elsewhere may offer attractive alternatives or differing semantics that could
-yield a more feature rich environment with little to no additional overhead.
-For example, the Doubly Distributed Transactions (\DDTns) protocol offers an
-alternative approach for incorporating transactional semantics into the data
-path. Another project, PreDatA, examined how to get the best throughput for
-data operators and may offer additional insights into further refinements of
-the Burst Buffer concept.
+The initial effort to define a next generation storage system has made
+admirable contributions in architecture and design.  Formalizing the general
+idea of data staging into burst buffers for the storage system will help manage
+the performance variability and offer additional data processing opportunities
+outside the main compute and storage system. Adding a transactional mechanism
+to manage faults and data visibility helps enable effective analytics without
+having to work around the IO stack semantics. While these and other
+contributions are valuable, similar efforts made elsewhere may offer attractive
+alternatives or differing semantics that could yield a more feature rich
+environment with little to no additional overhead. For example, the Doubly
+Distributed Transactions (\DDTns) protocol offers an alternative approach for
+incorporating transactional semantics into the data path. Another project,
+PreDatA, examined how to get the best throughput for data operators and may
+offer additional insights into further refinements of the Burst Buffer concept.
 
-This paper examines some of the choices made by the FastForward team and
+This paper examines some of the choices made by the Fast Forward team and
 compares them with other options and offers observations and suggestions based
@@
 
-The overall design seeks to offer byte-granular, multi-version concurrency
-control. Through the use of a copy-on-write style mechanism, multiple versions
-of an object can be stored in potentially greatly reduced space. It assumes the
-client interface will be through an IO library affording a more complicated
-interface that offers richer functionality requiring only minimal end-user code
-changes.  Managing most data access in a platform-local layer rather than
-requiring writing to centralized storage will better support the performance
-and energy requirements of extreme scale integrated application workflows. At a
-more detailed view, the various layers of the IO stack each contribute
-different functionality and performance implications.
+The overall design seeks to offer high availability, byte-granular,
+multi-version concurrency control. Through the use of a copy-on-write style
+mechanism, multiple versions of an object can be stored in potentially greatly
+reduced space. It assumes the client interface will be through an IO library
+affording a more complicated interface that offers richer functionality
+requiring only minimal end-user code changes.  Managing most data access in a
+platform-local layer rather than requiring writing to centralized storage will
+better support the performance and energy requirements of extreme scale
+integrated application workflows. At a more detailed view, the various layers
+of the IO stack each contribute different functionality and performance
+implications.
 
@@
 for the IO stack and is the only way to reach the persistent storage array in
-lower layers. The Data Access Object Storage (DAOS) layer serves as the
-persistent storage interface and is intended to be the foundation on which
+lower layers. The Distributed Application Object Storage (DAOS) layer serves as
+the persistent storage interface and is intended to be the foundation on which
 everything else is built with no dependence on any technologies specified above
@@
 stack~\cite{nowoczynski:2008:zest,bent:2012:challenges,bent:2012:burst-buffer}.
-The current FastForward IOD design recommends incorporating SSDs, but
+The current Fast Forward IOD design recommends incorporating SSDs, but
 specifically lists these devices as optional. Unfortunately, not incorporating
 burst buffers and the use of SSDs in the IOD layer may be problematic.  First,
-the IOD and DAOS designs currently are written assuming burst buffers. This
-means that the bulk of IO operations will only hit the IOD layer and proposed
-functionality, such as the function shipping, do not discuss the impact on the
-DAOS layer or the operators themselves should a burst buffer not be available.
-Consider the important functionality of data rearrangement and doing things
-like changing the fast array dimension on shared, spinning media.  The design
-of DAOS assumes that it will only be involved when persisting a completed
-transaction and only for a fraction of the total transactions created.
-Transaction frequency at the IOD layer can be higher since IOD does not
-``flatten'' transactions. Flattening is the process of transfering the set of
-copy-on-write changes from transaction/epoch $n$ to transaction $n+m$ into the
-DAOS layer. The DAOS layer mirrors the copy-on-write functionality, but stores
-the differences between epochs rather than individual transactions. The primary
-difference is that at the DAOS layer, data is stored in large, logically
-contiguous chunks to manage the metadata load and hopefully improve read
-performance.
+the IOD design currently is written assuming burst buffers. This means that the
+bulk of IO operations will only hit the IOD layer and proposed functionality,
+such as the function shipping, do not discuss the impact on the DAOS layer or
+the operators themselves should a burst buffer not be available.  Consider the
+important functionality of data rearrangement and doing things like changing
+the fast array dimension on shared, spinning media.  The design of DAOS assumes
+that it will only be involved when persisting a completed transaction and only
+for a fraction of the total transactions created.  Transaction frequency at the
+IOD layer can be higher since IOD does not ``flatten'' transactions. Flattening
+is the process of transferring the set of copy-on-write changes from
+transaction/epoch $n$ to transaction $n+m$ into the DAOS layer. The DAOS layer
+mirrors the copy-on-write functionality, but stores the differences between
+epochs rather than individual transactions. The primary difference is that at
+the DAOS layer, data is stored in large, logically contiguous chunks to manage
+the metadata load and hopefully improve read performance.
 
@@
 of an optional use SSD layer in the new Trinity machine at Los Alamos will
-offer a testbed to determine how likely these observed problems would affect a
+offer a test bed to determine how likely these observed problems would affect a
 production extreme scale platform.
@@
 Another concern that is acknowledged, but no thought has been applied to, is
-the requirement that a single IOD process be the master for any operation.
-Should the number of concurrent applications exceed the available nodes,
-sharing an IOD process will be required. The requirements both in terms of
-scheduling and resource management were cut from the project due to funding
-limitations. Since partitioning of the IOD processes for exclusive use by
-particular applications is the assumed operating mode, should insufficient IOD
-resources be available, either a job could be delayed or IOD resources could be
-reallocated from a different process could be redeployed for use by the new
-job. Handling resilience concerns for the IOD processes must also be address.
-These sorts of considerations still need to be made for a full production
-system.
+the requirement that a single IOD process of the set assigned to an application
+is the master for any operation.  Should the number of concurrent applications
+exceed the available nodes, sharing an IOD process will be required. The
+requirements both in terms of scheduling and resource management were cut from
+the project due to funding limitations. Since partitioning of the IOD processes
+for exclusive use by particular applications is the assumed operating mode,
+should insufficient IOD resources be available, either a job could be delayed
+or IOD resources could be reallocated from a different process could be
+redeployed for use by the new job. Handling resilience concerns for the IOD
+processes must also be address.  These sorts of considerations still need to be
+made for a full production system.
 
@@
 mechanism that can determine if a process or node is no longer reachable. This
-defines how a fault will be detected and what will trigger a passive fault
-recovery (i.e., transaction abort).
+detection happens at the DAOS layer and when a fault is detected, the rollback
+process is pushed up to the IOD layer for all non-persisted or non-committed
+transactions. This defines how a fault will be detected and what will trigger a
+passive fault recovery (i.e., transaction abort).
 
@@
 providing no inherent metadata support and having a fully integrated, but
-separate metadata manaegement system.
+separate metadata management system.
 

### 2014-04-28 paper.tex
IASDS submission version

@@
 
-\addtolength{\parskip}{-0.06in}
+% change the inter-paragraph spacing
+%\addtolength{\parskip}{-0.06in}
 
@@
 
+The overall target machine architecture is illustrated in
+Figure~\ref{fig:architecture}.
+
+\begin{figure}[ht]
+%\vspace{-0.15in}
+\centering
+\includegraphics[keepaspectratio=true, width=\columnwidth]{images/architecture}
+%\vspace{-0.15in}
+\caption{Target Machine Architecture}
+\label{fig:architecture}
+%\vspace{-0.15in}
+\end{figure}
+
 The basic architecture incorporates five layers. The top layer is a high level
@@
 
-The rest of the paper is organized as follows. Section~\ref{sec:burst}
-discusses some of the features of incorporating burst buffers as designed and
-suggests some considerations and alternatives for the next generation of this
-project. Section~\ref{sec:transactions} discusses the transactions approach
-offered in the IOD layer and the corresponding epochs in the DAOS layer. It
-also offers a comparison to the \DDT system given the very similar high-level
-design and motivating use case.  Section~\ref{sec:summary} discusses the system
-overall with recommendations on what design elements should be considered based
-on broader issues with current HPC data centers. The paper is concluded in
+The rest of the paper is organized as follows. First, an overview of the
+target machine architecture and how it maps to the proposed components is
+presented in Section~\ref{sec:overview}. Section~\ref{sec:burst} provides a
+more detailed overview of the IOD layer and discusses some of the features of
+incorporating burst buffers as designed and suggests some considerations and
+alternatives for the next generation of this project.
+Section~\ref{sec:transactions} discusses the transactions approach offered in
+the IOD layer and the corresponding epochs in the DAOS layer. It also offers a
+comparison to the \DDT system given the very similar high-level design and
+motivating use case.  Section~\ref{sec:summary} discusses the system overall
+with recommendations on what design elements should be considered based on
+broader issues with current HPC data centers. The paper is concluded in
 Section~\ref{sec:conclusion} with a summary of the broad issues covered in the
@@
 
-\section{Burst Buffers}
+\section{Overview}
+\label{sec:overview}
+
+The demonstration system uses some particular existing technology to enable
+testing the proposed concepts. The mapping is presented in
+Figure~\ref{fig:arch-mapping}. We examine this mapping by evaluating the
+descriptive blocks at the bottom of the figure.
+
+\begin{figure}[ht]
+%\vspace{-0.15in}
+\centering
+\includegraphics[keepaspectratio=true, width=\columnwidth]{images/arch-mapping}
+%\vspace{-0.15in}
+\caption{Mapping FFSIO Components to Machine Architecture}
+\label{fig:arch-mapping}
+%\vspace{-0.15in}
+\end{figure}
+
+The dark blue block that includes the HDF5 VOL block represents the User API
+layer. It is worth noting that while this is the intended interaction
+mechanism, it is still possible for other interfaces like MPI-IO and POSIX to
+interact with the system. The IO Forwarding layer is represented by the two
+black boxes. The represented split shows that the IO nodes can offer some
+control over how data is moved from the compute area towards storage. The
+IO dispatcher, the primary focus of the evaluation presented in this paper, is
+shown in the light green box. The one extension is the dark pink block
+representing technology like SSDs on the IO nodes. This is where a burst
+buffer, if the NVRAM is available, would be hosted. The DAOS layer is the
+yellow box. The VOSD is the purple box.
+
+The specification of HDF5 and Lustre at the top and bottom of the stack is 
+an expedience for demonstrating the functionality. In a production version of
+the proposed system, these may be represented by other technologies.
+
+Throughout the paper, multiple terms are used to define an operation applied to
+data outside of the main application computation. These include both function
+shipping and just the term operator. In all cases, the idea is a piece of code
+is placed at some location along the IO path processing data as it passes
+through that point. Note that this works in both directions as well as in
+place. For example, when reading, an operator may be used to only retrieve the
+pieces of a large data set from the DAOS and VOSD layers into the IOD caches
+automatically.  It is also possible that an operator is deployed to either
+rearrange data or generate a new copy with different ordering. For example,
+changing the fast dimension for a multi-dimensional array can greatly improve
+reading performance for particular reading patterns. In all cases, these
+operators offer a chance to apply a relatively light piece of processing to
+data rather than an intensive data analytics operation. These heavy-weight
+operations should be left to the main computation resources leaving capacity
+for other concurrent applications using the IOD layer.
+
+\section{IOD Layer and Burst Buffers}
 \label{sec:burst}
+
+The IOD layer is intended to function as a higher performance, machine local
+caching layer for the DAOS/VOSD layers. Even in a simplified system, by keeping
+the IOD layer as part of the stack, making portable code is simpler since it
+eliminates the need to use a different lower level APIs from the user layer
+(e.g., HDF5) if the system lacks a feature, such as burst buffers. In an ideal
+design, the IOD API could function solely an directly on the DAOS servers. The
+reality is that performance demands and budget constraints will force something
+like burst buffers to compensate for the performance afforded and to offer a
+location to process data down to volumes that can fit through the limited
+bandwidth. Given the criticality of burst buffers and the transaction mechanism
+for end-to-end performance, these ideas are examined in detail below with
+comparison to alternatives that may have been avoided due to perceived
+performance limitations. In particular, the \DDT protocol was designed to
+address these sorts of scenarios and has demonstrated excellent performance.
 
@@
 
-One of the bigger concerns is the observation that the original data staging
-proposals all used compute nodes while the newer proposals seek not only to
-make them a fixed portion of the IO stack, but also shared across all machine
-users. The PreDatA~\cite{zheng:2010:predata} paper in particular examines the
-potential costs and advantages of where to place operators similar to the IOD
-proposed function shipping. There are two key takeaways from PreDatA. First,
-placement matters.  Depending on the communication intensity vs. computation
-intensity, where along the IO path to place the operation can matter
-significantly. Second, and more importantly, the amount of time spent
-processing for the operators was stretched to the point where it consumed
-nearly all of the time between IO operations. The given ratios of compute
-processes to staging process examined is representative for future extreme
-scale platforms. If anything, the ratios offer more staging processes than IOD
-processes would be available.
+One of the bigger concerns is that the original data staging proposals all used
+compute nodes while the newer proposals seek not only to make them a fixed
+portion of the IO stack, but also shared across all machine users. The
+PreDatA~\cite{zheng:2010:predata} paper in particular examines the potential
+costs and advantages of where to place operators similar to the IOD proposed
+function shipping. There are two key takeaways from PreDatA. First, placement
+matters.  Depending on the communication intensity vs. computation intensity,
+where along the IO path to place the operation can matter significantly.
+Second, and more importantly, the amount of time spent processing for the
+operators was stretched to the point where it consumed nearly all of the time
+between IO operations. The given ratios of compute processes to staging process
+examined is representative for future extreme scale platforms. If anything, the
+ratios offer more staging processes than IOD processes would be available.
 
@@
 space and time partitioning is used instead, the functionality risks being too
-small to be useful. Third, the long-term hardware performance advantage for
-SSDs is questionable. Recent studies have shown that the erase-before-write
-and interference between reading and writing with flash-based SSDs can cause
-severe performance problems~\cite{skourtis:2013:ssd-performance}. The inclusion
-of an optional use SSD layer in the new Trinity machine at Los Alamos will
-offer a test bed to determine how likely these observed problems would affect a
-production extreme scale platform.
+small to be useful. For example, if some of the IO nodes are dedicated either
+throughout an application run (space partitioning) or for just phases of the
+application runtime (time partitioning), the allocated capacity or ownership
+duration may be insufficient to be useful. Third, the long-term hardware
+performance advantage for SSDs is questionable. Recent studies have shown that
+the erase-before-write and interference between reading and writing with
+flash-based SSDs can cause severe performance
+problems~\cite{skourtis:2013:ssd-performance}. The inclusion of an optional use
+SSD layer in the new Trinity machine at Los Alamos will offer a test bed to
+determine how likely these observed problems would affect a production extreme
+scale platform.
 
-%Current NAND-based flash devices top out at around 400 MB/sec.
-%The key spec that is missing from this number is that 400 MB/sec is a measure
-%of the fixed number of available IOPS multiplied by the block size. This
-%represents the ideal streaming performance possible. The problem is that it
-%costs an IOP to read 1 byte or 1 block (4KB or 8KB, depending on the device).
-%It costs 1 IOP to write a full block--usually. In some cases, it will cost 2
-%IOPs.  This accounts for the required pre-erase write prior to writing to a
-%reused block. In the worst case, it can be 3 IOPS per write. This would be for
-%a partial block write (read the old block, erase, write the modified block).
-%One-third of 400 MB/sec, about 133 MB/sec, is well below the streaming
-%performance of HDDs.  Granted, there is still rotational and seek latency to
-%deal with for HDDs, but the advantage for SSDs has evaporated and potentially
-%turned into a penalty at a considerable cost premium.  There are faster SSD
-%solutions on the market that incorporate DRAM for caching and using the PCIe
-%bus, for example, but their price precludes them from use in an extreme scale
-%platform.
+Current NAND-based flash devices top out at around 400 MB/sec.  The key spec
+that is missing from this number is that 400 MB/sec is a measure of the fixed
+number of available IOPS multiplied by the block size. This represents the
+ideal streaming performance possible. The problem is that it costs an IOP to
+read 1 byte or 1 block (4KB or 8KB, depending on the device).  It costs 1 IOP
+to write a full block--usually. In some cases, it will cost 2 IOPs.  This
+accounts for the required pre-erase write prior to writing to a reused block.
+In the worst case, it can be 3 or more IOPS per write. For example, if writing
+forces both garbage collection and block compaction, many blocks may have to be
+read, the data combined, the blocks erased, the combined data written, and then
+the intended write operation can finally proceed on the newly freed blocks
+(erase and then write).  One-third of 400 MB/sec, about 133 MB/sec, is well
+below the streaming performance of HDDs.  Granted, there is still rotational
+and seek latency to deal with for HDDs, but the advantage for SSDs has
+evaporated and potentially turned into a considerable penalty at a cost
+premium.  There are faster SSD solutions on the market that incorporate DRAM
+for caching and using the PCIe bus, for example, but their price precludes them
+from use in an extreme scale platform given budget constraints.
 
@@
 
-\section{Transactions and Epochs}
+\section{Transactions, Epochs, and Metadata}
 \label{sec:transactions}
@@
 Each container provides the single access context through which to access a
-collection of objects. Transactions are the way that a series of modifications
-to the objects within a container are treated atomically. Conceptually,
-containers corresponds to a something akin to an HDF5 file in a traditional
-file system. The objects in each container represent different data within a
-file.  The three initially defined object types are key-value stores,
-multi-dimensional arrays, and blobs.  The easiest way to understand these types
-is to evaluate these from the perspective of an HDF-5 file, the initial user
-interface layer. The key-value store represents a collection of attributes or
-groups. The array represents a potentially multi-dimensional array.  The blob
-represents a byte stream of arbitrary contents.  The fundamental difference
-between an array and a blob is that the array has metadata specifying the
-dimension(s). At the physical layer within the IO nodes, all of these objects
-may be striped across multiple IO nodes.  Given this context, the transactions
-come in two forms.
+collection of objects. Transactions are used to treat a series of modifications
+to the objects within a container atomically. Conceptually, containers
+corresponds to a something akin to an HDF5 file in a traditional file system.
+The objects in each container represent different data within a file.  The
+three initially defined object types are key-value stores, multi-dimensional
+arrays, and blobs.  The easiest way to understand these types is to evaluate
+these from the perspective of an HDF-5 file, the initial user interface layer.
+The key-value store represents a collection of attributes or groups. The array
+represents a potentially multi-dimensional array.  The blob represents a byte
+stream of arbitrary contents.  The fundamental difference between an array and
+a blob is that the array has metadata specifying the dimension(s). At the
+physical layer within the IO nodes, all of these objects may be striped across
+multiple IO nodes.  Given this context, the transactions come in two forms.
 
-First is a single leader transaction where the IOD manages based on calls from
-a single client. The underlying assumption is that the client side will manage
+First is a single leader transaction managed by IOD based on calls from a
+single client. The underlying assumption is that the client side will manage
 the transactional operations itself and the single client is capable of
-reporting to the IOD how to evolve the transaction state.
+reporting to the IOD how to evolve the transaction state. Passive failures are
+detected based on the single client only. Any actively detected failure will
+have to be managed by this single client.
 
 The second form is called multi-leader and has the IOD layer manage the
-transactions. In this case, when the transaction is created, a count of clients
-is provided to the IOD layer. As clients commit their changes to the container,
-the reference count is reduced. Once the count reaches 0, the transaction is
-automatically committed.
+transactions from a collection of clients. In this case, when the transaction
+is created, a count of clients is provided to the IOD layer. As clients commit
+their changes to the container, the reference count is reduced. Once the count
+reaches 0, the transaction is automatically committed. Failure detection is
+fully passive based on all participating clients.
 
@@
 
+\subsection{Comparison to Other Protocols}
+Alternatives, such as Paxos~\cite{Lamport:1998:paxos} algorithms like
+ZooKeeper~\cite{Hunt:2010:zookeeper}, suffer from two limitations making them
+unsuitable for this environment. First, the distributed servers in Paxos
+systems are all distributed copies of each other that eventually become
+consistent. Given the scale we wish to address, a single node's memory is
+unlikely to be able to hold all of the data necessary for many operations at
+scale. They also do not have a guarantee for when consensus will be achieved
+without using slower synchronous calls. For the tight timing we wish to
+support, we need guarantees of when a consistent state has been achieved.
+Second, these systems also all assume that updates are initiated from a single
+client process rather than a parallel set of processes as is the standard in
+HPC environments.
+
+Another effort to offer consistency and data integrity for the ZFS file
+system~\cite{zhang:2010:zfs} covers some of the same territory. Instead of a
+focus on the processes all having a notion of completion as a transaction, this
+work focuses on the integrity of the data movement operations. We view this
+work as something that should be considered hand-in-hand with a transaction
+approach to ensure the integrity of the movement of the data in addition to the
+agreement of processes about the successful completion of a parallel operation.
+
+The transactions and epochs approach offered by FFSIO is most similar to
+\DDT, but is a specialized implementation of a similar protocol. In this case,
+\DDT is general and must use more general techniques. The FFSIO transactions
+and epochs take advantage of simplifications based on the architecture and the
+very specific use.
+
+\DDT uses a second layer of coordination on the client side that greatly
+increases the scalability by consolidating messages from clients into unique
+sets prior to sending to the overall coordinator. A gossip
+protocol~\cite{ganesh:2003:gossip-protocols} may appear sufficient for this
+purpose, but the delay of eventual consistency is strictly avoided with this
+protocol to ensure guarantees at particular states in the code. For example, if
+a failure occurs, the global understanding of the role of all processes is
+required in order for effective communication to occur for operations like
+creating sub-transactions or voting. In this case, the protocol can offer
+stronger statements about consistency than these protocols offer.  These
+features offer a way to easily scale the transaction protocol given the
+guarantees we wish to offer. The IOD approach does nothing to address these
+concerns.
+
+The details of \DDT is discussed next to offer a performance evaluation that
+may be slower than FFSIO transactions and epochs for synchronous due to the
+explicit messaging rather than relying on the hardware mechanisms. At scale,
+the FFSIO performance should be similar to the \DDT approach.
+
 \subsection{Comparison to \DDTns}
@@
 by IOD offers different functionality than \DDTns's example services. In the
-case of \DDTns, the functionality is more minimal, but also avoids some of the
-concerns outlined below.
+case of \DDTns, the functionality is less, but also avoids some of the concerns
+outlined below.
 
@@
 explored in detail in a previous paper~\cite{lofstead:2013:pdsw-txn}.
-%The
-%breakdown of the number of participants in each role is shown in
-%Table~\ref{tab:scaling}. For comparison, consider the Number of
-%Sub-Coordinators equivalent to IOD processes. The Processes Per Sub-Coordinator
-%represents the number of clients that use a particular IOD process. For these
-%tests, we maintained a balanced distribution and always used at least two
-%sub-coordinators to slow down the processing.
+The
+breakdown of the number of participants in each role is shown in
+Table~\ref{tab:scaling}. For comparison, consider the Number of
+Sub-Coordinators equivalent to IOD processes. The Processes Per Sub-Coordinator
+represents the number of clients that use a particular IOD process. For these
+tests, we maintained a balanced distribution and always used at least two
+sub-coordinators to slow down the processing.
 
@@
 \centering
-\includegraphics[keepaspectratio=true, width=0.9\columnwidth]{images/performance}
-\vspace{-0.15in}
+\includegraphics[keepaspectratio=true, width=\columnwidth]{images/performance}
+%\vspace{-0.15in}
 \caption{Total Transaction Overhead}
@@
 
-%\begin{table}[ht]
-%    \vspace{-0.15in}
-%    \centering
-%    \caption[Scaling Configuration]{Performance Tests Scaling Configuration}
-%    \bigskip
-%    \vspace{-0.15in}
-%
-%    \begin{tabular}{|r|r|r|}
-%\hline
-%Processes & \vtop{\hbox{\strut Number of}\hbox{\strut Sub-Coordinators}} & \vtop{\hbox{\strut Processes Per} \hbox{\strut Sub-Coordinator}}\\
-%\hline
-%%8 & 2 & 4 \\
-%%16 & 2 & 8 \\
-%%32 & 2 & 16 \\
-%%64 & 2 & 32 \\
-%%128 & 2 & 64 \\
-%256 & 2 & 128 \\
-%512 & 2 & 256 \\
-%1024 & 4 & 256 \\
-%2048 & 8 & 256 \\
-%4096 & 16 & 256 \\
-%8192 & 32 & 256 \\
-%16384 & 64 & 256 \\
-%32768 & 128 & 256 \\
-%65536 & 256 & 256 \\
-%\hline
-%    \end{tabular}
-%    \label{tab:scaling}
-%\end{table}
+\begin{table}[ht]
+    %\vspace{-0.15in}
+    \centering
+    \caption[Scaling Configuration]{Performance Tests Scaling Configuration}
+    \bigskip
+    %\vspace{-0.15in}
+
+    \begin{tabular}{|r|r|r|}
+\hline
+Processes & \vtop{\hbox{\strut Number of}\hbox{\strut Sub-Coordinators}} & \vtop{\hbox{\strut Processes Per} \hbox{\strut Sub-Coordinator}}\\
+\hline
+%8 & 2 & 4 \\
+%16 & 2 & 8 \\
+%32 & 2 & 16 \\
+%64 & 2 & 32 \\
+%128 & 2 & 64 \\
+256 & 2 & 128 \\
+512 & 2 & 256 \\
+1024 & 4 & 256 \\
+2048 & 8 & 256 \\
+4096 & 16 & 256 \\
+8192 & 32 & 256 \\
+16384 & 64 & 256 \\
+32768 & 128 & 256 \\
+65536 & 256 & 256 \\
+\hline
+    \end{tabular}
+    \label{tab:scaling}
+\end{table}
 
@@
 \centering
-\includegraphics[keepaspectratio=true, width=0.5\columnwidth]{images/optimized-protocol}
-\vspace{-0.15in}
+\includegraphics[keepaspectratio=true, width=0.8\columnwidth]{images/optimized-protocol}
+%\vspace{-0.15in}
 \caption{Optimized Protocol}
@@
 potential issues can be addressed successfully.
-
-%\subsection{Comparison to Other Protocols}
-%Alternatives, such as Paxos~\cite{Lamport:1998:paxos} algorithms like
-%ZooKeeper~\cite{Hunt:2010:zookeeper}, suffer from two limitations making them
-%unsuitable for this environment. First, the distributed servers in Paxos
-%systems are all distributed copies of each other that eventually become
-%consistent. Given the scale we wish to address, a single node's memory is
-%unlikely to be able to hold all of the data necessary for many operations at
-%scale. They also do not have a guarantee for when consensus will be achieved
-%without using slower synchronous calls. For the tight timing we wish to
-%support, we need guarantees of when a consistent state has been achieved.
-%Second, these systems also all assume that updates are initiated from a single
-%client process rather than a parallel set of processes as is the standard in
-%HPC environments.
-%
-%\DDT uses a second layer of coordination on the client side that greatly
-%increases the scalability by consolidating messages from clients into unique
-%sets prior to sending to the overall coordinator. A gossip
-%protocol~\cite{ganesh:2003:gossip-protocols} may appear sufficient for this
-%purpose, but the delay of eventual consistency is strictly avoided with this
-%protocol to ensure guarantees at particular states in the code. For example, if
-%a failure occurs, the global understanding of the role of all processes is
-%required in order for effective communication to occur for operations like
-%creating sub-transactions or voting. In this case, the protocol can offer
-%stronger statements about consistency than these protocols offer.  These
-%features offer a way to easily scale the transaction protocol given the
-%guarantees we wish to offer. The IOD approach does nothing to address these
-%concerns.
-%
-%Another effort to offer consistency and data integrity for the ZFS file
-%system~\cite{zhang:2010:zfs} covers some of the same territory. Instead of a
-%focus on the processes all having a notion of completion as a transaction, this
-%work focuses on the integrity of the data movement operations. We view this
-%work as something that should be considered hand-in-hand with a transaction
-%approach to ensure the integrity of the movement of the data in addition to the
-%agreement of processes about the successful completion of a parallel operation.
 
@@
 
+We wish to thank Eric Barton for the some of the figures used to illustrate
+the system architecture.
+
 \bibliographystyle{abbrv}
